### 2016-08-12 main.tex
First commit

\documentclass{bioinfo}
\copyrightyear{2016} \pubyear{2016}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

\begin{document}
\firstpage{1}

\subtitle{Subject Section}

\title[scater package]{\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R}
\author[McCarthy \textit{et~al}.]{Davis J.~McCarthy\,$^{\text{\sfb 1,2,5}*}$, Kieran R.~Campbell\,$^{\text{\sfb 2,4}}$, Aaron T.~L.~Lun\,$^{\text{\sfb 6}}$ and Quin F.~Wills\,$^{\text{\sfb 2,3}}$}
\address{$^{\text{\sf 1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, Cambridgeshire, UK;\\
$^{\text{\sf 2}}$Wellcome Trust Centre for Human Genetics, University of Oxford,
Oxford, Oxfordshire, UK;\\
$^{\text{\sf 3}}$Weatherall Institute for Molecular Medicine, University of Oxford, Oxford, Oxfordshire, UK;\\
$^{\text{\sf 4}}$Department of Physiology, Anatomy and Genetics, University of Oxford, Oxford, Oxfordshire, UK;\\
$^{\text{\sf 5}}$St Vincent's Institute of Medical Research, 41 Victoria Parade Fitzroy Victoria 3065, Australia; and \\
$^{\text{\sf 6}}$CRUK Cambridge Institute, Robinson Way, Cambridge CB2 0RE,
Cambridgeshire, UK.
% $^{\text{\sf 4}}$Department of Statistics, University of Oxford, Oxford, Oxfordshire, UK;
}

\corresp{$^\ast$To whom correspondence should be addressed.}

\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

\editor{Associate Editor: XXXXXXX}

\abstract{\textbf{Motivation:} Single-cell RNA sequencing (scRNA-seq) is increasingly used to study gene expression at the level of individual cells. However, preparing raw sequence data for further analysis is not a straightforward process. Biases, artifacts, and other sources of unwanted variation are present in the data, requiring substantial time and effort to be spent on pre-processing, quality control (QC) and normalisation.\\
\textbf{Results:} We have developed the R/Bioconductor package \emph{scater} to facilitate rigorous pre-processing, quality control, normalisation and visualisation of scRNA-seq data. The package provides a convenient, flexible workflow to process raw sequencing reads into a high-quality expression dataset ready for downstream analysis. \emph{scater} provides a rich suite of plotting tools for single-cell data and a flexible data structure that is compatible with existing tools and can be used as infrastructure for future software development.\\
\textbf{Availability:} The open-source code, along with installation instructions, vignettes and case studies, is available through Bioconductor at
\href{http://bioconductor.org/packages/scater}{http://bioconductor.org/packages/scater}.\\
\textbf{Contact:} \href{davis@ebi.ac.uk}{davis@ebi.ac.uk}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}

\maketitle

\section{Introduction}\label{introduction}

Single-cell RNA sequencing (scRNA-seq) describes a broad class of techniques which profile the transcriptomes of individual cells. This provides insights into cellular processes at a resolution that cannot be matched by bulk RNA-seq experiments \citep{Hebenstreit2011-ig,Shalek2013-kf}. With scRNA-seq data, the contributions of different cell types to the expression profile of a heterogeneous population can be explicitly determined. Rare cell types can be interrogated and new cell subpopulations can be discovered. Graduated processes such as development and differentiation can also be studied in greater detail. However, this improvement in resolution comes at the cost of increased technical noise and biases. This means that pre-processing, quality control and normalisation are critical to a rigorous analysis of scRNA-seq data. The increased complexity of the data across hundreds or thousands of cells also requires sophisticated visualisation tools to assist interpretation of the results.

Numerous statistical methods and software tools have been published for scRNA-seq data \citep{Guo2015-wx,Kharchenko2014-rx,Finak2015-rd,Delmans2016-qc,Angerer2015-sw,Kiselev2016-fu,Julia2015-jt,Trapnell2014-gj}. However, all of these assume that quality control and normalisation have already been applied. Fewer methods are available in the literature to perform these basic steps in scRNA-seq data processing \citep{Ilicic2016-dm}. This issue is exacerbated by the diversity of scRNA-seq data sets with respect to the experimental protocol and the biological context of the study, meaning that a single processing pipeline with fixed parameters is unlikely to be universally applicable. Rather, software tools are required that support an interactive approach to analysis. This allows parameters to be fine-tuned for the study at hand in response to any issues diagnosed during data exploration. The provided functionality should also process the data in a statistically rigorous manner and encourage reproducible bioinformatics analyses.

One of the most popular frameworks for interactive analysis is the R programming language, extended for biological data analysis through the Bioconductor project \citep{Huber2015-en}. While Bioconductor packages have been widely used for bulk RNA-seq data, the existing data structures (like the ExpressionSet class) and methods are not sufficient for scRNA-seq data. This is because they do not support data types or processing methods that are specific to single-cell studies (e.g., cell-cell distance matrices for clustering, cell-based quality control). Extensions to the current computational infrastructure are required to add these capabilities for scRNA-seq analyses.

Furthermore, large-scale single-cell studies will contain data types beyond the expression profiles, such as intensity values from fluorescence-activated cell sorting, cell imaging data, and information from epigenetic and targeted genotyping assays. Appropriate data structures and methods are required that can accommodate these rich datasets for integrative analyses of expression and other assay data along with the accompanying metadata.

In addition, current visualisation methods designed for exploratory data analysis of bulk transcriptomic experiments are unsuited to scRNA-seq data sets containing hundreds or thousands of cells. The large size of each dataset also favours methods such as \emph{kallisto} \citep{Bray2016-pj} and \emph{Salmon} \citep{Patro2015-jf} for rapidly quantifying gene expression, and there is a need for new computational infrastructure to process raw scRNA-seq sequence data into a high-quality expression dataset ready for downstream analysis.

% \enlargethispage{12pt}

Here we present \emph{scater}, an open-source R/Bioconductor software package that implements a convenient data structure for representing scRNA-seq data and contains functions for pre-processing, quality control, normalisation and visualisation. The package provides wrapper functions for running \emph{kallisto} and \emph{Salmon} on raw read data and converting their output into gene-level expression values, methods for computing and visualising quality-control metrics for cells and genes, and methods for normalisation and correction of uninteresting covariates. This is done in a single software environment which enables seamless integration with a large number of existing tools for scRNA-seq data analysis in R. The \emph{scater} package provides basic infrastructure upon which customized scRNA-seq analyses can be constructed, and we anticipate the package to be useful across the whole spectrum of users, from experimentalists to computational scientists.\vspace*{-12pt}

\begin{methods}
\section{Methods, Data and Implementation}\label{methods-data-and-implementation}

\subsection{Case study with scRNA-seq data}\label{case-study-with-scrna-seq-data}

The results presented in the main paper and supplementary case study use an unpublished single-cell RNA-seq dataset consisting of 73 cells from two lymphoblast cell lines of two unrelated individuals. Cells were captured, lysed, and cDNA generated using the popular C1 platform from Fluidigm, Inc. (\href{https://www.fluidigm.com/products/c1-system}{www.fluidigm.com/products/c1-system}). The processing of the two cell lines was replicated across two machines, with the nuclei of the two cell lines stained with different dyes before mixing on each machine. Cells were imaged before lysis, with an example image provided together with these data (see Case Study in Supplementary Material). Samples were sequenced with paired-end sequencing using the HiSeq 2500 Sequencing system (Illumina). RNA-seq reads were were mapped to a custom genome reference, consisting of Homo sapiens GRCh37 (primary assembly from \href{ftp://ftp.ensembl.org/pub/release-75/fasta/homo_sapiens/dna/}{ftp://ftp.ensembl.org/pub/release-75/fasta/homo_sapiens/dna/}, last accessed 14/08/2015), Epstein-Barr Virus type 1 (B95-8 strain, Accession NC\_007605.1) and ERCC RNA spike-ins (ThermoFisher).  Reads in fastq format were aligned using TopHat2 v2.0.12 \citep{Kim2013-qb} using Bowtie2 v2.2.3.0 \citep{Langmead2012-yc} as the core-mapping engine (\verb|--mate-inner-dist 190 --mate-std-dev 40 --report-secondary-alignments|) and other default parameters. Potential PCR duplicates were marked with Picard MarkDuplicates v1.92(1464). Reads mapping uniquely to annotated exon features were counted using htseq-count implemented in HTSeq, version 0.6.1p1 \citep{Anders2015-wf}.

Further case studies using \emph{scater} on published data, for example from 3000 mouse cortex cells \citep{Zeisel2015-ab} and 1200 cells from early-development mouse embryos \citep{Scialdone2016-oa} are available at \href{http://dx.doi.org/10.5281/zenodo.59897}{dx.doi.org/10.5281/zenodo.59897}.

% We used these two genomes -
% Human genome GrCh37 (ENSEMBL release 75)
% EBV : Accession NC_007605.1 (also known as EBV B95-8)
% The reference genome was a merged file of Human, EBV and ERCC spike-ins and all reads were aligned with Tophatv2.0.12, duplicates with Picard MarkDuplicates VN:1.92(1464), reads counted with HTSeq-counts v0.6.1p1
% The remaining methods are same as what I shared with Davis when he was writing his methods (also quoted below).

%These are the commands:
%@PG ID:TopHat VN:2.0.12 CL:/usr/local/genetics/bin/tophat2 --mate-inner-dist 100 --mate-std-dev 50 --num-threads 4 --library-type fr-unstranded -o REX/WTCHG_146282_709508_tophat --transcriptome-index=/data/Genomes/BOWTIE2/GRCh37.EBVB95-8wt.ERCC-transcriptome /data/Genomes/BOWTIE2/GRCh37.EBVB95-8wt.ERCC FASTQ/WTCHG_146282_709508_1.fastq.gz FASTQ/WTCHG_146282_709508_2.fastq.gz

%@PG ID:TopHat.1 VN:2.0.12 CL:/usr/local/genetics/bin/tophat2 --mate-inner-dist 100 --mate-std-dev 50 --num-threads 4 --library-type fr-unstranded -o REX/WTCHG_146283_709508_tophat --transcriptome-index=/data/Genomes/BOWTIE2/GRCh37.EBVB95-8wt.ERCC-transcriptome /data/Genomes/BOWTIE2/GRCh37.EBVB95-8wt.ERCC FASTQ/WTCHG_146283_709508_1.fastq.gz FASTQ/WTCHG_146283_709508_2.fastq.gz

%@PG ID:MarkDuplicates PN:MarkDuplicates VN:1.92(1464) CL:net.sf.picard.sam.MarkDuplicates INPUT=[/well/bsg/projects/CP140108/merge/merged_BAM/lol_1_43_accepted_hits_merge.bam] OUTPUT=/well/bsg/projects/CP140108/merge/merged_deDup_BAM/lol_1_43_accepted_hits_merge_deDup.bam METRICS_FILE=/well/bsg/projects/CP140108/merge/merged_deDup_BAM/Metrics/lol_1_43_accepted_hits_merge_deDup.bam.metrics.txt REMOVE_DUPLICATES=true TMP_DIR=[/well/htseq/TMP] VERBOSITY=ERROR    PROGRAM_RECORD_ID=MarkDuplicates PROGRAM_GROUP_NAME=MarkDuplicates ASSUME_SORTED=false MAX_SEQUENCES_FOR_DISK_READ_ENDS_MAP=50000 MAX_FILE_HANDLES_FOR_READ_ENDS_MAP=8000 SORTING_COLLECTION_SIZE_RATIO=0.25 READ_NAME_REGEX=[a-zA-Z0-9]+:[0-9]:([0-9]+):([0-9]+):([0-9]+).* OPTICAL_DUPLICATE_PIXEL_DISTANCE=100 QUIET=false VALIDATION_STRINGENCY=STRICT COMPRESSION_LEVEL=5 MAX_RECORDS_IN_RAM=500000 CREATE_INDEX=false CREATE_MD5_FILE=false

\begin{figure}[!tpb]%figure1
\centerline{\includegraphics[width=0.48\textwidth]{figures/scater_qc_workflow.pdf}}
\caption{An overview of the \emph{scater} workflow, from raw sequenced reads to a tidy data set ready for higher-level downstream analysis. For step 5, explanatory variables include experimental covariates like batch, cell source and other recorded information, as well as QC metrics computed from the data. Step 6 describes an optional round of normalisation to remove effects of particular explanatory variables from the data. Automated computation of QC metrics and extensive plotting functionality support the workflow.}\label{fig:01}
\end{figure}

\subsection{Implementation}\label{implementation}

The \emph{scater} package is an open-source R package available through     Bioconductor. Key aspects of the code are written in C++ to minimise computational time and memory use. The package builds on many other R packages, including \emph{Biobase} and \emph{BiocGenerics} for core Bioconductor functionality \citep{Huber2015-en}; \emph{destiny} \citep{Angerer2015-sw} and \emph{Rtsne} \citep{Krijthe2015-is} for dimensionality reduction; and \emph{edgeR} \citep{Robinson2010-ky} and \emph{limma} \citep{Ritchie2015-so} for model fitting and statistical analyses. The plotting functionality in the package uses \emph{ggplot2} \citep{Wickham2016-dc}. A full set of dependencies is provided in the Supplementary Materials.

\end{methods}

\section{Results}\label{results}

\subsection{The \emph{scater} package}\label{the-scater-package}

The \emph{scater} package offers a workflow to convert raw read sequences into a data set ready for higher-level analysis within the R programming environment (Figure~\ref{fig:01}). In addition, \emph{scater} provides basic computational infrastructure to standardise and streamline scRNA-seq data analyses. Key features of \emph{scater} include: (1) the ``single-cell expression set'' (SCESet) class, a data structure specialized for scRNA-seq data; (2) wrapper methods to run \emph{kallisto} and \emph{Salmon} and process their output into gene-level expression values; (3) automated calculation of quality control metrics, with QC visualisation and filtering methods to retain high-quality cells and informative features; (4) extensive visualisation capabilities for inspection of scRNA-seq data; and (5) methods to identify and remove uninteresting covariates affecting expression across cells. The package integrates many commonly used tools for scRNA-seq data analysis and provides a foundation on which future methods can be built. The methods in \emph{scater} are agnostic to the form of the input data and are compatible with counts, transcripts-per-million, counts-per-million, FPKM or any other appropriate transformation of the expression values.

\subsection{SCESet: a data structure for single-cell expression
data}\label{sceset-a-data-structure-for-single-cell-expression-data}

The \emph{scater} package is built around the SCESet class (Supplementary Figure~1) which provides a sophisticated container for scRNA-seq data. This class inherits from the ExpressionSet class in Bioconductor's \emph{Biobase} package \citep{Huber2015-en}, which allows assay data (and multiple transformations thereof), gene or transcript metadata and sample metadata to be combined in a single object to empower robust analyses. While the ExpressionSet class is the basis of many microarray and bulk RNA-seq anaysis methods in Bioconductor, extensions to the class design are necessary to support scRNA-seq data analyses. Specifically, the SCESet class adds slots to store a reduced-dimension representation of the expression profiles, to easily visualize the relationships between cells; cell-cell and gene-gene pairwise distance matrices, for clustering or regulatory network reconstruction; bootstrapped expression results (such as from \emph{kallisto}), to gauge the accuracy of expression quantification; consensus clustering results, where cluster assignments for each cell are combined from different methods to improve reliability; information about feature controls (such as ERCC spike-ins), which is required in downstream steps such as normalization, QC and detection of highly variable genes; and several more (Supplementary Figure~1). With these extra slots, SCESet objects can support analyses of scRNA-seq data that ExpressionSet cannot. In addition, extra data types such as FACS marker expression or epigenetic information can be easily stored in each SCESet object for integration with the single-cell expression profiles.

An SCESet data object can be easily subsetted by row or column to remove unwanted genes or cells, respectively, from all data and metadata fields stored in the object. Furthermore, data and metadata in multiple SCESet objects can be easily combined e.g., to incorporate cells from different experimental batches. SCESet objects can also be converted to other R data structures, or saved to disk in structured, shareable formats. Further details on the class, including its motivation and execution, are available in the Supplementary Case Study and the package documentation. All methods available in \emph{scater} are applicable to instances of the SCESet class and exploit the availability and richness of (meta)data stored in each SCESet object.

\subsection{Data pre-processing}\label{data-pre-processing}

An important initial step in scRNA-seq data processing is to quantify the expression level of genomic features such as transcripts or genes from the raw sequencing data. Approaches to expression quantification from raw reads are, in principle, the same for scRNA-seq as they are for bulk RNA-seq \citep{Kanitz2015-wp,Teng2016-sc}. Read counts obtained from conventional quantification methods such as \emph{HTSeq} \citep{Anders2015-wf} and \emph{featureCounts} \citep{Liao2014-ui} can be readily stored in an SCESet object and used in a \emph{scater} workflow (Figure~\ref{fig:01}). Another option is to use computationally-efficient pseudoalignment methods such as \emph{kallisto} and \emph{Salmon}. This is especially appealing for large scRNA-seq data sets containing hundreds to tens of thousands of cells. To this end, \emph{scater} also provides wrapper functions for \emph{kallisto} and \emph{Salmon} so that fast quantification of transcript-level expression can be managed completely within an R programming environment. A common subsequent step for these methods is to collapse transcript-level expression to gene-level expression. Exploiting the \emph{biomaRt} R/Bioconductor package, \emph{scater} provides a convenient function for using Ensembl annotations to obtain
gene-level expression values and gene or transcript annotations \citep{Yates2016-ia}.

\enlargethispage{6pt}

\begin{figure*}[!htpb]%figure3
\centerline{\includegraphics[width=\textwidth]{figures/figure3.pdf}}
\caption{\textbf{Different types of QC plot that can be generated with \emph{scater}.} (\textbf{a}) Cumulative expression plot showing the proportion of the library accounted for by the top 1--500 most highly-expressed features. (\textbf{b)} PCA plot produced using a subset of the QC metrics computed with \emph{scater's} calculateQCMetrics function. (\textbf{c)} Plot of the 50 most-expressed features (computed according to the highest total read counts) across all cells in the data set. For each feature, the circle represents the percentage of counts that the gene accounts for computed from counts pooled across all cells. The genes are ordered by this value. The bars for each gene show the percentage of counts accounted for by the gene for each individual cell, providing a visualisation of the distribution across cells. (\textbf{d}) Plot of frequency of expression (percentage of cells in which the feature is deemed expressed) against mean expression level across cells. (\textbf{e}) Density plot showing the percentage of variance explained by a set of explanatory variables across all genes. (\textbf{f}) Violin, scatter- and boxplots of principal component values against the C1 machine used for each cell for the six principal components most strongly correlated with C1 machine used. Each individual plot is produced by a single call with either the function plot (\textbf{a}), plotPCA (\textbf{b}) or plotQC (\textbf{c-f}).}\label{fig:03}
\end{figure*}

\subsection{Data quality control}\label{data-quality-control}

The \emph{scater} package provides methods to compute relevant QC metrics for an SCESet object. Given a set of control genes and/or cells, a variety of QC metrics will be computed and returned to the object in a single call to the calculateQCMetrics function (see package documentation). Cell-specific QC metrics include the total count across all genes, the total number of expressed genes, and the percentage of counts allocated to control genes like spike-in transcripts or mitochondrial genes. These metrics are useful for identifying low-quality cells---for example, a high percentage of counts mapping to spike-ins typically indicates a small amount of RNA captured for the cell, suggesting protocol failure or death of the cell in processing and means it is unlikely to be suitable for downstream analyses. For each gene, QC metrics such as the average expression level and the proportion of cells in which the gene is expressed are computed. This can be used to identify low-abundance genes or genes with high dropout rates that should be filtered out prior to downstream analyses. All of these metrics are used by scater to construct QC plots to diagnose potential issues with data quality. This facilitates quality control which---despite attempts at automation \citep{Ilicic2016-dm}---still requires manual intervention to account for aspects of the data specific to each study. The package documentation provides full details of the QC metrics produced.

In \emph{scater}, the default plot method for an SCESet object produces
a cumulative expression plot (Figure~\ref{fig:03}a). This plot describes how reads are distributed across genes, distinguishing between low-complexity libraries (where very few genes contain most of the counts) and their high-complexity counterparts (where counts are distributed more evenly across genes).
For example, there is substantial variability in library complexity among cells in the case study dataset in Figure~\ref{fig:03}a. Some cells have profiles similar to the blank wells,
suggesting that library preparation or sequencing failed for these cells and
that the corresponding libraries should be removed prior to further analysis.
Cell phenotype variables can be incorporated into these plots to highlight
differences in expression distributions for different types of cells. For
example, the curve for each cell is coloured by the type of well that produced
the library (Figure~\ref{fig:03}a), while cells can also be split into separate
facets by library type to show more metadata variables simultaneously (see
Supplementary Case Study). Cumulative expression plots should be favoured over
boxplots as the default method for visualising expression distributions across
cells in a dataset, as the latter performs poorly at handling the long tail of
low- and zero-expression observations in scRNA-seq data.

The plotPCA function implements an approach to automatic outlier
detection using multivariate normal methods applied to the cell-level QC
metrics \citep{Ilicic2016-dm}. Specifically, PCA is applied to the QC metrics
for all cells and a plot is produced to automatically detect outliers in the
higher-dimensional QC metric space (Figure~\ref{fig:03}b). These outliers
correspond to low-quality cells with abnormal library characteristics (e.g.,
low total counts and few expressed genes) that should be removed prior to
downstream analysis. This automated approach is powerful but also difficult to interpret, and so complements simpler filtering approaches that apply thresholds to particular QC metrics.

The plotQC function generates many types of plots useful for quality control, such as a plot of visualising the most
highly-expressed features in the dataset (Figure~\ref{fig:03}c). This
provides a feature-centric overview of the dataset that simultaneously
visualises the features with highest total expression across all cells,
while also displaying the distribution of cell-level expression values
for these features. It is common to see ERCC spike-ins (if used),
mitochondrial and ribosomal genes among the highest expressed genes,
while datasets consisting of healthy cells will also show high levels of
constitutively expressed genes like \emph{GAPDH} and \emph{ACTB}. This
plot allows the analyst to quickly check that the gene- or
transcript-level quantification is behaving as expected, and to flag
datasets where it is not.

A key feature of scRNA-seq is the high frequency of ``dropout'' events,
that is, no observed expression (such as no read counts) in a particular
cell for a gene that is actually expressed in that cell. Typically only
a small set of genes are observed with detectable expression in every
cell. The plotQC function can also be used to visualise the frequency of
expression (inverse of the dropout percentage) of features
against their average expression level (Figure~\ref{fig:03}d). Control features
can be highlighted easily in the plot, and typical scRNA-seq datasets will show
a broadly sigmoidal relationship between average expression level and
frequency of expression across cells. This is consistent with expected
behaviour where genes with greater average expression are more readily
captured during library preparation and are detected at a greater
frequency \citep{Brennecke2013-zv,Kim2015-xd,Vallejos2015-ww}.

Another important step in quality control is to identify variables (experimental factors or computed QC metrics) that drive variation in expression data across cells. The plotQC function
provides a novel approach to identifying variables that
have substantial explanatory power for many genes. For each variable in
the phenoData slot of the SCESet object, we fit a linear model for each
feature with just that variable as the explanatory variable. We then plot the distribution of the marginal $R^2$ values across all features for the variables with the most
explanatory power for the dataset (Figure~\ref{fig:03}e). The variables are ranked by median $R^2$ across features in the plot, allowing users to identify variables that may need to be considered
during normalisation or statistical modelling. The plotQC function can
also assess the influence of variables of interest by plotting principal
components of the expression matrix most strongly correlated with a
variable of interest against that variable. For example, in the Case
Study data, the first principal component is correlated with the
C1 machine used to process the cell (Figure~\ref{fig:03}f).

We also introduce the plotPhenoData function for convenient plotting of
cell phenotype information (including QC metrics), and the
plotFeatureData function for plotting feature information (see examples in the
Supplementary Case Study). These methods will work not only on the
SCESet class defined in \emph{scater}, but also on any ExpressionSet object,
providing sophisticated plotting functionality for many other Bioconductor
packages and contexts.

The \emph{scater} graphical user interface (GUI) provides convenient
access to \emph{scater}'s QC and visualisation methods (Supplementary
Figures~3--5). This opens an interactive interface in a web browser that facilitates exploration of the data through QC plots and other intuitive visualisations. The GUI allows users of any background to easily examine the effects of changing multiple parameters, which can be helpful for quickly conducting exploratory data analysis. Useful settings can then be stored in R scripts to ensure that data analyses are reproducible.

In summary, \emph{scater} provides a variety of novel and convenient
methods to visualise an scRNA-seq dataset for QC. Low-quality cells and
uninteresting genes can then be easily removed by filtering and
subsetting the SCESet data structure prior to further analysis.

\subsection{Data visualisation}\label{data-visualisation}

Dimensionality reduction techniques are necessary to convert
high-dimensional expression data into low-dimensional representations
for intuitive visualisation of the relationships, similarities and
differences between cells. To this end, \emph{scater} provides
convenient functions to apply a variety of dimensionality reduction
procedures to the cells in an SCESet object. Functions include plotPCA,
to perform a principal components analysis; plotTSNE, to perform
t-distributed stochastic neighbour embedding \citep{Van_der_Maaten2008-oe}, which has
been widely used for scRNA-seq data \citep{Amir2013-nf,Bendall2014-gf,Macosko2015-vt};
plotDiffusionMap, to generate a diffusion map \citep{Haghverdi2015-fm} for
visualising differentiation processes; and plotMDS, to generate
multi-dimensional scaling plots (Figure~\ref{fig:04}a--c). The plotReducedDim
function can also be used to plot any reduced-dimension representation of cells
(e.g., an independent component analysis produced by \emph{monocle}
\citep{Trapnell2013-fv} or similar) that is stored in an SCESet object.

By default, the PCA and t-SNE plots are produced using the features with
the most variable expression across all cells. We focus on the most variable genes to highlight any
heterogeneity in the data that might be driving interesting differences between
cells. Alternatively, we can apply \emph{a priori} knowledge to define a set of
genes that are associated with a biological process of interest, and construct
plots using only these features. For example, \citealp{Scialdone2015-gj}
found that using prior knowledge to define feature sets is vital for
exploring processes like the cell cycle, which can have substantial effects on
single-cell expression measurements \citep{Buettner2015-jg}. The subsetting
and filtering methods for SCESet objects make it easy to construct
reduced-dimension plots for particular gene sets, in order to
investigate certain effects in the data such as those due to the cell
cycle (Figure~\ref{fig:04}d--f).

The various types of reduced-dimension plots can be used to examine the structure of the cell population, including the formation of distinct subpopulations or the presence of continuous trajectories. Cell-level variables stored in the
SCESet object can be used to define the shape, colour and size of points
plotted, allowing more information to be conveniently incorporated into
each plot (e.g., cells are coloured by \emph{CCND2} expression in
Figure~\ref{fig:04}d--f). The plotExpression function is also provided for
plotting expression levels of a particular gene against any of the cell
phenotype variables or the expression level of
another feature (Figure~\ref{fig:04}g). This allows the user to inspect the expression levels
of a feature or set of features in full detail, rather than relying only
on summary information and reduced-dimension plots where information is
necessarily lost.

\begin{figure}[!tpb]%figure4
\centerline{\includegraphics[width=0.48\textwidth]{figures/figure4.pdf}}
\caption{\textbf{Reduced dimension representations of cells and gene expression plots with \emph{scater.}} Plots are shown using all genes (\textbf{a-c}) and cell cycle genes only (\textbf{d-f}) using PCA (\textbf{a,d}), t-SNE (\textbf{b,e}) and diffusion maps (\textbf{c,f}). In the top row (\textbf{a-c}), points are coloured by patient of origin, sized by total features (number of genes with detectable expression) and the shape indicates the C1 machine used to process the cells. In the second row (\textbf{d-f}), points are coloured by the expression of \emph{CCND2} (ENSG00000118971), a gene associated with the G1/S phase transition of the cell cycle. With the plotExpression function, expression for sets of genes can be plotted against any cell metadata variables or the expression of a given gene (\textbf{g}). The function automatically detects whether the x-axis variable is categorical or continuous and plots the data accordingly, with x-axis values ``jittered'' to avoid excessive overplotting of points with the same x coordinate.}\label{fig:04}
\end{figure}

\subsection{Data normalisation and batch correction} \label{data-normalisation-and-batch-correction}

Scaling normalisation is typically required in RNA-seq data analysis to remove biases caused by differences in sequencing depth, capture efficiency or composition effects between samples. Frequently used methods for scaling normalization include the trimmed mean of M-values \citep{Robinson2010-dy}, relative log-expression \citep{Anders2010-kv} and upper-quartile methods \citep{Bullard2010-ui}, all of which are available for use in \emph{scater}. In addition, \emph{scater} is tightly integrated with the \emph{scran} package that implements a method utilising cell pooling and deconvolution to compute size factors better suited to scRNA-seq data \citep{Lun2016-sk}.

After scaling normalisation, further correction is typically required to ameliorate or remove batch effects. For example, in the case study dataset, cells from two patients were each processed on two C1 machines. Although C1 machine is not one of the most important explanatory variables on a per-gene level (Figure~\ref{fig:03}e), this factor is correlated with the first principal component of the log-expression data (Figure~\ref{fig:03}f). This effect cannot be removed by scaling normalisation methods, which target cell-specific biases and are not sufficient for removing large-scale batch effects that vary on a gene-by-gene basis (Figure~\ref{fig:05}a). Here we present two possibilities, all easily implemented in a scater workflow.

The C1 machine effect is known from the design of the experiment, so we
can easily regress out this effect in \emph{scater}. With the
normaliseExprs function the user can supply a design matrix of variables
to regress out of the expression values, and residuals from the linear
model fit can be used as expression values for downstream analyses. For
the dataset here, we fit a linear model to the \emph{scran} normalised
log-expression values with the C1 machine as an explanatory factor. (We
also use the log-total counts from endogenous genes, percentage of
counts from the top 100 most highly-expressed genes and percentage of
counts from control genes as additional covariates.) We then use the
residuals from the fitted model for further analyses (see Case Study in
Supplementary Material). This approach successfully removes the C1
machine effect as a major source of variation between cells; the first
principal component now separates the cells from the two patients, as
expected (Figure~\ref{fig:05}b). This approach needs to be used carefully
as single-cell data often deviate from normal distributions, but in many
cases, as here, it can successfully ameliorate large-scale known batch
effects.

In addition to removing known batch effects, it can be important for
large data sets to identify (potentially unknown) sources of unwanted variation
\citep{Leek2010-nq,Hicks2015-qy,Tung2016-jy,Bacher2016-ay,Grun2015-xi}.
\emph{scater} is compatible with existing methods such as \emph{svaseq}
\citep{Leek2007-rg,Leek2014-nu} and \emph{RUVSeq} \citep{Risso2014-np} to
identify and remove these unwanted sources of variation, and the
removeBatchEffect method in the \emph{limma} package \citep{Ritchie2015-so}
to account for known batch effects. Here, just removing the first latent
variable identified by the RUVs method from \emph{RUVSeq} is sufficient to
remove the machine effect, as the PCA plot now separates cells by patient
rather than C1 machine (Figure~\ref{fig:05}c).

% Applying removeBatchEffect from
% \emph{limma} yields normalised data for which PCs are no longer correlated with
% C1 machine (see Case Study in Supplementary Material), but the patient effect is
% not the primary driver of variation amongst the PCs (Figure~\ref{fig:05}c).

We emphasise that it is generally preferable to incorporate batch effects or latent variables into statistical models used for inference. Where this is not possible (e.g., for visualisation), directly regressing out these uninteresting factors is required to obtain "corrected" expression values for further analysis.

\begin{figure}[!tpb]%figure5
\centerline{\includegraphics[width=0.48\textwidth]{figures/figure5.pdf}}
\caption{\textbf{Figure 5: Normalisation approaches made easy with
\emph{scater}.} Principal component analysis plots showing cell
structure in the first two PCA dimensions using various normalisation
methods methods that can be easily applied in \emph{scater}, including
endogenous size-factor normalisation using methods from the \emph{scran}
package (\textbf{a}); expression residuals after applying size-factor
normalisation and regressing out known, unwanted sources of variation
(\textbf{b}); and removal of one hidden factor identified using
the RUVs method from the \emph{RUV} package (\textbf{c}). In all plots,
the colour of points is determined by the patient from which cells were
obtained, shape is determined by the C1 machine used to process the
cells and size reflects the total number of genes with detectable
expression in the cell.}\label{fig:05}
\end{figure}

\subsection{Software and data integration}\label{software-and-data-integration}

As part of the R/Bioconductor ecosystem, \emph{scater} can be easily
integrated with other software for scRNA-seq data analysis (Supplementary Figure~2). Because the
SCESet class builds on existing Bioconductor data structures, most
Bioconductor packages for expression analyses are able to operate
seamlessly with SCESet objects. Tools that can integrate easily with
\emph{scater} include many options for data normalisation
\citep{Lun2016-sk,Vallejos2015-ww,Ding2015-jv}, differential expression analysis
\citep{Vallejos2016-qy,Trapnell2014-gj,Finak2015-rd,Vu2016-sk,Kharchenko2014-rx,Korthauer2015-wg,Andrews2016-at}, heterogeneous gene expression analyses
\citep{Lun2016-sk,Vallejos2015-ww}, clustering
\citep{Kiselev2016-fu,Guo2015-wx,Fan2016-wh,Grun2015-fy}, latent or hidden
variable analysis
\citep{Leek2014-nu,Risso2014-np,Stegle2012-is,Chikina2015-lq},
cell cycle phase identification \citep{Scialdone2015-gj} and pseudotime
computation \citep{Trapnell2014-gj,Angerer2015-sw,Julia2015-jt,Campbell2016-es,Campbell2015-nj,Haghverdi2016-is}. The \emph{scater} package bridges
the gap between raw reads and these downstream analysis tools by providing the
pre-processing, QC, visualisation and normalisation methods and a data
structure combining multiple data modalities and metadata necessary for
convenient, robust and reproducible analyses of scRNA-seq data.

\section{Discussion}\label{discussion}

Single-cell RNA sequencing is widely used for high-resolution gene
expression studies investigating the behaviour of individual cells.
While scRNA-seq data can provide substantial biological insights, the
complexity and noise of the data is also much greater than that of
conventional bulk RNA-seq. Thus, rigorous analysis of scRNA-seq data
requires careful quality control to remove low-quality cells and genes,
as well as normalisation to adjust for biases and batch effects in the
expression data. Failure to carry out these procedures correctly is
likely to compromise the validity of all downstream analyses
\citep{Leek2010-nq,Hicks2015-qy,Tung2016-jy,Bacher2016-ay,Grun2015-xi}.

Here, we present an R/Bioconductor package, \emph{scater}, that provides
crucial infrastructure and methods for low-level scRNA-seq data
analysis. The package introduces a data structure tailored to scRNA-seq
data that is compatible with a vast number of existing tools in the
Bioconductor project. The \emph{scater} data structure combines multiple
transformations of the expression data with cell and feature (gene or
transcript) metadata and allows data sets to be easily standardised and
shared. Wrapper functions for the popular RNA-seq quantification methods
\emph{kallisto} and \emph{Salmon} facilitate the processing of raw read
sequences to a SCESet object in R with expression data and accompanying
metadata.

Quality control is a vital preliminary step for scRNA-seq and can be a
time-consuming manual task. We present a tool for automated computation
of QC metrics, novel plotting methods for QC and convenient subsetting
and filtering methods to substantially simplify the process of filtering
out unwanted or problematic cells and genes. The package provides a
large array of sophisticated plotting functions so that cells can be
visualised with a variety of popular dimensionality-reduction techniques
in plots that incorporate cell metadata and expression values as
plotting variables.

Normalisation is a critical aspect of scRNA-seq data processing that is supported by \emph{scater}. Scaling normalisation methods, including the single-cell specific methods in the \emph{scran} package, are seamlessly integrated into a \emph{scater} workflow. Methods for identifying and removing batch effects and other types of unwanted variation are supported both with internal methods and through integration with a multitude of tools available in the R/Bioconductor environment. Once identified, important covariates and latent variables can be flagged for inclusion in downstream statistical models or their effects regressed out of normalised expression values.

Future development will include further extensions to data structures that will enable tight integration of single-cell transcriptomic, genetic and epigenetic data, as well as further refinement of the methods available as the single-cell field matures. Although \emph{scater} has been produced for
scRNA-seq data, its capabilities are well suited for single-cell qPCR data
and bulk RNA-seq data, and will prove useful for supporting analyses of these data types too.

\section{Conclusion}\label{conclusion}

The \emph{scater} package eases the burden for a user tasked with
producing a high-quality single-cell expression dataset for downstream analysis. The intuitive GUI implemented in \emph{scater} provides an easier entry point into rigorous analysis of scRNA-seq data for users without a
computational background, enabling them to process raw reads into high-quality expression data within a single computing environment. Experienced users can take advantage of \emph{scater}'s data structures, wide array of tools, suitability for scripted analyses and seamless integration with many other R/Bioconductor analysis tools. The data structures and methods in
\emph{scater} provide basic infrastructure upon which new scRNA-seq
analysis tools can be developed. We anticipate that \emph{scater} will be a
useful resource for both analysts and software developers in the
single-cell RNA sequencing field.\vspace*{-10pt}

\section*{Acknowledgements}\label{acknowledgements}

We thank Marco Salvetti for provision of the two samples and Zam Cader for processing the samples. We would also like to acknowledge Peter Donnelly and Oliver Stegle for support and helpful discussions.\vspace*{-12pt}

\section*{Funding}

DJM was supported by funding through an Early Career Fellowship from the National Health and Medical Research Council of Australia, and by core funding from EMBL. ATLL was supported by core funding from Cancer Research UK (award no. A17197). KRC is supported by a UK Medical Research Council studentship.\vspace*{-12pt}

\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{plain}
%\bibliographystyle{bioinformatics}
\bibliography{main}

\end{document}

### 2016-08-12 scater_case_study/case_study.Rmd
First commit

---
title: "SCATER CASE STUDY"
author: Davis McCarthy & Quin Wills
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        theme: united
        highlight: tango
        number_sections: false
        code_folding: hide
---

```{r, echo=F, eval=T, message=F, warning=F}

# setwd( "~/Dropbox/scater_case_study/" )
# setwd( "~/Google\ Drive/Manuscripts/scater_paper_nov-2015/scater_case_study/" )
library( "scater" )
library( "EBImage" )
library( "scran" )
library( "RUVSeq" )
```

1. **OVERVIEW:** Brief introduction to scater and the case study data.
2. **GETTING STARTED:** Import data/meta-data,  and run automated QC.
3. **QC THE CELLS:** Exploratory plots and quality control of the cells.  
4. **QC THE GENES:** Exploratory plots and quality control of the genes.  
5. **NORMALISE THE DATA**: Unwanted technical effects in the data, and corrective normalisation.  
6. **DOWNSTREAM ANALYSIS:** An example of further analysis.
7. **TECHNICAL STUFF:** Information for bioinformaticians, such as the R session used to run this case study.

---

### 1. OVERVIEW

This case study is designed to walk you through key scater features from importing and easily QC'ing your single-cell data,  through to seamless integration with other data types and [Bioconductor]( https://www.bioconductor.org/ ) packages. If you're new to single cell RNA-seq,  this case study should help you get started with your own data,  irrespective of the platforms used to process the cells for sequencing. As scater is an [R]( https://www.r-project.org/ ) package,  some of the basic work flow is also implemented as a [shiny]( http://shiny.rstudio.com/ ) GUI for those less familiar with R coding. The aim is to get you from sequencer to higher level analysis and modelling with minimal fuss.

The RNA-seq data provided here are 73 cells from two lymphoblast cell lines of two unrelated individuals. Cells were captured,  lysed,  and cDNA generated using a popular [microfluidics platform]( https://www.fluidigm.com/products/c1-system?gclid=CIO41cywp8kCFdgaGwodwYUCHA ). The processing of the two cell lines was replicated across two machines,  with the nuclei of the two cell lines stained with different dyes before mixing on each machine. Cells were imaged before lysis,  with an example image provided together with this data.

To run this case study you will require the following packages in addition to scater: [EBImage]( http://bioconductor.org/packages/release/bioc/html/EBImage.html ),  [qvalue]( https://bioconductor.org/packages/release/bioc/html/qvalue.html ),  and [mclust]( https://cran.r-project.org/web/packages/mclust/index.html ).  
\

---

### 2. GETTING STARTED

Scater processing can begin with sequencer FASTQ files,  using the inbuilt [Kallisto]( http://nextgenseek.com/2015/05/kallisto-a-new-ultra-fast-rna-seq-quantitation-method/ ) pseudo-aligner. This is dicussed [here](  ) and in the supporting help files. Here we assume that read counts have already been provided from a more traditional RNA-seq pipeline,  and read them in as follows:  
\

```{r, echo=T, eval=T, message=F, warning=F}

reads <- as.matrix( read.csv( "reads.csv", row.names = 1 ) )

```
\

In keeping with best practice in object-oriented bioinformatics, `scater` not only allows you to store your data in a standardised format, but also to associate it with experimental and bioinformatic meta-data. Experimental meta-data are any information associated with individual cells (also called 'samples'), from sequencing metrics to experimental batching and other experimental cell phenotypes. The bioinformatic meta-data are any information related to the genes (also called 'features'). These can be read in,  combined with the sequencing data,  and extracted as shown below. For this example we'll use the automated log-scaled counts per million generated by `scater`,  but see ```calculateTPM()``` as an example function for generating alternative RNA-seq read count formats.  
\

View the first few entris for the cell metadata:

```{r, echo=T, eval=T, message=F, warning=F}

cells <- AnnotatedDataFrame( read.csv( "cells.csv", row.names = 1 ) ) # row names match column names in reads
genes <- AnnotatedDataFrame( read.csv( "genes.csv", row.names = 1 ) ) # row names match row names in reads
scData <- newSCESet( countData = reads, phenoData = cells, featureData = genes, lowerDetectionLimit = 3 )

pData( scData )[1:3, 1:2] # view some sample meta-data
```

View the first few entries for the feature metadata:

```{r, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
fData( scData )[1:3, 1:2] # view some feature meta-data

```
\

Scater comes with an inbuilt QC function,  which is worth running once all data has been imported as an SCESet object. To do this,  first define any feature (gene) and sample (cell) controls. Here we define ERCC synthetic RNA spike-ins as our feature controls,  with bulk sequencing samples and empty capture sites (i.e. no cells) as blank controls.   
\

```{r, echo=T, eval=T, message=F, warning=F}

spikes <- grepl( "ERCC-", row.names( scData ) ) # ERCC spike-ins
blanks <- scData$sample_type == "blank (empty site) control" # negative/background controls
bulks <- grepl( "bulk control", scData$sample_type ) # samples of several hundred cells
scData <- calculateQCMetrics(
    scData, feature_controls = list( spikes = spikes ), 
    cell_controls = list( bulks = bulks, blanks = blanks ) )

```
\

---

### 3. QC THE CELLS

A useful first step is flagging/failing poorly performing cells. This can be done from the sample meta-data using the automated QC metrics generated above,  any additional sequencing metrics from sequencing aligner/mapping software,  and additional cell phenotypes such as from imaging. For the sake of demonstration, here we focus on four metrics. Others you may want to consider are % reads mapped to mitochondrial genes,  library PCR duplication rate,  and mean sequencing bias per cell.  
\

```{r, echo=T, eval=T, message=F, warning=F}

scData$use <- (scData$cdna_recovered_in_ng_per_ul > 0.5 & #sufficient cDNA per cell 
                   scData$pct_counts_feature_controls < 1 & # sufficient endogenous RNA
                   !scData$total_counts < 1e5 & # sufficient reads mapped to features
                   !scData$filter_on_total_features & # remove cells with unusual numbers of genes
                   !scData$is_cell_control # controls shouldn't be used in downstream analysis
)

```
\

Box plots aren't particularly useful for visualising sparse data, so ```plot()``` applied to an SCESet object helps visualise all cells as a cumulative proportion of reads per cell. You can see from the plot below that the two failed cells have curves that look more like the blank controls.  
\

```{r, echo=T, eval=T, message=F, warning=F, fig.width=10, fig.height=6, fig.align="center"}

plot( scData, block1 = "sample_type", block2 = "c1_machine", colour_by = "use",
      exprs_values = "counts")

```
\

Scater allows users total flexibility to run their favourite dimension reduction methods,  as decribed [here]() and in the supporting help files. Here we use ```plotPCA()``` to further explore the cells. You can see that the two cell lines cluster separately,  with the respective bulk control samples occuring within the correct clusters. With all single-cell work there is background RNA (secreted or from lysed cells),  with the blanks here clustering between the two samples. One feature to look out for is if individual blanks cluster strongly with one or other cell type,  suggesting that the background admixture isn't homogeneous. As there are more samples here from patient B,  blanks  demonstrate a stronger patient B admixture.  
\

```{r, echo=T, eval=T, message=F, warning=F, fig.width=7, fig.height=6, fig.align="center"}

plotPCA( scData, colour_by = "use", shape_by = "sample_type" )

plotPCA( scData[, grep("cell", scData$sample_type)], colour_by = "use", 
         shape_by = "sample_type")

```
\  
Another option available in `scater` is to conduct PCA on a set of QC metrics.
The advantage of doing this is that the QC metrics focus on technical aspects of
the libraries that are likely to distinguish problematics cells. Automatic 
outlier detection on PCA plots using QC metrics is available to help identify 
potentially problematic cells.

By default, the following metrics are used for PCA-based outlier detection:

* `pct_counts_top_100_features`
* `total_features`
* `pct_counts_feature_controls`
* `n_detected_feature_controls`
* `log10_counts_endogenous_features`
* `log10_counts_feature_controls`

A particular set of variables to be used can be specified with the 
`selected_variables` argument as shown in the example below.

```{r pca-qc-metrics, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=7, fig.height=5, fig.align="center"}

## PCA on the phenoData cannot handle missing values
## for the exercise here we thus set cdna_recovered_in_ng_per_ul to 100 where 
## there are NA values actually (creating a new dummy variable)
cdna_temp <- scData$cdna_recovered_in_ng_per_ul
cdna_temp[is.na(cdna_temp)] <- 100
scData$cdna_recovered_in_ng_per_ul_no_miss <- cdna_temp
scData <- plotPCA( scData, size_by = "use", shape_by = "sample_type", 
                   pca_data_input = "pdata", detect_outliers = TRUE,
                   return_SCESet = TRUE, 
                   selected_variables = c("cdna_recovered_in_ng_per_ul_no_miss",
                                          "pct_counts_top_100_features",
                                          "total_features", 
                                          "pct_counts_feature_controls",
                                          "n_detected_feature_controls", 
                                          "log10_counts_endogenous_features",
                                          "log10_counts_feature_controls"))

plotReducedDim(scData, colour_by = "outlier", shape_by = "use", ncomponents = 3)

plotReducedDim(scData[, grep("cell", scData$sample_type)], 
               colour_by = "outlier", shape_by = "use", ncomponents = 3)

plotPCA( scData[, grep("cell", scData$sample_type)], colour_by = "outlier", 
         shape_by = "use", pca_data_input = "pdata")
plotDiffusionMap( scData[, grep("cell", scData$sample_type)], 
                  colour_by = "outlier", shape_by = "use", 
                  pca_data_input = "pdata")
```

Having detected potential outliers on the phenotype data (i.e. cell metrics) we
can see where these cells appear on PCA and diffusion map plots based on the 
expression data.

```{r pca-difmap-outliers, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5, fig.align="center"}
plotPCA( scData[, grep("cell", scData$sample_type)], ncomponents = 3,
         shape_by = "use", pca_data_input = "pdata", colour_by = "outlier")

p1 <- plotPCA( scData, colour_by = "use", shape_by = "sample_type", 
         size_by = "outlier" ) + ggtitle("PCA")
p2 <- plotDiffusionMap( scData, colour_by = "use", shape_by = "sample_type", 
         size_by = "outlier" ) + ggtitle("Diffusion map")
multiplot(p1, p2, cols = 2)
```

With `scater`, any specific set of features based on prior knowledge can be used for PCA, t-SNE or diffusion maps. A feature set to use can be defined by supplying the `feature_set` argument to `plotPCA`, `plotTSNE` or `plotDiffusionMap`. This allows, for example, using only housekeeping features or control features or cell cycle genes to produce reduced-dimension plots. The plots below use only the spike-in genes defined as such earlier.

```{r pca-difmap-ercc-genes, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5, fig.align="center"}
p1 <- plotPCA( scData, feature_set = fData(scData)$is_feature_control, 
               colour_by = "use", shape_by = "sample_type", 
               size_by = "outlier" ) + ggtitle("PCA")
p2 <- plotDiffusionMap( scData, feature_set = fData(scData)$is_feature_control, 
                        colour_by = "use", shape_by = "sample_type", 
                        size_by = "outlier" ) + ggtitle("Diffusion map")
multiplot(p1, p2, cols = 2)
```

The following plot uses only the genes annotated as cell cycle genes (this annotation was included in the genes information loaded into the SCESet object at the start). Points are coloured by expression of RAN (ENSG00000132341), a gene associated with the G2/M phase of the cell cycle.

```{r pca-diffmap-ccn-genes, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=10, fig.height=5, fig.align="center"}
cell_cycle_genes <- !is.na(fData(scData)$cell_cycle_phase)
sum(cell_cycle_genes)
most_exprs_ccycle <- which(cell_cycle_genes & 
                               fData(scData)$total_feature_counts > 100000)
knitr::kable(fData(scData)[most_exprs_ccycle,] )
p1 <- plotPCA(scData, feature_set = cell_cycle_genes, 
               colour_by = featureNames(scData)[most_exprs_ccycle[8]], 
               shape_by = "sample_type", 
               size_by = "use" ) + ggtitle("PCA")
p2 <- plotDiffusionMap( scData, feature_set = cell_cycle_genes, 
                        colour_by = featureNames(scData)[most_exprs_ccycle[8]], 
                        shape_by = "sample_type", 
                        size_by = "use" ) + ggtitle("Diffusion map")
multiplot(p1, p2, cols = 2)

```

```plotPhenoData()``` can be used to explore specific sample meta-data values. For example,  in the plot below we can see strong association between cDNA recovery and number of genes detected in a cell. The two failed cells demonstrate low gene coverage (total features) and cDNA recovery.

\

```{r, echo=T, eval=T, message=F, warning=F, fig.width=6, fig.height=4.5, fig.align="center"}

plotPhenoData( scData, aesth = aes_string( x = "cdna_recovered_in_ng_per_ul", 
                                         y = "total_features", colour = "use" ) )

```
\

One element of QC that can prove challenging to judge from the above criteria is if more than one cell ( or cell debris from other cells ) are being sequenced. With most cell processing platforms a small percentage of cells won't be truly single. One solution is imaging individual cells,  which also allows for the incorporation of fluorescence phenotypes such as viability and nuclear content ( for cell cycle analysis ). Using the [EBImage]( http://bioconductor.org/packages/release/bioc/html/EBImage.html ) package,  an example image of a stained cell is processed and displayed below,  with results incorporated into the sample meta-data.  
\

```{r, echo=T, eval=T, fig.width=4, fig.height=4, fig.align="center"}

imageQC <- function(cell) { 
    require( "EBImage" )
    pass <- TRUE # if the cell should be passed or failed

    im <- readImage( paste( cell, ".tif", sep = "" ) ) # read in image TIFF for cell of interest
    im <- thresh( im, w = 20, h = 20, offset = 0.1 ) # convert to binary fluorescence matrix 
    display( im, method = "raster" ) # view fluorescent spot(s)
    im <- bwlabel(  opening(  im,  makeBrush( 3, "gaussian" ) ) ) # label clusters of glowing spots

    cl <- sort( table( im )[-1], decreasing = TRUE ) # pixels per fluorescent spot
    if ( sum( cl > 1e2 ) != 1 ) pass <- FALSE # pass if a single spot greater than 100 pixels

    sh <- prcomp( which( im == names( cl )[1],  arr.ind = TRUE ) )  # a non-circular shape may be a doublet/dividing cell
    if ( sh$sdev[1] ^ 2/sum( sh$sdev ^ 2 ) > 0.75 ) pass <- FALSE

    pass
}

pData( scData )["cellB_1_04", "use"] <- 
    (pData( scData )["cellB_1_04", "use"] & imageQC( "cellB_1_04" ) )

```
\

---

### 4. QC THE GENES

Scater allows you to set minimum QC thresholds for a gene to be considered sufficiently expressed in your downstream analysis. Here,  using the inbuilt  ```is_exprs()``` function,  we enforce that a gene must have least three reads in at least three cells in both replicates.  
\

```{r, echo=TRUE, eval=TRUE, }
reps <- scData$c1_machine # the machine replicates
is_exprs(scData) <- counts(scData) >= 3
exprsThresh <- sapply( 1:2,  
               function(r) apply( is_exprs( scData )[, scData$use & reps == r], 1, 
                                    function(x) sum( x ) >= 3 ) )
fData( scData )$use <- (apply( exprsThresh, 1, all ) & # endogenous genes above the threshold are used downstream
                            !fData( scData )$is_feature_control_spikes)

```
\

It can be useful to plot gene expression frequency versus mean expression level
to assess the effects of technical dropout in the dataset. We fit a non-linear
least squares curve for the relationship between expression frequency and mean
expression and use this to define the number of genes above high technical 
dropout and the numbers of genes that are expressed (here defined as at least 4
counts) in at least 50% and at least 25% of cells. A subset of genes to be 
treated as feature controls can be specified, otherwise any feature controls 
previously defined are used.
\

```{r, echo=TRUE, eval=TRUE, fig.width=7, fig.height=5, fig.align="center"}
plotQC(scData, type = "exprs") 
```
\

It can also be useful to look at total expression levels acrsoo gene biotypes. 
The plot below shows the distributions of log-10 scale total gene counts for 
genes of different annotated biotypes. The biotypes are ordered by median total
counts, and the jittered points beneath the violin plots indiciate the number of 
genes in the biotype. We want to see many protein-coding genes with reasonably 
high expression. This is exactly what we see here, with protein-coding genes 
appearing as the biotype with second-highest median expression behind the very 
small (but highly expressed) category of Mt_rRNA genes.
\

```{r, echo=T, eval=T, message=F, warning=F, fig.width=8, fig.height=6, fig.align="center"}
oo <- order(unlist(dplyr::summarise(dplyr::group_by(fData(scData), gene_biotype),
                 median(log10_total_feature_counts))[,2]), decreasing = TRUE)
fData(scData)$biotype <- factor(fData(scData)$gene_biotype, 
                                levels = levels(fData(scData)$gene_biotype)[oo])

plotFeatureData(scData, aes(x = biotype, y = log10_total_feature_counts),
                theme_size = 6) +
    theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5))

```
\

The ```plotQC()``` function provides several useful  QC plots,  such as the example below that considers the the number of reads consumed by the top 50 expressed genes. Aside from spike-ins, these are typically mitochondrial and housekeeping genes. Here,  as with most single-cell experiments,  a large proportion of reads are being are taken up by uninteresting biology.

To make gene labels more interpretable we append the gene symbol to the Ensembl
gene IDs.

```{r append-gene-symbols}
featureNames(scData) <- paste(fData(scData)$hgnc_symbol, featureNames(scData),
                              sep = "_")
```

```{r, echo=T, eval=T, message=F, warning=F, fig.width=6, fig.height=6, fig.align="center"}
plotQC( scData[fData( scData )$use, pData( scData )$use], 
        type = "highest-expression", 
        col_by_variable = "sample_type" )
```

---

### 5. NORMALISE THE DATA

Once you have filtered cells and genes,  a next step is to explore technical drivers of variability in the data to inform data normalisation before downstream analysis.

Experimental design is a critical, but neglected, aspect of \gls{scrnaseq} studies. To the best of my knowledge, methods like those described in this section for exploring experimental and QC variables and the experimental design, do not feature in any scRNA-seq software packages apart from `scater`. There are a very large number of potential confounders, artifacts and biases in \gls{scrnaseq} studies. Exploring the effects of such explanatory variables (both those recorded during the experiment and computed QC metrics) is crucial for appropriate modeling of the data. The \scater\ package provides a set of methods specifically for quality control of experimental and explanatory variables, which will be demonstrated briefly here.

Using the ```plotPCA()``` function we can see that principal component one is driven by differences between the two machine replicates,  which in turn is due to differences in gene coverage. Differences in number of detected genes is a common driver of cell clustering  and can be result of biology (e.g. different cell types, cell cycle). However, it often has a strong technical component due to variably recovered RNA,  reverse transcription,  or library amplification. Its effect can also be notably non-linear,  affecting low expressed and high expressed genes differently. The ```plotQC()``` function can be used to explore the the marginal % variance explained (per gene) of the various technical factors. In the second plot we can see that it's not unusual for gene coverage to explain more than 10% of the expression variance of a gene.  
\

```{r pca-diffmap-normalisation, echo=T, eval=T, message=F, warning=F, fig.width=7, fig.height=5, fig.align="center"}
plotPCA( scData[fData( scData )$use, scData$use],  # only plot genes and cells of interest
         colour_by = "sample_type", 
         shape_by = "c1_machine", 
         size_by = "total_features" )

plotDiffusionMap( scData[fData( scData )$use, scData$use], ntop = 5000,
                  colour_by = "sample_type", 
                  shape_by = "c1_machine", 
                  size_by = "total_features" )
```

The relative importance of different explanatory variables can be explored with some of the `plotQC` function options. Supplying the `type = "expl"` argument to `plotQC` computes the marginal $R^2$ for each variable in the \textsf{SCESet} when fitting a linear model regressing expression values for each gene against just that variable, and displays a density plot of the gene-wise marginal $R^2$ values for the variables. The default approach looks at all variables in the \textsf{phenoData} slot of the object and plots the top `nvars_to_plot` variables (default is 10).

Alternatively, one can choose a subset of variables to plot in this manner, which we do here. The density curves for marginal $R^2$ show the relative importance of different variables for explaining variance in expression between cells.

```{r plotqc-expl-vars, echo=T, eval=T, message=F, warning=F, fig.width=8, fig.height=4.5, fig.align="center"}
plotQC( scData[fData( scData )$use, scData$use], 
        type = "explanatory-variables", 
        variables = c("pct_counts_top_100_features", "total_features", 
                      "pct_counts_feature_controls", "c1_machine",
                      "n_detected_feature_controls", 
                      "log10_counts_endogenous_features",
                      "log10_counts_feature_controls", "sample_type") )
```

This analysis indicates that total number of features detected and the sequencing depth (number of counts) for endogenous genes, in particular, have substantial
explanatory power for many genes, so these variables are good candidates for
conditioning out in a normalisation step, or including in downstream statistical
models. The number of detected feature controls (spike-in genes) does not appear to be an important explanatory variable.

One can also easily produce plots to identify principal components that correlate with experimental and QC variables of interest. The function `plotQC` with the option `type = "find-pcs"` ranks the principal components in decreasing order of $R^2$ from a linear model regressing PC value against the variable of interest. The default behaviour is to show the relationships between the variable of interest and the six principal components with the strongest relationship to the variable (as measured by $R^2$). This works both for continous and categorical variables. This type of plot  can indicate which explanatory variables may be driving differences between cells as detected by PCA and highlight which PCs are associated with the variable. The "total features" variable shows very strong correlation with both principal components 1 and 2.

```{r, echo=T, eval=T, message=F, warning=F, fig.width=10, fig.height=6, fig.align="center", message=FALSE, results='hide'}
p1 <- plotQC( scData[fData( scData )$use, scData$use], 
        type = "find-pcs", variable = "total_features" )
p2 <- plotQC( scData[fData( scData )$use, scData$use], 
        type = "find-pcs", variable = "sample_type" )
multiplot(p1, p2, cols = 2)
```

The cells in this dataset were processed on two C1 machines, which gives rise to a substantial batch effect. The QC plots below show that the C1 machine factor is correlated with the first principal component. This effect is likely caused by the difference in amount of cDNA recovered per cell, which is even more strongly correlated with the first principal component. Thus, differences in batches are a major source of variation between the cells.

```{r plot-qc-pcs-c1-machine}
scData$c1_machine <- as.factor(scData$c1_machine)
p1 <- plotQC( scData[fData( scData )$use, scData$use], 
        type = "find-pcs", variable = "c1_machine" )
p2 <- plotQC( scData[fData( scData )$use, scData$use], 
        type = "find-pcs", variable = "cdna_recovered_in_ng_per_ul")
multiplot(p1, p2, cols = 2)
```

The `plotQC()` function can also be used to produce a pairs plot of explanatory variables (with the same calls as above, but with `method = "pairs"`). The plot below shows this use case for looking at the % counts from the top 100 most-expressed genes, the total number of expressed genes, the % of counts from feature controls, the number of detected feature controls, the  number of counts (on the log-10 scale) from endogenous features, the number of counts (log-10 scale) from feature controls and sample type. The explanatory variables are ordered by the median $R^2$ of the variable across all genes, and this value is reported on the plot. This type of plot is useful for finding correlations between experimental and QC variables with substantial explanatory power.

```{r, echo=T, eval=T, message=F, warning=F, fig.width=8, fig.height=8, fig.align="center"}
plotQC( scData[fData( scData )$use, scData$use], 
        type = "expl", method = "pairs", 
        variables = c("pct_counts_top_100_features", "total_features", 
                      "pct_counts_feature_controls", "c1_machine",
                      "n_detected_feature_controls", 
                      "log10_counts_endogenous_features",
                      "log10_counts_feature_controls", "sample_type"),
        theme_size = 6)

```
\

After important explanatory variables have been identified with the tools shown above, their effects can be accounted for in subsequent statistical models, or they can be conditioned out using `normaliseExprs()`, if so desired. If a design matrix incorporating a selection of explanatory variables is supplied as an argument to `normaliseExprs`, then normalised expression values returned for each feature will be the residuals from a linear model fitted with the design matrix, after any size-factor normalisation has been applied to the expression data.

Normalising single-cell RNA-seq data is a topic in its infancy, but many of the basic principles still apply. How much you choose to initially correct for technical factors depends on your question of interest and the ease with which they can be accounted for in downstream models.
\

In `scater` it is easy to perform size-factor normalisation using the "TMM" approach (the default in the [edgeR]() package), the "RLE" approach (default in the [DESeq]() package), an "upper-quartile approach" (as proposed by Bullard et al (2010)) or simple scaling by total counts. Size-factor normalisation can be carried out on a subsetted SCESet object, normalising either to ERCC spike-in genes or to endogenous genes if desired. Under certain circumstances either may be appropriate. Careful thought should be given to applying scale-factor normalisation as underlying assumptions from methods developed for bulk RNA-seq may not be appropriate for all single-cell datasets.

However, [Lun et al (2016)](http://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-0947-7) recently published a size-factor normalisation method specifically designed
for scRNA-seq data. This method performs much better on single-cell data and its
implementation in the Bioconductor package `scran` allows seamless integration 
into the `scater` workflow. (The `scran` package itself depends on `scater`).

We recommend using the `scran` size-factor normalisation approach and 
demonstrate its use here.

In the code below we use the `filter` function to select a subset of cells from
an `SCESet` object (this method works just as the `filter` function in the 
`dplyr` package on which it is based).

```{r pca-norm-plots, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align="center", fig.width=10, fig.height=5.5}
## subset to form a QC's version of the data
scDataQC <- filter(scData, use)
scDataQC <- scDataQC[fData(scData)$use | fData(scData)$is_feature_control,]
ercc_genes <- fData(scDataQC)$is_feature_control
endog_genes <- !ercc_genes

## size factor normalisation with scran 
qclust <- quickCluster(scDataQC, min.size = 20)
scDataQC <- computeSumFactors(scDataQC, sizes = 10, clusters = qclust)
summary(scDataQC$size_factor)
scDataQC <- normalise(scDataQC)

scDataQC_spikenorm <- computeSumFactors(scDataQC, sizes = 10, 
                              get.spikes = TRUE)
summary(scDataQC_spikenorm$size_factor)
scDataQC_spikenorm <- normalise(scDataQC_spikenorm)

set_exprs(scDataQC, "ercc_norm_exprs") <- exprs(scDataQC_spikenorm)
set_exprs(scDataQC, "endog_norm_exprs") <- exprs(scDataQC)
## subset again so that only endogenous genes are used
scDataQC <- scDataQC[fData(scDataQC)$use,]
plt_pca_ercc_norm <- plotPCA(
    scDataQC, exprs_values = "ercc_norm_exprs", size_by = "total_features",  
    colour_by = "sample_type", shape_by = "c1_machine") + 
    ggtitle("PCA - ERCC size-factor normalisation") + 
    theme(legend.position = "bottom")
plt_pca_endog_norm <- plotPCA(
    scDataQC, exprs_values = "endog_norm_exprs", size_by = "total_features",  
    colour_by = "sample_type", shape_by = "c1_machine") + 
    ggtitle("PCA - endogenous size-factor normalisation") + 
    theme(legend.position = "bottom")
multiplot(plt_pca_ercc_norm, plt_pca_endog_norm, cols = 2)
```

In the PCA plots above we see that the first principal component seems to separate cells based on the C1 machine used to process them, even after appropriate size-factor normalisation. The `plotQC` function confirms this in the plot below: principal component 1 is the PC most highly correlated with C1 machine.

```{r  pca-norm-plot-find-pcs, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.align="center", fig.width=6, fig.height=6}
scDataQC$c1_machine <- as.factor(scDataQC$c1_machine)
plotQC( scDataQC, type = "find-pcs", variable = "c1_machine")
```

We would prefer such technical effects to be removed by normalisation, so below we demonstrate a further alternative approach. We see how "customised" normalisation approaches can be easily incorporated into the `scater` workflow.

In this example we normalise and standardise counts conditioned on expression level, showing the PCA for the corrected data. One advantage of this approach is that a biologically 'noisy' gene is naturally defined as one with greater dispersion than other genes at a similar expression level. In the normalised data these are genes with a mean absolute deviation greater than 1.

```{r custom-norm-exprs, echo=T, eval=T, message=F, warning=F, fig.width=8, fig.height=6, fig.align="center"}

rp <- scData$c1_machine[scData$use] # replicates
mn <- sapply( 1:2,  # mean log expression per replicate
              function(r) rowMeans(exprs( scData )[fData( scData )$use, 
                                                   scData$use][, rp == r]))

exprsNorm <- counts( scData )[fData( scData )$use, pData( scData )$use] # read counts to normalise
wn <- floor( 0.05*nrow( scData ) ) # consider a sliding window of 10% of genes
for (r in 1:2) { # normalise per replicate
  exprsNorm[, rp == r] <- t( sapply( row.names( exprsNorm ), function(g){
    nh <- head( sort( abs( mn[, r] - mn[g, r] ) ), wn ) # neighbourhood of similarly expressing genes
    nh <- exprsNorm[names( nh ), rp == r]
    nh <- log2( t( t( nh ) / colSums( nh ) ) * 1e5 + 1 ) # locally normalise
    nh <- nh - mean( nh ) # locally standardise 
    nh <- nh / mean( abs( nh ) )
    nh[1, ]
  } ) )  
}

### add normalised expression values to SCESet object and make PCA plot
norm_exprs(scDataQC) <- exprsNorm
plt_cond_norm <- plotPCA(
    scDataQC, exprs_values = "norm_exprs", shape_by = "c1_machine", 
        colour_by = "sample_type", size_by = "total_features", 
        scale_features = FALSE) + 
    ggtitle("PCA - expression conditional normalisation") + 
    theme(legend.position = "bottom")
plt_cond_norm
```
\

After this normalisation procedure we see that the first principal component now separates cells from the two patients and neither the first nor second principal component appears influenced by the C1 machine used to process the cells.

Furthermore, C1 machine shows no correlation with the top principal components, 
as shown in the plot below.

```{r  pca-norm-plot-find-pcs-2, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.align="center", fig.width=5, fig.height=4}
plotQC( scDataQC, type = "find-pcs", variable = "c1_machine", 
        exprs_values = "norm_exprs")
```

Another option is to regress out effects of technical factors in the 
normalisation step.

```{r regress-factors-norm,  echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align="center", fig.width=6, fig.height=9}
design <- model.matrix(~scDataQC$log10_counts_endogenous_features +
                           scDataQC$pct_counts_top_100_features +
                           scDataQC$pct_counts_feature_controls +
                           scDataQC$c1_machine)
set_exprs(scDataQC, "norm_exprs_resid") <- norm_exprs(
    normaliseExprs(scDataQC, design = design,
                   method = "none", exprs_values = "exprs",
                   return_norm_as_exprs = FALSE) )

plt_sf_norm_resid <- plotPCA(
    scDataQC, exprs_values = "norm_exprs_resid", shape_by = "c1_machine", 
        colour_by = "sample_type", size_by = "total_features", 
        scale_features = TRUE) + 
    ggtitle("PCA - size-factor normalisation residuals") + 
    theme(legend.position = "bottom")
plt_sf_norm_resid
```

As we see here, this simple regression approach is here also able to 
substantially remove the C1 machine effect from that the data so that the first
principal component now separates cells by patient.

We can also check the results that would be obtained when regressing out 
technical effects computed from the expression values but *not* C1 machine. This
can give us an idea of what effects could be removed if we did not know that 
cells were process on different C1 machines and only had access to the QC metrics
we could compute from the expression data, or other sample metadata information
provided.

The PCA plots below show  the results after regressing out log10(counts) 
from endogenous features, % counts from the top 100 most-expressed features, 
the percentage counts from control features and the log10(counts) from feature
controls (top) and after regressing out all of those effects plus cDNA recovered
from each cell in ng per microlitre.

```{r  regress-factors-norm-2,  echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align="center", fig.width=6, fig.height=9}
design2 <- model.matrix(~scDataQC$log10_counts_endogenous_features +
                            scDataQC$pct_counts_top_100_features +
                            scDataQC$pct_counts_feature_controls +
                            scDataQC$log10_counts_feature_controls)
set_exprs(scDataQC, "norm_exprs_resid2") <- norm_exprs(
    normaliseExprs(scDataQC, design = design2,
                   method = "none", exprs_values = "exprs",
                   return_norm_as_exprs = FALSE) )

plt_cond_norm_resid2 <- plotPCA(
    scDataQC, exprs_values = "norm_exprs_resid2", shape_by = "c1_machine", 
        colour_by = "sample_type", size_by = "total_features", 
        scale_features = FALSE) + 
    ggtitle("PCA - size-factor normalisation residuals technical factors") + 
    theme(legend.position = "bottom")

design3 <- model.matrix(~scDataQC$log10_counts_endogenous_features +
                            scDataQC$pct_counts_top_100_features +
                            scDataQC$pct_counts_feature_controls +
                            scDataQC$log10_counts_feature_controls +
                            scDataQC$cdna_recovered_in_ng_per_ul_no_miss)
set_exprs(scDataQC, "norm_exprs_resid3") <- norm_exprs(
    normaliseExprs(scDataQC, design = design3,
                   method = "none", exprs_values = "exprs",
                   return_norm_as_exprs = FALSE) )

plt_cond_norm_resid3 <- plotPCA(
    scDataQC, exprs_values = "norm_exprs_resid3", shape_by = "c1_machine", 
        colour_by = "sample_type", size_by = "total_features", 
        scale_features = FALSE) + 
    ggtitle("PCA - size-factor normalisation residuals incl cDNA") + 
    theme(legend.position = "bottom")

multiplot(plt_cond_norm_resid2, plt_cond_norm_resid3)
```

Here, regressing out only the technical QC metrics does not remove the major 
effect of C1 machine, but regressing out cDNA concentration as well does (so 
the first principal component now separates the two patients). Thus, for 
removing known effects, the regression approach implemented in `scater` is 
effective, but more needs to be done to find _hidden_ effects such as unknown
batch effects, or similar.

Look at the performance of `removeBatch` from the `limma` package:

```{r limma-remove-batch, echo=TRUE, eval=TRUE}
design_rbe <- model.matrix(~as.character(scDataQC$sample_type))
rbe_norm_exprs <- removeBatchEffect(exprs(scDataQC), batch = scDataQC$c1_machine,
                                    design = design_rbe)
rbe_norm_exprs <- removeBatchEffect(get_exprs(scDataQC, "endog_norm_exprs"), 
                                    batch = scDataQC$c1_machine,
                                    design = design_rbe)
set_exprs(scDataQC, "norm_exprs_rbe") <- rbe_norm_exprs
plt_rbe_norm <- plotPCA(
    scDataQC, exprs_values = "norm_exprs_rbe", shape_by = "c1_machine", 
        colour_by = "sample_type", size_by = "total_features") + 
    ggtitle("PCA - removeBatchEffect normalisation") + 
    theme(legend.position = "bottom")
plt_rbe_norm
```

After normalisation with `removeBatchEffect`, the patient effect is still not the strongest one picked out by PCA. Nevertheless, there is now no correlation between PCs and C1 machine, as the plot below shows.

```{r limma-remove-batch-pcs-corr, fig.height=7, fig.width=7}
plotQC(scDataQC, type = "find-pcs", variable = "c1_machine", 
       exprs_values = "norm_exprs_rbe")
```

Look at the performance of the RUVs method:

```{r ruv-seq-s, echo=TRUE, eval=TRUE, fig.height=14, fig.width=9}
# One approach to finding and removing hidden, unwanted sources of variability is
# the `RUVSeq` package.

scDataRUV <- filter(scData, use)
scDataRUV <- scDataRUV[fData(scData)$use | fData(scData)$is_feature_control,]
scIdx <- matrix(-1, ncol = max(table(scDataRUV$sample_type)), nrow = 2)
scIdx[1,] <- which(scDataRUV$sample_type == "cell from patient B")
tmp <- which(scDataRUV$sample_type == "cell from patient A")
scIdx[2, 1:length(tmp)] <- tmp
# cIdx <- (fData(scDataRUV)$is_feature_control & fData(scDataRUV)$n_cells_exprs >= 10) 
## Running RUVs
cIdx <- rownames(scDataRUV)
ruv_list <- list()
for (k in 1:9) {
    ruvs <- RUVs(counts(scDataRUV), cIdx, k = k,
                 scIdx = scIdx, isLog = FALSE)
    norm_counts(scDataRUV) <- ruvs$normalizedCounts
    exprs(scDataRUV) <- edgeR::cpm.default(norm_counts(scDataRUV), log = TRUE, 
                                           prior.count = scDataRUV@logExprsOffset)
    plt_norm_ruv <- plotPCA(
        scDataRUV, exprs_values = "exprs", shape_by = "c1_machine", 
        colour_by = "sample_type", size_by = "total_features", 
        scale_features = TRUE) + 
        ggtitle(paste("PCA - RUVs normalisation: k =", k)) + 
        theme_bw(8)
    ruv_list[[k]] <- plt_norm_ruv
}

figplot_ruv <- cowplot::plot_grid(plotlist = ruv_list[1:8], labels = letters[1:8], ncol = 2)
figplot_ruv
figplot_ruv <- cowplot::plot_grid(plotlist = ruv_list, labels = letters[1:9])
cowplot::save_plot("../figures/supplement/ruvs_figure.png", figplot_ruv, 
                   ncol = 3, nrow = 3, base_height = 5, base_aspect_ratio = 0.9)
```

```{r ruv-seq-s-qc-plots, echo=TRUE, eval=TRUE, message=FALSE, results='hide'}
plotQC(scDataRUV, type = "explanatory-variables")
plotQC(scDataRUV, type = "find-pcs", variable = "sample_type")
```

```{r ruvseq-red-dim-k6}
ruvs <- RUVs(counts(scDataRUV), cIdx, k = 6, scIdx = scIdx, isLog = FALSE)
reducedDimension(scDataRUV) <- ruvs$W
plotReducedDim(scDataRUV, shape_by = "c1_machine", 
               colour_by = "sample_type", size_by = "total_features") + theme_bw()

```

Look at the performance of the RUVg method:

```{r ruv-seq-g, echo=TRUE, eval=TRUE, fig.height=14, fig.width=9}
# One approach to finding and removing hidden, unwanted sources of variability is
# the `RUVSeq` package.
## Running RUVs
cIdx <- fData(scDataRUV)$is_feature_control
ruvg_list <- list()
for (k in 1:9) {
    ruvs <- RUVg(counts(scDataRUV), cIdx, k = k, isLog = FALSE)
    norm_counts(scDataRUV) <- ruvs$normalizedCounts
    exprs(scDataRUV) <- edgeR::cpm.default(norm_counts(scDataRUV), log = TRUE, 
                                           prior.count = scDataRUV@logExprsOffset)
    plt_norm_ruv <- plotPCA(
        scDataRUV, exprs_values = "exprs", shape_by = "c1_machine", 
        colour_by = "sample_type", size_by = "total_features", 
        scale_features = TRUE) + 
        ggtitle(paste("PCA - RUVg normalisation: k =", k)) + 
        theme_bw(8)
    ruvg_list[[k]] <- plt_norm_ruv
}

figplot_ruvg <- cowplot::plot_grid(plotlist = ruvg_list[1:8], labels = letters[1:8],
                                   ncol = 2)
figplot_ruvg
figplot_ruvg <- cowplot::plot_grid(plotlist = ruvg_list, labels = letters[1:9])
cowplot::save_plot("../figures/supplement/ruvg_figure.png", figplot_ruv, 
                   ncol = 3, nrow = 3, base_height = 5, base_aspect_ratio = 0.9)

```

```{r ruv-g-qc-plots, message=FALSE, results='hide'}
plotQC(scDataRUV, type = "explanatory-variables")
plotQC(scDataRUV, type = "find-pcs", variable = "sample_type")
scDataRUV$c1_machine <- as.factor(scDataRUV$c1_machine)
plotQC(scDataRUV, type = "find-pcs", variable = "c1_machine")

```

Unfortunately, the RUVg method does not do a good job of removing the C1 machine effect. The machine used is still correlated with the first PC even after removing 9 hidden factors of unwanted variation.

Thus, after convenient pre-processing, QC and normalisation with \scater, the data
are well organised (with feature and cell metadata and many data transformations),
clean and tidy, and are ready for further statistical modeling and analysis.

---

### 7. DOWNSTREAM ANALYSIS

Expression data at a single-cell resolution not only allows testing for differential expression,  but exploring how this is dependent on sub-types of cells and/or how genes coexpress with each other across cells. As with normalisation,  single-cell analysis methodology is an area in its infancy and deserving of discussion that is beyond this case study. Here we simply demonstrate how QC'ed and normalised data contained within an SCESet object allows for easy downstream interrogation. Let's assume we are interested in defining differential expression as change in expression frequency. This can be done with a standard generalised linear model and the ```qvalue``` package to control false discovery rates. 
\

```{r de-testing, echo=T, eval=T, message=F, warning=F, fig.align="center"}
library( "qvalue" )

sm <- scDataQC$sample_type # the two samples
de <- data.frame( t( apply( norm_exprs(scDataQC) > 0, 1, # test change in expression frequency 
                            function( y ) coef(summary(glm(y ~ sm, family = "binomial" ) ) )[2, c(1,4)])),  check.names = FALSE)
de$qvalue <- qvalue( de[, "Pr(>|z|)"], fdr.level = 0.05 )$qvalues # control for false disovery rate
de <- de[order( de$qvalue, decreasing = FALSE ), ] # order by global statistical evidence
knitr::kable(head( de ))
```

In `scater` the `plotExpression` function enables the convenient visualisation of expression values for a set of features. Here, the expression values for the six most DE genes for expression frequency between patients are shown. The units for expression in the plot can be defined with the `exprs_values` argument (the expression values must exist with the provided name in the `assayData` slot of the SCESet object; if not the default `exprs` values will be used, with a warning). As with other plots in `scater` we can use phenotype data variables to define the colour and shapr of the points.

```{r plot-exprs-de, echo=T, eval=T, message=F, warning=F, fig.align="center", fig.width=9, fig.height=5}
plotExpression(scDataQC, rownames(de)[1:6], x = "sample_type", ncol = 3)
plotExpression(scDataQC, rownames(de)[1:6], ncol = 3, exprs_values = "counts",
               x = "sample_type", colour_by = "log10_counts_endogenous_features")
plotExpression(scDataQC, rownames(de)[1:6], exprs_values = "norm_exprs", 
               x = "sample_type", colour_by = "total_features", 
               shape_by = "c1_machine", ncol = 3)
```
\

One natural question is if differentially expressed genes are co-expressed. Clustering of the genes,  using Gaussian mixture models,  suggests that the immunoglobulin genes with  increased expression in patient B form a tight coexpression cluster (shown in red) that is distinct from the other differentially expressed genes.  
\

```{r, echo=T, eval=T, message=F, warning=F, fig.width=6, fig.height=12, fig.align="center", cache=TRUE}

library( "mclust" )

coexprsA <- cor( t( norm_exprs(scDataQC)[row.names(de)[de$qvalue < 0.05], 
                              sm == "cell from patient A"] ),  # rank coexpression
                 method = "spearman" )
coexprsA <- data.frame( cmdscale( (1 - coexprsA) / 2 ) ) # multidimensional scaling of the coexpression
clustA <- Mclust( coexprsA, modelNames = "VVV" ) # model based clustering of the differentially expressed genes

coexprsB <- cor( t( norm_exprs(scDataQC)[row.names(de)[de$qvalue < 0.05], 
                                         sm == "cell from patient B"] ),  # repeat
                method = "spearman" )
coexprsB <- data.frame( cmdscale( (1 - coexprsB) / 2 ) ) 
clustB <- Mclust( coexprsB, modelNames = "VVV" )

gns <- fData( scDataQC )[row.names( de )[de$qvalue < 0.05], "hgnc_symbol"] # differentially expressed genes' symbols
cls <- c( "firebrick", "grey30" )[clustB$classification] # colour by clustering
eff <- rank( de$Estimate[de$qvalue < 0.05] ) / 60 # size by relative expression in patient B

par( mfrow = c( 2, 1 ) )
plot( clustA, what = "density", bty = "n", xlab = "patient A dimension one",  ylab = "patient A dimension two" )
text( coexprsA[, 1], coexprsA[, 2], gns, cex = eff, col = cls )
plot( clustB, what = "density", bty = "n", xlab = "patient B dimension one",  ylab = "patient B dimension two" )
text( coexprsB[, 1], coexprsB[, 2], gns, cex = eff, col = cls )

```
\

---

### 8.  TECHNICAL STUFF

Scater has been tested on Mac OS X and Linux environments and requires the R packages:

* Biobase
* BiocGenerics
* ggplot2
* methods

and imports the packages:

* biomaRt
* data.table
* dplyr
* edgeR
* grid
* limma
* matrixStats
* plyr
* reshape2
* rhdf5
* rjson
* viridis

This case study was run using the following platform and R package versions:  
\

```{r, echo=FALSE, eval=TRUE}

sessionInfo()

```

```{r figures-for-paper, echo=TRUE, results='hide'}
### Figures for the paper
if ( requireNamespace("cowplot") ) {
    ## QC figure
    scData$well_type <- "single cell"
    scData$well_type[scData$sample_type == "blank (empty site) control"] <-
        "blank control"
    scData$well_type[scData$sample_type == "bulk control from patient A"] <-
        "bulk control"
    scData$well_type[scData$sample_type == "bulk control from patient B"] <-
        "bulk control"
    p1 <- plot( scData, colour_by = "well_type", exprs_values = "counts")
    p2 <- plotPCA( scData, shape_by = "sample_type", 
                   pca_data_input = "pdata", detect_outliers = TRUE,
                   selected_variables = c("cdna_recovered_in_ng_per_ul_no_miss",
                                          "pct_counts_top_100_features",
                                          "total_features", 
                                          "pct_counts_feature_controls",
                                          "n_detected_feature_controls", 
                                          "log10_counts_endogenous_features",
                                          "log10_counts_feature_controls"))
    p3 <- plotQC( scData[fData( scData )$use, pData( scData )$use], 
        type = "highest-expression", 
        col_by_variable = "sample_type" ) + coord_cartesian(xlim = c(0, 6))
    p4 <- plotQC(scData, type = "exprs") 
    p5 <- plotQC( scData[fData( scData )$use, scData$use], 
        type = "explanatory-variables", 
        variables = c("pct_counts_top_100_features", "total_features", 
                      "c1_machine",
                      "n_detected_feature_controls", 
                      "log10_counts_endogenous_features",
                      "sample_type") ) + 
        xlab(expression(paste(R ^ 2, " (% variance explained; log10-scale)"))) +
    theme(legend.position = "bottom", legend.text = element_text(size = 7))
    p6 <- plotQC( scData[fData( scData )$use, scData$use], 
        type = "find-pcs", variable = "c1_machine")
    figplot1 <- cowplot::plot_grid(p1, p2, p3, p4, p5, p6, 
                                  labels = letters, ncol = 3,
                                  label_size = 20)
    cowplot::save_plot("../figures/figure3.pdf", figplot1, ncol = 3, nrow = 2,
                       base_height = 3.5, base_aspect_ratio = 1.4)
    cowplot::save_plot("../figures/figure3.png", figplot1, ncol = 3, nrow = 2,
                       base_height = 3.5, base_aspect_ratio = 1.4)
    cowplot::save_plot("../figures/figure3.jpg", figplot1, ncol = 3, nrow = 2,
                       base_height = 3.5, base_aspect_ratio = 1.4, dpi = 350)

    ## dimension reduction figure
    ww <- which(featureNames(scDataQC) %in% rownames(de)[c(1:2, 5)])
    # ww <- which(featureNames(scDataQC) %in% c("IGHM_ENSG00000211899", 
    #                                           "IGHV4-34_ENSG00000211956",
    #                                           "CD86_ENSG00000114013"))
    # ww <- which(featureNames(scDataQC) %in% c("IGHM", 
    #                                           "IGHV4-34",
    #                                           "CD86"))
    featureNames(scDataQC)[ww] <- as.character(fData(scDataQC)$hgnc_symbol[ww])
    feats_to_plot <- as.character(fData(scDataQC)$hgnc_symbol[ww])

    p1 <- plotPCA(scData[, scData$use], size_by = "total_features", 
                  colour_by = "sample_type", theme_size = 14,
                  shape_by = "c1_machine") + ggtitle("PCA - all genes") + 
        ylab("Component 2") +
        theme(legend.position = "bottom", legend.text = element_text(size = 10), 
              legend.title = element_text(size = 11))
    p2 <- plotTSNE(scData[, scData$use], size_by = "total_features", 
                   colour_by = "sample_type", theme_size = 14,
                   shape_by = "c1_machine", rand_seed = 20151225) + 
        ggtitle("t-SNE - all genes") + 
        theme(legend.position = "bottom", legend.text = element_text(size = 10),
              legend.title = element_text(size = 11))
    p3 <- plotDiffusionMap( scData[, scData$use], size_by = "total_features", 
                            colour_by = "sample_type", 
                            shape_by = "c1_machine", theme_size = 14) + 
        ggtitle("Diffusion map - all genes") +
        theme(legend.position = "bottom", legend.text = element_text(size = 10),
              legend.title = element_text(size = 11))
    p4 <- plotPCA(scData[, scData$use], feature_set = cell_cycle_genes, 
                  colour_by = featureNames(scData)[most_exprs_ccycle[7]], 
                  shape_by = "sample_type", theme_size = 14) + 
        ggtitle("PCA - cell cycle genes") + ylab("Component 2") +
        geom_point(aes(colour = colour_by, shape = shape_by), alpha = 0.6, size = 4) +
        theme(legend.position = "bottom", legend.text = element_text(size = 10),
              legend.title = element_text(size = 11))
    p5 <- plotTSNE(scData[, scData$use], feature_set = cell_cycle_genes, 
                   colour_by = featureNames(scData)[most_exprs_ccycle[7]], 
                   shape_by = "sample_type", rand_seed = 20151225, theme_size = 14) + 
        geom_point(aes(colour = colour_by, shape = shape_by), alpha = 0.6, size = 4) +
        ggtitle("t-SNE - cell cycle genes") + 
        theme(legend.position = "bottom", legend.text = element_text(size = 10),
              legend.title = element_text(size = 11))
    p6 <- plotDiffusionMap( scData[, scData$use], feature_set = cell_cycle_genes, 
                            colour_by = featureNames(scData)[most_exprs_ccycle[7]], 
                            shape_by = "sample_type", theme_size = 14) + 
        geom_point(aes(colour = colour_by, shape = shape_by), alpha = 0.6, size = 4) +
        ggtitle("Diffusion map - cell cycle genes") +
        theme(legend.position = "bottom", legend.text = element_text(size = 10),
              legend.title = element_text(size = 11))

    plt_exprs <- plotExpression(
        scDataQC, feats_to_plot, exprs_values = "norm_exprs", 
        x = "sample_type", colour_by = "total_features", ncol = 3, size = 4,
        theme_size = 14) + ylab("Expression") + xlab(NULL) +
        theme(legend.text = element_text(size = 10),
              legend.title = element_text(size = 11))

    figplot2 <- cowplot::ggdraw() +
        cowplot::draw_plot(p1, 0, .62, 0.33, .38) +
        cowplot::draw_plot_label(letters[1], x = 0, y = 1, size = 24) +
        cowplot::draw_plot(p2, 0.33, .62, 0.33, .38) +
        cowplot::draw_plot_label(letters[2], x = 0.33, y = 1, size = 24) +
        cowplot::draw_plot(p3, 0.66, .62, 0.33, .38) +
        cowplot::draw_plot_label(letters[3], x = 0.66, y = 1, size = 24) +
        cowplot::draw_plot(p4, 0, .27, 0.33, .35) +
        cowplot::draw_plot_label(letters[4], x = 0, y = 0.65, size = 24) +
        cowplot::draw_plot(p5, 0.33, .27, 0.33, .35) +
        cowplot::draw_plot_label(letters[5], x = 0.33, y = 0.65, size = 24) +
        cowplot::draw_plot(p6, 0.66, .27, 0.33, .35) +
        cowplot::draw_plot_label(letters[6], x = 0.66, y = 0.65, size = 24) +
        cowplot::draw_plot(plt_exprs, 0, 0, 1, 0.27) +
        cowplot::draw_plot_label(letters[7],  x = 0, y = 0.3, size = 24)
    figplot2

    cowplot::save_plot("../figures/figure4.pdf", figplot2, nrow = 3, ncol = 3,
                       base_height = 3.5, base_aspect_ratio = 1.2)
    cowplot::save_plot("../figures/figure4.png", figplot2,  ncol = 3, nrow = 3,
                       base_height = 3.5, base_aspect_ratio = 1.2)
    cowplot::save_plot("../figures/figure4.jpg", figplot2,  ncol = 3, nrow = 3,
                       base_height = 3.5, base_aspect_ratio = 1.2, dpi = 350)

    if ( requireNamespace("RUVSeq")) {
        library( "RUVSeq" )
        scDataRUV <- filter(scData, use)
        scDataRUV <- scDataRUV[fData(scData)$use | fData(scData)$is_feature_control,]
        scIdx <- matrix(-1, ncol = max(table(scDataRUV$sample_type)), nrow = 2)
        scIdx[1,] <- which(scDataRUV$sample_type == "cell from patient B")
        tmp <- which(scDataRUV$sample_type == "cell from patient A")
        scIdx[2, 1:length(tmp)] <- tmp
        cIdx <- featureNames(scDataRUV)
        ## RUVs
        ruvs <- RUVs(counts(scDataRUV), cIdx, k = 1, scIdx = scIdx, isLog = FALSE)
        norm_counts(scDataRUV) <- ruvs$normalizedCounts
        exprs(scDataRUV) <- edgeR::cpm.default(norm_counts(scDataRUV), log = TRUE, 
                                               prior.count = scDataRUV@logExprsOffset)
        plt_norm_ruvs <- plotPCA(
            scDataRUV, exprs_values = "exprs", shape_by = "c1_machine", 
            colour_by = "sample_type", size_by = "total_features", 
            scale_features = TRUE) + 
            ggtitle(paste("PCA - RUVs normalisation: k =", 1)) +
            theme(legend.position = "bottom")

        ## RUVg
        cIdx <- (fData(scDataRUV)$is_feature_control & fData(scDataRUV)$n_cells_exprs >= 10)
        ruvg <- RUVg(counts(scDataRUV), cIdx, k = 1, isLog = FALSE)
        norm_counts(scDataRUV) <- ruvs$normalizedCounts
        exprs(scDataRUV) <- edgeR::cpm.default(norm_counts(scDataRUV), log = TRUE, 
                                               prior.count = scDataRUV@logExprsOffset)
        plt_norm_ruvg <- plotPCA(
            scDataRUV, exprs_values = "exprs", shape_by = "c1_machine", 
            colour_by = "sample_type", size_by = "total_features", 
            scale_features = TRUE) + 
            ggtitle(paste("PCA - RUVg normalisation: k =", 1)) +
            theme(legend.position = "bottom")

        plt_pca_endog_norm <- plt_pca_endog_norm + 
            theme(legend.position = "none")

        plt_sf_norm_resid <- plt_sf_norm_resid + 
            theme(legend.position = "none")

        figplot3 <- cowplot::plot_grid(plt_pca_endog_norm, plt_sf_norm_resid,
                                       plt_norm_ruvs,
                                       labels = letters, ncol = 1,
                                       rel_heights = c(0.75, 0.75, 1), label_size = 18)
        cowplot::save_plot("../figures/figure5.pdf", figplot3, ncol = 1, nrow = 3,
                           base_height = 3.5, base_aspect_ratio = 1.2)
        cowplot::save_plot("../figures/figure5.png", figplot3, ncol = 1, nrow = 3,
                           base_height = 3.5, base_aspect_ratio = 1.2)
        cowplot::save_plot("../figures/figure5.jpg", figplot3, ncol = 1, nrow = 3,
                           base_height = 3.5, base_aspect_ratio = 1.2, dpi = 350)

    }

}

```

### 2016-08-12 supplement_figs.tex
First commit

\documentclass{revtex4}
\usepackage{hyperref}
\usepackage{graphicx}
\begin{document}

\title[scater supplementary figures]{Supplementary Figures---``\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R''}
\author{Davis J.~McCarthy\,$^{\text{1,2,4,6}*}$, Kieran R.~Campbell\,$^{\text{2,5}}$, Aaron T.~L.~Lun\,$^{\text{7}}$ and Quin F.~Wills\,$^{\text{2,3}}$}
\address{$^{\text{1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, Cambridgeshire, UK;\\
$^{\text{2}}$Wellcome Trust Centre for Human Genetics, University of Oxford,
Oxford, Oxfordshire, UK;\\
$^{\text{3}}$ Weatherall Institute for Molecular Medicine, University of Oxford, Oxford, Oxfordshire, UK;\\
$^{\text{4}}$Department of Statistics, University of Oxford, Oxford, Oxfordshire, UK;\\
$^{\text{5}}$Department of Anatomy, Physiology and Genetics, University of Oxford, Oxford, Oxfordshire, UK;\\
$^{\text{6}}$St Vincent's Institute of Medical Research, 41 Victoria Parade Fitzroy Victoria 3065, Australia; and \\
$^{\text{7}}$CRUK Cambridge Institute, Robinson Way, Cambridge CB2 0RE,
Cambridgeshire, UK.}

\begin{abstract}
Supplementary Figures---an overview of the SCESet class; an overview of the \emph{scater} ecosystem; examples of using the \emph{scater} GUI.\\ \\
\textbf{Contact:} \href{davis@ebi.ac.uk}{davis@ebi.ac.uk}
\end{abstract}

\maketitle

\section*{Supplementary Figures}

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics[width=0.95\textwidth]{figures/sceset_outline.pdf}}
\caption{An overview of the SCESet class that underpins the \emph{scater} package. Building on Bioconductor's ExpressionSet class, it is a fully-featured, sophisticated and flexible data class tailored to scRNA-seq data.}\label{fig:02}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.95\textwidth]{figures/scater_ecosystem.pdf}}
\caption{An overview of the \emph{scater} ecosystem. The SCESet class in \emph{scater} acts as a convenient hub for datasets so that many other methods and tools implemented in R can be applied.}\label{fig:scater-eco}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.95\textwidth]{figures/scater_gui_landing_page.pdf}}
\caption{The landing page for the \emph{scater} graphical user interface (GUI).}\label{fig:scater-gui-landing}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.95\textwidth]{figures/scater_gui_plotqc.pdf}}
\caption{The \texttt{plotQC} page for the \emph{scater} graphical user interface (GUI).}\label{fig:scater-plotqc}
\end{figure}

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.95\textwidth]{figures/scater_gui_pca_qc.pdf}}
\caption{The \texttt{plotPCA - QC} page for the \emph{scater} graphical user interface (GUI).}\label{fig:scater-pca-qc}
\end{figure}

\section*{Details of package dependencies}

The package builds on many other R packages: \emph{Biobase} and \emph{BiocGenerics} for core Bioconductor functionality  \citep{Huber2015-en}; \emph{plyr} \citep{Wickham2015-kj}, \emph{reshape2} \citep{Wickham2012-ec}, \emph{dplyr} \citep{Wickham2015-la}, \emph{data.table} \citep{Dowle2015-zc} and \emph{magrittr} \citep{Bache2014-sa} for reading and
tidying data; \emph{ggplot2} \citep{Wickham2016-dc} for plotting; \emph{biomaRt} \citep{Durinck2005-yz} for feature annotation; \emph{edgeR} \citep{Robinson2010-ky} for computation of normalisation size factors and counts-per-million values; \emph{limma} \citep{Ritchie2015-so} for efficient fitting of linear models to features; \emph{rhdf5} \citep{Fischer2016-me}, \emph{rjson} \citep{Couture-Beil2014-kk} and \emph{tximport} \citep{Soneson2015-fw} for reading in transcript-level expression values;
\emph{viridis} \citep{Garnier2016-hk} for perceptually-uniform colour
maps for plotting; \emph{parallel} for parallel computation; \emph{matrixStats} \citep{Bengtsson2016-tn} for computation of summary statistics from matrices; \emph{cowplot} \citep{Wilke2016-hj} for attractive plotting themes; \emph{destiny} \citep{Angerer2015-sw} for producing diffusion maps; \emph{Rtsne} \citep{Krijthe2015-is} for producing t-SNE plots; \emph{mvoutlier}
\citep{Filzmoser2015-kx} for multivariate outlier detection from PCA of QC metrics; \emph{roxygen2} \citep{Wickham2015-pu}, \emph{BiocStyle} \citep{Huber2015-en}, \emph{knitr} \citep{Xie2013-bn} and \emph{rmarkdown} \citep{Allaire2016-dl} for generating documentation; and \emph{testthat} \citep{Wickham2011-cj} for unit testing. As well as functioning in the usual R environments, \emph{scater} also has a GUI built using
\emph{shiny} \citep{Chang2016-of} and \emph{shinydashboard} \citep{Chang2015-bn} for intuitive and interactive data visualisation. Calling the \verb|scater_gui| function from within an R session opens up the GUI in a web browser.

\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{plain}
%\bibliographystyle{bioinformatics}
\bibliography{main}

\end{document}

### 2016-08-14 main.tex
scater paper ready for submission

@@
 \author[McCarthy \textit{et~al}.]{Davis J.~McCarthy\,$^{\text{\sfb 1,2,5}*}$, Kieran R.~Campbell\,$^{\text{\sfb 2,4}}$, Aaron T.~L.~Lun\,$^{\text{\sfb 6}}$ and Quin F.~Wills\,$^{\text{\sfb 2,3}}$}
-\address{$^{\text{\sf 1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, Cambridgeshire, UK;\\
+\address{$^{\text{\sf 1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, United Kingdom;\\
 $^{\text{\sf 2}}$Wellcome Trust Centre for Human Genetics, University of Oxford,
-Oxford, Oxfordshire, UK;\\
-$^{\text{\sf 3}}$Weatherall Institute for Molecular Medicine, University of Oxford, Oxford, Oxfordshire, UK;\\
-$^{\text{\sf 4}}$Department of Physiology, Anatomy and Genetics, University of Oxford, Oxford, Oxfordshire, UK;\\
-$^{\text{\sf 5}}$St Vincent's Institute of Medical Research, 41 Victoria Parade Fitzroy Victoria 3065, Australia; and \\
-$^{\text{\sf 6}}$CRUK Cambridge Institute, Robinson Way, Cambridge CB2 0RE,
-Cambridgeshire, UK.
+Roosevelt Drive, Oxford OX3 7BN, United Kingdom;\\
+$^{\text{\sf 3}}$Weatherall Institute for Molecular Medicine, University of Oxford, John Radcliffe Hospital, Oxford OX3 9DS, United Kingdom;\\
+$^{\text{\sf 4}}$Department of Physiology, Anatomy and Genetics, University of Oxford, South Parks Road, Oxford OX1 3QX, United Kingdom;\\
+$^{\text{\sf 5}}$St Vincent's Institute of Medical Research, 41 Victoria Parade, Fitzroy Victoria 3065, Australia; and \\
+$^{\text{\sf 6}}$CRUK Cambridge Institute, University of Cambridge, Robinson Way, Cambridge CB2 0RE, United Kingdom.
 % $^{\text{\sf 4}}$Department of Statistics, University of Oxford, Oxford, Oxfordshire, UK;
@@
 One of the most popular frameworks for interactive analysis is the R programming language, extended for biological data analysis through the Bioconductor project \citep{Huber2015-en}.
-While Bioconductor packages have been widely used for bulk RNA-seq data, the existing data structures (like the ExpressionSet class) and methods are not sufficient for scRNA-seq data.
-This is because they do not support data types or processing methods that are specific to single-cell studies (e.g., cell-cell distance matrices for clustering, cell-based quality control).
-Extensions to the current computational infrastructure are required to add these capabilities for scRNA-seq analyses.
-
-Furthermore, large-scale single-cell studies will contain data types beyond the expression profiles, such as intensity values from fluorescence-activated cell sorting, cell imaging data, and information from epigenetic and targeted genotyping assays.
-Appropriate data structures and methods are required that can accommodate these rich datasets for integrative analyses of expression and other assay data along with the accompanying metadata.
-
-In addition, current visualisation methods designed for exploratory data analysis of bulk transcriptomic experiments are unsuited to scRNA-seq data sets containing hundreds or thousands of cells.
-The large size of each dataset also favours methods such as \emph{kallisto} \citep{Bray2016-pj} and \emph{Salmon} \citep{Patro2015-jf} for rapidly quantifying gene expression, and there is a need for new computational infrastructure to process raw scRNA-seq sequence data into a high-quality expression dataset ready for downstream analysis.
+While Bioconductor packages have been widely used for bulk RNA-seq data, the existing data structures (like the ExpressionSet class) are not sufficient for scRNA-seq data.
+This is because they do not support data types that are specific to single-cell studies, e.g., cell-cell distance matrices for clustering.
+For larger studies, this also includes data beyond expression profiles such as intensity values from fluorescence-activated cell sorting, cell imaging data, and information from epigenetic and targeted genotyping assays.
+Existing methods for processing and applying quality control to scRNA-seq data are similarly inadequate.
+In particular, current visualisation methods designed for exploratory data analysis of bulk transcriptomic experiments are unsuited to datasets containing hundreds or thousands of cells.
+The large size of each dataset also favours methods such as \emph{kallisto} \citep{Bray2016-pj} and \emph{Salmon} \citep{Patro2015-jf} for rapidly quantifying gene expression.
+Extensions to the current computational infrastructure are required to provide appropriate data structures and methods that can accommodate these rich scRNA-seq datasets for integrative analyses of expression and other assay data along with the accompanying metadata.
 
@@
 Samples were sequenced with paired-end sequencing using the HiSeq 2500 Sequencing system (Illumina).
-RNA-seq reads were were mapped to a custom genome reference, consisting of Homo sapiens GRCh37 (primary assembly from \href{ftp://ftp.ensembl.org/pub/release-75/fasta/homo_sapiens/dna/}{ftp://ftp.ensembl.org/pub/release-75/fasta/homo_sapiens/dna/}, last accessed 14/08/2015), Epstein-Barr Virus type 1 (B95-8 strain, Accession NC\_007605.1) and ERCC RNA spike-ins (ThermoFisher).
-Reads in fastq format were aligned using TopHat2 v2.0.12 \citep{Kim2013-qb} using Bowtie2 v2.2.3.0 \citep{Langmead2012-yc} as the core-mapping engine (\verb|--mate-inner-dist 190 --mate-std-dev 40 --report-secondary-alignments|) and other default parameters.
+RNA-seq reads were were mapped to a custom genome reference, consisting of Homo sapiens GRCh37 (primary assembly from \href{ftp://ftp.ensembl.org/pub/release-75/fasta/homo_sapiens/dna/}{ftp.ensembl.org/pub/release-75/fasta/homo\_sapiens/dna/}, last accessed 14.08.2015), Epstein-Barr Virus type 1 (B95-8 strain, Accession NC\_007605.1) and ERCC RNA spike-ins (ThermoFisher).
+Reads in fastq format were aligned using TopHat2 v2.0.12 \citep{Kim2013-qb} using Bowtie2 v2.2.3.0 \citep{Langmead2012-yc} as the core-mapping engine (\verb|--mate-inner-dist 190| \verb|--mate-std-dev 40| \verb|--report-secondary-alignments|) and other default parameters.
 Potential PCR duplicates were marked with Picard MarkDuplicates v1.92(1464).
@@
 \begin{figure*}[!htpb]%figure3
-\centerline{\includegraphics[width=\textwidth]{figures/figure3.pdf}}
-\caption{\textbf{Different types of QC plot that can be generated with \emph{scater}.} (\textbf{a}) Cumulative expression plot showing the proportion of the library accounted for by the top 1--500 most highly-expressed features.
-(\textbf{b)} PCA plot produced using a subset of the QC metrics computed with \emph{scater's} calculateQCMetrics function.
-(\textbf{c)} Plot of the 50 most-expressed features (computed according to the highest total read counts) across all cells in the data set.
+\centerline{\includegraphics[width=\textwidth]{figures/figure3_alt.pdf}}
+\caption{\textbf{Different types of QC plots that can be generated with \emph{scater}.} (\textbf{a}) Cumulative expression plot showing the proportion of the library accounted for by the top 1--500 most highly-expressed features.
+(\textbf{b}) PCA plot produced using a subset of the QC metrics computed with \emph{scater's} calculateQCMetrics function.
+(\textbf{c}) Plot of frequency of expression (percentage of cells in which the feature is deemed expressed) against mean expression level across cells.
+The vertical dotted line shows the median of the gene mean expression levels, and the horizontal dotted line indicates 50\% frequency of expression.
+(\textbf{d}) Plot of the 20 most highly-expressed features (computed according to the highest total read counts) across all cells in the data set.
 For each feature, the circle represents the percentage of counts that the gene accounts for computed from counts pooled across all cells.
@@
 The bars for each gene show the percentage of counts accounted for by the gene for each individual cell, providing a visualisation of the distribution across cells.
-(\textbf{d}) Plot of frequency of expression (percentage of cells in which the feature is deemed expressed) against mean expression level across cells.
 (\textbf{e}) Density plot showing the percentage of variance explained by a set of explanatory variables across all genes.
-(\textbf{f}) Violin, scatter- and boxplots of principal component values against the C1 machine used for each cell for the six principal components most strongly correlated with C1 machine used.
-Each individual plot is produced by a single call with either the function plot (\textbf{a}), plotPCA (\textbf{b}) or plotQC (\textbf{c-f}).}\label{fig:03}
+Each individual plot is produced by a single call with either the function plot (\textbf{a}), plotPCA (\textbf{b}) or plotQC (\textbf{c--e}).}\label{fig:03}
 \end{figure*}
@@
 Cell-specific QC metrics include the total count across all genes, the total number of expressed genes, and the percentage of counts allocated to control genes like spike-in transcripts or mitochondrial genes.
-These metrics are useful for identifying low-quality cells---for example, a high percentage of counts mapping to spike-ins typically indicates a small amount of RNA captured for the cell, suggesting protocol failure or death of the cell in processing and means it is unlikely to be suitable for downstream analyses.
+These metrics are useful for identifying low-quality cells---for example, a high percentage of counts mapping to spike-ins typically indicates that a small amount of RNA was captured for the cell, suggesting protocol failure or death of the cell in processing that renders it unsuitable for downstream analyses.
 For each gene, QC metrics such as the average expression level and the proportion of cells in which the gene is expressed are computed.
@@
 low total counts and few expressed genes) that should be removed prior to
-downstream analysis. This automated approach is powerful but also difficult to interpret, and so complements simpler filtering approaches that apply thresholds to particular QC metrics.
+downstream analysis.
+This automated approach is powerful but also somewhat opaque with respect to how outliers are defined, and so complements simpler filtering approaches that apply thresholds to particular QC metrics.
 
-The plotQC function generates many types of plots useful for quality control, such as a plot of visualising the most
-highly-expressed features in the dataset (Figure~\ref{fig:03}c). This
-provides a feature-centric overview of the dataset that simultaneously
-visualises the features with highest total expression across all cells,
+The plotQC function generates many types of plots useful for quality control,
+such as a plot to visualise the frequency of expression of features against their average expression level (Figure~\ref{fig:03}c).
+Such plots are useful, because scRNA-seq is characterised by a high frequency of ``dropout'' events,
+that is, no observed expression (such as no read counts) in a particular
+cell for a gene that is actually expressed in that cell. Typically only
+a small set of genes are observed with detectable expression in every
+cell. With plotQC, control features can be highlighted easily in the plot, and
+typical scRNA-seq datasets will show a broadly sigmoidal relationship between
+average expression level and frequency of expression across cells. This is
+consistent with expected behaviour where genes with greater average expression
+are more readily captured during library preparation and are detected at a
+greater frequency \citep{Brennecke2013-zv,Kim2015-xd,Vallejos2015-ww}.
+
+With plotQC we can also produce a plot to visualise the most
+highly-expressed features in the dataset (Figure~\ref{fig:03}d). This
+provides a feature-centric overview of the dataset that visualises the features
+with highest total expression across all cells,
 while also displaying the distribution of cell-level expression values
@@
 while datasets consisting of healthy cells will also show high levels of
-constitutively expressed genes like \emph{GAPDH} and \emph{ACTB}. This
+constitutively expressed genes like \emph{ACTB}. This
 plot allows the analyst to quickly check that the gene- or
@@
 datasets where it is not.
-
-A key feature of scRNA-seq is the high frequency of ``dropout'' events,
-that is, no observed expression (such as no read counts) in a particular
-cell for a gene that is actually expressed in that cell. Typically only
-a small set of genes are observed with detectable expression in every
-cell. The plotQC function can also be used to visualise the frequency of
-expression (inverse of the dropout percentage) of features
-against their average expression level (Figure~\ref{fig:03}d). Control features
-can be highlighted easily in the plot, and typical scRNA-seq datasets will show
-a broadly sigmoidal relationship between average expression level and
-frequency of expression across cells. This is consistent with expected
-behaviour where genes with greater average expression are more readily
-captured during library preparation and are detected at a greater
-frequency \citep{Brennecke2013-zv,Kim2015-xd,Vallejos2015-ww}.
 
@@
 Study data, the first principal component is correlated with the
-C1 machine used to process the cell (Figure~\ref{fig:03}f).
+C1 machine used to process the cell (Supplementary Figure~2).
 
@@
 
+\begin{figure*}[!htpb]%figure4
+\centerline{\includegraphics[width=\textwidth]{figures/figure4.pdf}}
+\caption{\textbf{Reduced dimension representations of cells and gene expression plots with \emph{scater.}} Plots are shown using all genes (\textbf{a-c}) and cell cycle genes only (\textbf{d-f}) using PCA (\textbf{a,d}), t-SNE (\textbf{b,e}) and diffusion maps (\textbf{c,f}), where each point represents a cell.
+In the top row (\textbf{a-c}), points are coloured by patient of origin, sized by total features (number of genes with detectable expression) and the shape indicates the C1 machine used to process the cells.
+In the second row (\textbf{d-f}), points are coloured by the expression of \emph{CCND2} (a gene associated with the G1/S phase transition of the cell cycle) in each cell.
+With the plotExpression function, gene expression can be plotted against any cell metadata variables or the expression of another gene---here, expression for the CD86, IGH44 and IGHV4-34 genes in each cell is plotted against the patient of origin (\textbf{g}).
+The function automatically detects whether the x-axis variable is categorical or continuous and plots the data accordingly, with x-axis values ``jittered'' to avoid excessive overplotting of points with the same x coordinate.}\label{fig:04}
+\end{figure*}
+
 The \emph{scater} graphical user interface (GUI) provides convenient
 access to \emph{scater}'s QC and visualisation methods (Supplementary
-Figures~3--5).
+Figures~4--6).
 This opens an interactive interface in a web browser that facilitates exploration of the data through QC plots and other intuitive visualisations.
@@
 exploring processes like the cell cycle, which can have substantial effects on
-single-cell expression measurements \citep{Buettner2015-jg}. The subsetting
-and filtering methods for SCESet objects make it easy to construct
+single-cell expression measurements \citep{Buettner2015-jg}. The subsetting and
+filtering methods for SCESet objects facilitate the generation of
 reduced-dimension plots for particular gene sets, in order to
@@
 necessarily lost.
-
-\begin{figure}[!tpb]%figure4
-\centerline{\includegraphics[width=0.48\textwidth]{figures/figure4.pdf}}
-\caption{\textbf{Reduced dimension representations of cells and gene expression plots with \emph{scater.}} Plots are shown using all genes (\textbf{a-c}) and cell cycle genes only (\textbf{d-f}) using PCA (\textbf{a,d}), t-SNE (\textbf{b,e}) and diffusion maps (\textbf{c,f}).
-In the top row (\textbf{a-c}), points are coloured by patient of origin, sized by total features (number of genes with detectable expression) and the shape indicates the C1 machine used to process the cells.
-In the second row (\textbf{d-f}), points are coloured by the expression of \emph{CCND2} (ENSG00000118971), a gene associated with the G1/S phase transition of the cell cycle.
-With the plotExpression function, expression for sets of genes can be plotted against any cell metadata variables or the expression of a given gene (\textbf{g}).
-The function automatically detects whether the x-axis variable is categorical or continuous and plots the data accordingly, with x-axis values ``jittered'' to avoid excessive overplotting of points with the same x coordinate.}\label{fig:04}
-\end{figure}
 
@@
 counts from the top 100 most highly-expressed genes and percentage of
-counts from control genes as additional covariates.) We then use the
+counts from control genes as additional covariates to control for these other unwanted technical effects.) We then use the
 residuals from the fitted model for further analyses (see Case Study in
@@
 
-\begin{figure}[!tpb]%figure5
-\centerline{\includegraphics[width=0.48\textwidth]{figures/figure5.pdf}}
-\caption{\textbf{Figure 5: Normalisation approaches made easy with
+\begin{figure*}[!tpb]%figure5
+\centerline{\includegraphics[width=\textwidth]{figures/figure5.pdf}}
+\caption{\textbf{Figure 5: Normalisation and batch correction with
 \emph{scater}.} Principal component analysis plots showing cell
@@
 expression in the cell.}\label{fig:05}
-\end{figure}
+\end{figure*}
 
@@
 As part of the R/Bioconductor ecosystem, \emph{scater} can be easily
-integrated with other software for scRNA-seq data analysis (Supplementary Figure~2). Because the
+integrated with other software for scRNA-seq data analysis (Supplementary Figure~3). Because the
 SCESet class builds on existing Bioconductor data structures, most
@@
 \citep{Vallejos2016-qy,Trapnell2014-gj,Finak2015-rd,Vu2016-sk,Kharchenko2014-rx,Korthauer2015-wg,Andrews2016-at}, heterogeneous gene expression analyses
-\citep{Lun2016-sk,Vallejos2015-ww}, clustering
+\citep{Vallejos2015-ww}, clustering
 \citep{Kiselev2016-fu,Guo2015-wx,Fan2016-wh,Grun2015-fy}, latent or hidden
@@
 scRNA-seq data, its capabilities are well suited for single-cell qPCR data
-and bulk RNA-seq data, and will prove useful for supporting analyses of these data types too.
+and bulk RNA-seq data, and may prove useful for supporting analyses of these data types.
 
@@
 
+\section*{Funding}
+
+This work was supported by the National Health and Medical Research Council of Australia [Early Career Fellowship to D.J.M.], by core funding from the European Molecular Biology Laboratory [D.J.M.], core funding from Cancer Research UK [A17197 to A.T.L.L.), the United Kingdom Medical Research Council [studentship to K.R.C.) and the Oxford Single Cell Biology Consortium [Q.F.W.].\vspace*{-12pt}
+
 \section*{Acknowledgements}\label{acknowledgements}
@@
 We would also like to acknowledge Peter Donnelly and Oliver Stegle for support and helpful discussions.\vspace*{-12pt}
-
-\section*{Funding}
-
-DJM was supported by funding through an Early Career Fellowship from the National Health and Medical Research Council of Australia, and by core funding from EMBL.
-ATLL was supported by core funding from Cancer Research UK (award no.
-A17197).
-KRC is supported by a UK Medical Research Council studentship.\vspace*{-12pt}
 

### 2016-08-14 scater_case_study/case_study.Rmd
scater paper ready for submission

@@
 title: "SCATER CASE STUDY"
-author: Davis McCarthy & Quin Wills
+author: Davis McCarthy, Kieran Campbell, Aaron Lun & Quin Wills
 date: "`r Sys.Date()`"
@@
         number_sections: false
-        code_folding: hide
+        code_folding: show
 ---
@@
 library( "RUVSeq" )
+library( "BiocStyle" )
+options(stringsAsFactors = FALSE)
 ```
@@
 6. **DOWNSTREAM ANALYSIS:** An example of further analysis.
-7. **TECHNICAL STUFF:** Information for bioinformaticians, such as the R session used to run this case study.
+7. **SESSION INFORMATION:** Information for bioinformaticians, such as the R session used to run this case study.
 
@@
 
-### 1. OVERVIEW
+## 1. OVERVIEW
 
@@
 
-### 2. GETTING STARTED
+## 2. GETTING STARTED
 
@@
 
-In keeping with best practice in object-oriented bioinformatics, `scater` not only allows you to store your data in a standardised format, but also to associate it with experimental and bioinformatic meta-data.
+In keeping with best practice in object-oriented bioinformatics, `r Biocpkg("scater")` not only allows you to store your data in a standardised format, but also to associate it with experimental and bioinformatic meta-data.
 Experimental meta-data are any information associated with individual cells (also called 'samples'), from sequencing metrics to experimental batching and other experimental cell phenotypes.
@@
 These can be read in,  combined with the sequencing data,  and extracted as shown below.
-For this example we'll use the automated log-scaled counts per million generated by `scater`,  but see ```calculateTPM()``` as an example function for generating alternative RNA-seq read count formats.
-\
-
-View the first few entris for the cell metadata:
+For this example we'll use the automated log-scaled counts per million generated by `r Biocpkg("scater")`,  but see ```calculateTPM()``` as an example function for generating alternative RNA-seq read count formats.
+\
+
+View the first few entries for the cell metadata:
 
@@
 scData <- newSCESet( countData = reads, phenoData = cells, featureData = genes, lowerDetectionLimit = 3 ) 
-
-pData( scData )[1:3, 1:2] # view some sample meta-data
+head(pData(scData))  # view some sample meta-data
 ```
@@
 
-### 3. QC THE CELLS
+## 3. QC THE CELLS
 
@@
 \  
-Another option available in `scater` is to conduct PCA on a set of QC metrics.
+Another option available in `r Biocpkg("scater")` is to conduct PCA on a set of QC metrics.
 The advantage of doing this is that the QC metrics focus on technical aspects of
@@
 
-plotReducedDim(scData, colour_by = "outlier", shape_by = "use", ncomponents = 3)
-
-plotReducedDim(scData[, grep("cell", scData$sample_type)], 
-               colour_by = "outlier", shape_by = "use", ncomponents = 3)
-
 plotPCA( scData[, grep("cell", scData$sample_type)], colour_by = "outlier", 
@@
 
-With `scater`, any specific set of features based on prior knowledge can be used for PCA, t-SNE or diffusion maps.
+With `r Biocpkg("scater")`, any specific set of features based on prior knowledge can be used for PCA, t-SNE or diffusion maps.
 A feature set to use can be defined by supplying the `feature_set` argument to `plotPCA`, `plotTSNE` or `plotDiffusionMap`.
@@
                                fData(scData)$total_feature_counts > 100000)
-knitr::kable(fData(scData)[most_exprs_ccycle,] )
 p1 <- plotPCA(scData, feature_set = cell_cycle_genes, 
@@
 
-One element of QC that can prove challenging to judge from the above criteria is if more than one cell ( or cell debris from other cells ) are being sequenced.
+One element of QC that can prove challenging to judge from the above criteria is if more than one cell ( or cell debris from other cells ) is being sequenced.
 With most cell processing platforms a small percentage of cells won't be truly single.
@@
 
-### 4. QC THE GENES
+## 4. QC THE GENES
 
@@
 
-It can also be useful to look at total expression levels acrsoo gene biotypes. 
+It can also be useful to look at total expression levels across gene biotypes. 
 The plot below shows the distributions of log-10 scale total gene counts for 
@@
                  median(log10_total_feature_counts))[,2]), decreasing = TRUE)
-fData(scData)$biotype <- factor(fData(scData)$gene_biotype, 
-                                levels = levels(fData(scData)$gene_biotype)[oo])
+fData(scData)$biotype <- factor(
+    fData(scData)$gene_biotype, 
+    levels = levels(as.factor(fData(scData)$gene_biotype))[oo])
 
@@
 ```{r append-gene-symbols}
+fData(scData)$gene_symbol <- fData(scData)$hgnc_symbol
+fData(scData)$gene_symbol[is.na(fData(scData)$gene_symbol)] <-
+    featureNames(scData)[is.na(fData(scData)$gene_symbol)]
 featureNames(scData) <- paste(fData(scData)$hgnc_symbol, featureNames(scData),
@@
         type = "highest-expression", 
-        col_by_variable = "sample_type" )
+        col_by_variable = "sample_type", feature_names_to_plot = "gene_symbol")
 ```
@@
 
-### 5. NORMALISE THE DATA
+## 5. NORMALISE THE DATA
 
@@
 Experimental design is a critical, but neglected, aspect of \gls{scrnaseq} studies.
-To the best of my knowledge, methods like those described in this section for exploring experimental and QC variables and the experimental design, do not feature in any scRNA-seq software packages apart from `scater`.
+To the best of our knowledge, methods like those described in this section for exploring experimental and QC variables and the experimental design, do not feature in any scRNA-seq software packages apart from `r Biocpkg("scater")`.
 There are a very large number of potential confounders, artifacts and biases in \gls{scrnaseq} studies.
@@
 The \scater\ package provides a set of methods specifically for quality control of experimental and explanatory variables, which will be demonstrated briefly here.
+
+### Finding important explanatory variables
 
@@
 
-In `scater` it is easy to perform size-factor normalisation using the "TMM" approach (the default in the [edgeR]() package), the "RLE" approach (default in the [DESeq]() package), an "upper-quartile approach" (as proposed by Bullard et al (2010)) or simple scaling by total counts.
+### Scaling normalisation
+
+In `r Biocpkg("scater")` it is easy to perform scaling normalisation using the "TMM" approach (the default in the [edgeR]() package), the "RLE" approach (default in the [DESeq]() package), an "upper-quartile approach" (as proposed by Bullard et al (2010)) or simple scaling by total counts.
 Size-factor normalisation can be carried out on a subsetted SCESet object, normalising either to ERCC spike-in genes or to endogenous genes if desired.
@@
 for scRNA-seq data. This method performs much better on single-cell data and its
-implementation in the Bioconductor package `scran` allows seamless integration 
-into the `scater` workflow. (The `scran` package itself depends on `scater`).
-
-We recommend using the `scran` size-factor normalisation approach and 
+implementation in the Bioconductor package `r Biocpkg("scran")` allows seamless integration 
+into the `r Biocpkg("scater")` workflow. (The `r Biocpkg("scran")` package itself depends on `r Biocpkg("scater")`).
+
+We recommend using the `r Biocpkg("scran")` size-factor normalisation approach and 
 demonstrate its use here.
@@
 
-scDataQC_spikenorm <- computeSumFactors(scDataQC, sizes = 10, 
-                              get.spikes = TRUE)
+isSpike(scDataQC) <- "spikes"
+scDataQC_spikenorm <- computeSpikeFactors(scDataQC)
 summary(scDataQC_spikenorm$size_factor)
@@
 We would prefer such technical effects to be removed by normalisation, so below we demonstrate a further alternative approach.
-We see how "customised" normalisation approaches can be easily incorporated into the `scater` workflow.
-
-In this example we normalise and standardise counts conditioned on expression level, showing the PCA for the corrected data.
+We see how "customised" normalisation approaches can be easily incorporated into the `r Biocpkg("scater")` workflow.
+
+### Expression-conditional normalisation
+
+In this example we normalise (by cell) and standardise (by gene) counts conditioned on expression level, showing the PCA for the corrected data.
 One advantage of this approach is that a biologically 'noisy' gene is naturally defined as one with greater dispersion than other genes at a similar expression level.
@@
 
+### Regressing out known sources of variation
+
 Another option is to regress out effects of technical factors in the 
@@
 
-```{r regress-factors-norm,  echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align="center", fig.width=6, fig.height=9}
+```{r regress-factors-norm,  echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE, fig.align="center", fig.width=6, fig.height=6}
 design <- model.matrix(~scDataQC$log10_counts_endogenous_features +
@@
 the first principal component now separates the two patients). Thus, for 
-removing known effects, the regression approach implemented in `scater` is 
+removing known effects, the regression approach implemented in `r Biocpkg("scater")` is 
 effective, but more needs to be done to find _hidden_ effects such as unknown
 batch effects, or similar.
+
+### Normalisation with `removeBatchEffect` from `limma`
 
@@
 
+### Batch correction with `RUVSeq`
+
 Look at the performance of the RUVs method:
@@
 ruv_list <- list()
-for (k in 1:9) {
+for (k in 1:4) {
     ruvs <- RUVs(counts(scDataRUV), cIdx, k = k,
@@
         ggtitle(paste("PCA - RUVs normalisation: k =", k)) + 
-        theme_bw(8)
+        theme_bw(8) + theme(legend.position = "bottom")
     ruv_list[[k]] <- plt_norm_ruv
@@
 
-figplot_ruv <- cowplot::plot_grid(plotlist = ruv_list[1:8], labels = letters[1:8], ncol = 2)
+figplot_ruv <- cowplot::plot_grid(plotlist = ruv_list, labels = letters, ncol = 2)
 figplot_ruv
-figplot_ruv <- cowplot::plot_grid(plotlist = ruv_list, labels = letters[1:9])
 cowplot::save_plot("../figures/supplement/ruvs_figure.png", figplot_ruv, 
@@
 
+Just removing the first latent variable found with RUVs now effectively splits cells from the two patients in the first principal component of the adjusted expression values.
+
+The QC plot below confirms that the first principal component is now highly correlated with patient.
+
 ```{r ruv-seq-s-qc-plots, echo=TRUE, eval=TRUE, message=FALSE, results='hide'}
-plotQC(scDataRUV, type = "explanatory-variables")
 plotQC(scDataRUV, type = "find-pcs", variable = "sample_type")
 ```
-
-```{r ruvseq-red-dim-k6}
-ruvs <- RUVs(counts(scDataRUV), cIdx, k = 6, scIdx = scIdx, isLog = FALSE)
-reducedDimension(scDataRUV) <- ruvs$W
-plotReducedDim(scDataRUV, shape_by = "c1_machine", 
-               colour_by = "sample_type", size_by = "total_features") + theme_bw()
-
-```
-
-Look at the performance of the RUVg method:
-
-```{r ruv-seq-g, echo=TRUE, eval=TRUE, fig.height=14, fig.width=9}
-# One approach to finding and removing hidden, unwanted sources of variability is
-# the `RUVSeq` package.
-## Running RUVs
-cIdx <- fData(scDataRUV)$is_feature_control
-ruvg_list <- list()
-for (k in 1:9) {
-    ruvs <- RUVg(counts(scDataRUV), cIdx, k = k, isLog = FALSE)
-    norm_counts(scDataRUV) <- ruvs$normalizedCounts
-    exprs(scDataRUV) <- edgeR::cpm.default(norm_counts(scDataRUV), log = TRUE, 
-                                           prior.count = scDataRUV@logExprsOffset)
-    plt_norm_ruv <- plotPCA(
-        scDataRUV, exprs_values = "exprs", shape_by = "c1_machine", 
-        colour_by = "sample_type", size_by = "total_features", 
-        scale_features = TRUE) + 
-        ggtitle(paste("PCA - RUVg normalisation: k =", k)) + 
-        theme_bw(8)
-    ruvg_list[[k]] <- plt_norm_ruv
-}
-
-figplot_ruvg <- cowplot::plot_grid(plotlist = ruvg_list[1:8], labels = letters[1:8],
-                                   ncol = 2)
-figplot_ruvg
-figplot_ruvg <- cowplot::plot_grid(plotlist = ruvg_list, labels = letters[1:9])
-cowplot::save_plot("../figures/supplement/ruvg_figure.png", figplot_ruv, 
-                   ncol = 3, nrow = 3, base_height = 5, base_aspect_ratio = 0.9)
-
-```
-
-```{r ruv-g-qc-plots, message=FALSE, results='hide'}
-plotQC(scDataRUV, type = "explanatory-variables")
-plotQC(scDataRUV, type = "find-pcs", variable = "sample_type")
-scDataRUV$c1_machine <- as.factor(scDataRUV$c1_machine)
-plotQC(scDataRUV, type = "find-pcs", variable = "c1_machine")
-
-```
-
-Unfortunately, the RUVg method does not do a good job of removing the C1 machine effect.
-The machine used is still correlated with the first PC even after removing 9 hidden factors of unwanted variation.
 
@@
 
-### 7. DOWNSTREAM ANALYSIS
+## 6. DOWNSTREAM ANALYSIS
 
@@
 
-In `scater` the `plotExpression` function enables the convenient visualisation of expression values for a set of features.
+In `r Biocpkg("scater")` the `plotExpression` function enables the convenient visualisation of expression values for a set of features.
 Here, the expression values for the six most DE genes for expression frequency between patients are shown.
 The units for expression in the plot can be defined with the `exprs_values` argument (the expression values must exist with the provided name in the `assayData` slot of the SCESet object; if not the default `exprs` values will be used, with a warning).
-As with other plots in `scater` we can use phenotype data variables to define the colour and shapr of the points.
+As with other plots in `r Biocpkg("scater")` we can use phenotype data variables to define the colour and shapr of the points.
 
@@
 
-### 8.  TECHNICAL STUFF
+## 7. SESSION INFORMATION
 
@@
 ```{r figures-for-paper, echo=TRUE, results='hide'}
-### Figures for the paper
+## Figures for the paper
 if ( requireNamespace("cowplot") ) {
@@
         "bulk control"
-    p1 <- plot( scData, colour_by = "well_type", exprs_values = "counts")
+    p1 <- plot( scData, colour_by = "well_type", exprs_values = "counts") +
+        theme(legend.position = c(0.8, 0.25))
     p2 <- plotPCA( scData, shape_by = "sample_type", 
@@
     p3 <- plotQC( scData[fData( scData )$use, pData( scData )$use], 
-        type = "highest-expression", 
-        col_by_variable = "sample_type" ) + coord_cartesian(xlim = c(0, 6))
+        type = "highest-expression", feature_names_to_plot = "gene_symbol",
+        col_by_variable = "sample_type", n = 20) + coord_cartesian(xlim = c(0, 6)) +
+        theme(axis.text.y = element_text(size = 10),
+              axis.title = element_text(size = 11))
     p4 <- plotQC(scData, type = "exprs")
@@
                       "sample_type") ) + 
-        xlab(expression(paste(R ^ 2, " (% variance explained; log10-scale)"))) +
-    theme(legend.position = "bottom", legend.text = element_text(size = 7))
+        xlab(expression(paste(R ^ 2, " (% variance explained; log10-scale)")))
     p6 <- plotQC( scData[fData( scData )$use, scData$use], 
@@
 
+    ## alternative version for better readability
+    figplot1_alt <- cowplot::ggdraw() +
+        cowplot::draw_plot(p1, 0.01, .5, 0.28, .48) +
+        cowplot::draw_plot_label(letters[1], x = 0, y = 1, size = 20) +
+        cowplot::draw_plot(p2, 0.29, .5, 0.35, .48) +
+        cowplot::draw_plot_label(letters[2], x = 0.29, y = 1, size = 20) +
+        cowplot::draw_plot(p4, 0.64, .5, 0.36, .48) +
+        cowplot::draw_plot_label(letters[3], x = 0.63, y = 1, size = 20) +
+        cowplot::draw_plot(p3, 0.01, 0, 0.5, .48) +
+        cowplot::draw_plot_label(letters[4], x = 0, y = 0.49, size = 20) +
+        cowplot::draw_plot(p5, 0.52, 0, .48, .48) +
+        cowplot::draw_plot_label(letters[5], x = 0.515, y = 0.49, size = 20)
+
+    cowplot::save_plot("../figures/figure3_alt.pdf", figplot1_alt, ncol = 3, nrow = 2,
+                       base_height = 3.2, base_aspect_ratio = 1.4)
+    cowplot::save_plot("../figures/figure3_alt.png", figplot1_alt, ncol = 3, nrow = 2,
+                       base_height = 3.2, base_aspect_ratio = 1.4)
+    cowplot::save_plot("../figures/figure3_alt.jpg", figplot1_alt, ncol = 3, nrow = 2,
+                       base_height = 3.2, base_aspect_ratio = 1.4, dpi = 350)
+
+    ## save find-pcs plot to add to supplement
+    cowplot::save_plot("../figures/find-pcs_c1_machine.pdf", p6,
+                       base_height = 5, base_aspect_ratio = 1.1)
+    cowplot::save_plot("../figures/find-pcs_c1_machine.png", p6,
+                       base_height = 5, base_aspect_ratio = 1.1, dpi = 350)
+    cowplot::save_plot("../figures/find-pcs_c1_machine.jpg", p6,
+                       base_height = 5, base_aspect_ratio = 1.1, dpi = 350)
+
     ## dimension reduction figure
@@
     feats_to_plot <- as.character(fData(scDataQC)$hgnc_symbol[ww])
-
-    p1 <- plotPCA(scData[, scData$use], size_by = "total_features", 
+    featureNames(scData)[most_exprs_ccycle[7]] <- 
+        fData(scData)$hgnc_symbol[most_exprs_ccycle[7]]
+
+    p1 <- plotPCA(scData[, scData$use], 
                   colour_by = "sample_type", theme_size = 14,
@@
         ylab("Component 2") +
-        theme(legend.position = "bottom", legend.text = element_text(size = 10), 
-              legend.title = element_text(size = 11))
-    p2 <- plotTSNE(scData[, scData$use], size_by = "total_features", 
+        theme(legend.position = "none")
+    p2 <- plotTSNE(scData[, scData$use], 
                    colour_by = "sample_type", theme_size = 14,
@@
         ggtitle("t-SNE - all genes") + 
-        theme(legend.position = "bottom", legend.text = element_text(size = 10),
-              legend.title = element_text(size = 11))
-    p3 <- plotDiffusionMap( scData[, scData$use], size_by = "total_features", 
-                            colour_by = "sample_type", 
+        theme(legend.position = "none")
+    p3 <- plotDiffusionMap( scData[, scData$use], 
+                            colour_by = "sample_type",  rand_seed = 20160814,
                             shape_by = "c1_machine", theme_size = 14) + 
         ggtitle("Diffusion map - all genes") +
-        theme(legend.position = "bottom", legend.text = element_text(size = 10),
-              legend.title = element_text(size = 11))
+        theme(legend.position = "right")
     p4 <- plotPCA(scData[, scData$use], feature_set = cell_cycle_genes, 
                   colour_by = featureNames(scData)[most_exprs_ccycle[7]], 
-                  shape_by = "sample_type", theme_size = 14) + 
+                  theme_size = 14) + 
         ggtitle("PCA - cell cycle genes") + ylab("Component 2") +
-        geom_point(aes(colour = colour_by, shape = shape_by), alpha = 0.6, size = 4) +
-        theme(legend.position = "bottom", legend.text = element_text(size = 10),
-              legend.title = element_text(size = 11))
+        #geom_point(aes(colour = colour_by, shape = shape_by), alpha = 0.6, size = 4) +
+        theme(legend.position = "none")
     p5 <- plotTSNE(scData[, scData$use], feature_set = cell_cycle_genes, 
                    colour_by = featureNames(scData)[most_exprs_ccycle[7]], 
-                   shape_by = "sample_type", rand_seed = 20151225, theme_size = 14) + 
-        geom_point(aes(colour = colour_by, shape = shape_by), alpha = 0.6, size = 4) +
+                   rand_seed = 20160814, theme_size = 14) + 
+        #geom_point(aes(colour = colour_by, shape = shape_by), alpha = 0.6, size = 4) +
         ggtitle("t-SNE - cell cycle genes") + 
-        theme(legend.position = "bottom", legend.text = element_text(size = 10),
-              legend.title = element_text(size = 11))
+        theme(legend.position = "none")
     p6 <- plotDiffusionMap( scData[, scData$use], feature_set = cell_cycle_genes, 
                             colour_by = featureNames(scData)[most_exprs_ccycle[7]], 
-                            shape_by = "sample_type", theme_size = 14) + 
-        geom_point(aes(colour = colour_by, shape = shape_by), alpha = 0.6, size = 4) +
+                            theme_size = 14, rand_seed = 20160814) + 
+        #geom_point(aes(colour = colour_by, shape = shape_by), alpha = 0.6, size = 4) +
         ggtitle("Diffusion map - cell cycle genes") +
-        theme(legend.position = "bottom", legend.text = element_text(size = 10),
-              legend.title = element_text(size = 11))
+        theme(legend.position = "right")
 
@@
         scDataQC, feats_to_plot, exprs_values = "norm_exprs", 
-        x = "sample_type", colour_by = "total_features", ncol = 3, size = 4,
-        theme_size = 14) + ylab("Expression") + xlab(NULL) +
-        theme(legend.text = element_text(size = 10),
-              legend.title = element_text(size = 11))
+        x = "sample_type", colour_by = "total_features", ncol = 3, size = 3,
+        theme_size = 14) + ylab("Expression") + xlab(NULL)
 
     figplot2 <- cowplot::ggdraw() +
-        cowplot::draw_plot(p1, 0, .62, 0.33, .38) +
+        cowplot::draw_plot(p1, 0, .65, 0.29, .35) +
         cowplot::draw_plot_label(letters[1], x = 0, y = 1, size = 24) +
-        cowplot::draw_plot(p2, 0.33, .62, 0.33, .38) +
-        cowplot::draw_plot_label(letters[2], x = 0.33, y = 1, size = 24) +
-        cowplot::draw_plot(p3, 0.66, .62, 0.33, .38) +
-        cowplot::draw_plot_label(letters[3], x = 0.66, y = 1, size = 24) +
-        cowplot::draw_plot(p4, 0, .27, 0.33, .35) +
+        cowplot::draw_plot(p2, 0.3, .65, 0.29, .35) +
+        cowplot::draw_plot_label(letters[2], x = 0.3, y = 1, size = 24) +
+        cowplot::draw_plot(p3, 0.59, .65, 0.42, .35) +
+        cowplot::draw_plot_label(letters[3], x = 0.6, y = 1, size = 24) +
+        cowplot::draw_plot(p4, 0, .3, 0.29, .35) +
         cowplot::draw_plot_label(letters[4], x = 0, y = 0.65, size = 24) +
-        cowplot::draw_plot(p5, 0.33, .27, 0.33, .35) +
-        cowplot::draw_plot_label(letters[5], x = 0.33, y = 0.65, size = 24) +
-        cowplot::draw_plot(p6, 0.66, .27, 0.33, .35) +
-        cowplot::draw_plot_label(letters[6], x = 0.66, y = 0.65, size = 24) +
-        cowplot::draw_plot(plt_exprs, 0, 0, 1, 0.27) +
-        cowplot::draw_plot_label(letters[7],  x = 0, y = 0.3, size = 24)
+        cowplot::draw_plot(p5, 0.3, .3, 0.29, .35) +
+        cowplot::draw_plot_label(letters[5], x = 0.29, y = 0.65, size = 24) +
+        cowplot::draw_plot(p6, 0.59, .3, 0.37, .35) +
+        cowplot::draw_plot_label(letters[6], x = 0.59, y = 0.65, size = 24) +
+        cowplot::draw_plot(plt_exprs, 0.005, 0, 0.99, 0.29) +
+        cowplot::draw_plot_label(letters[7],  x = 0, y = 0.29, size = 24)
     figplot2
@@
     cowplot::save_plot("../figures/figure4.pdf", figplot2, nrow = 3, ncol = 3,
-                       base_height = 3.5, base_aspect_ratio = 1.2)
+                       base_height = 2.7, base_aspect_ratio = 2)
     cowplot::save_plot("../figures/figure4.png", figplot2,  ncol = 3, nrow = 3,
-                       base_height = 3.5, base_aspect_ratio = 1.2)
+                       base_height = 2.7, base_aspect_ratio = 2)
     cowplot::save_plot("../figures/figure4.jpg", figplot2,  ncol = 3, nrow = 3,
-                       base_height = 3.5, base_aspect_ratio = 1.2, dpi = 350)
+                       base_height = 2.7, base_aspect_ratio = 2, dpi = 350)
 
@@
             ggtitle(paste("PCA - RUVs normalisation: k =", 1)) +
-            theme(legend.position = "bottom")
-
-        ## RUVg
-        cIdx <- (fData(scDataRUV)$is_feature_control & fData(scDataRUV)$n_cells_exprs >= 10)
-        ruvg <- RUVg(counts(scDataRUV), cIdx, k = 1, isLog = FALSE)
-        norm_counts(scDataRUV) <- ruvs$normalizedCounts
-        exprs(scDataRUV) <- edgeR::cpm.default(norm_counts(scDataRUV), log = TRUE, 
-                                               prior.count = scDataRUV@logExprsOffset)
-        plt_norm_ruvg <- plotPCA(
-            scDataRUV, exprs_values = "exprs", shape_by = "c1_machine", 
-            colour_by = "sample_type", size_by = "total_features", 
-            scale_features = TRUE) + 
-            ggtitle(paste("PCA - RUVg normalisation: k =", 1)) +
-            theme(legend.position = "bottom")
+            theme(legend.position = "right")
 
@@
                                        plt_norm_ruvs,
-                                       labels = letters, ncol = 1,
-                                       rel_heights = c(0.75, 0.75, 1), label_size = 18)
-        cowplot::save_plot("../figures/figure5.pdf", figplot3, ncol = 1, nrow = 3,
-                           base_height = 3.5, base_aspect_ratio = 1.2)
-        cowplot::save_plot("../figures/figure5.png", figplot3, ncol = 1, nrow = 3,
-                           base_height = 3.5, base_aspect_ratio = 1.2)
-        cowplot::save_plot("../figures/figure5.jpg", figplot3, ncol = 1, nrow = 3,
-                           base_height = 3.5, base_aspect_ratio = 1.2, dpi = 350)
+                                       labels = letters, ncol = 3,
+                                       rel_widths = c(0.75, 0.75, 1), label_size = 18)
+        figplot3
+
+        cowplot::save_plot("../figures/figure5.pdf", figplot3, ncol = 3, nrow = 1,
+                           base_height = 2.7, base_aspect_ratio = 1.5)
+        cowplot::save_plot("../figures/figure5.png", figplot3, ncol = 3, nrow = 1,
+                           base_height = 2.7, base_aspect_ratio = 1.5)
+        cowplot::save_plot("../figures/figure5.jpg", figplot3, ncol = 3, nrow = 1,
+                           base_height = 2.7, base_aspect_ratio = 1.5, dpi = 350)
 

### 2016-08-14 supplement_figs.tex
scater paper ready for submission

@@
 \title[scater supplementary figures]{Supplementary Figures---``\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R''}
-\author{Davis J.~McCarthy\,$^{\text{1,2,4,6}*}$, Kieran R.~Campbell\,$^{\text{2,5}}$, Aaron T.~L.~Lun\,$^{\text{7}}$ and Quin F.~Wills\,$^{\text{2,3}}$}
-\address{$^{\text{1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, Cambridgeshire, UK;\\
+\author{Davis J.~McCarthy\,$^{\text{1,2,5}*}$, Kieran R.~Campbell\,$^{\text{2,4}}$, Aaron T.~L.~Lun\,$^{\text{6}}$ and Quin F.~Wills\,$^{\text{2,3}}$}
+\address{$^{\text{1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, United Kingdom;\\
 $^{\text{2}}$Wellcome Trust Centre for Human Genetics, University of Oxford,
-Oxford, Oxfordshire, UK;\\
-$^{\text{3}}$ Weatherall Institute for Molecular Medicine, University of Oxford, Oxford, Oxfordshire, UK;\\
-$^{\text{4}}$Department of Statistics, University of Oxford, Oxford, Oxfordshire, UK;\\
-$^{\text{5}}$Department of Anatomy, Physiology and Genetics, University of Oxford, Oxford, Oxfordshire, UK;\\
-$^{\text{6}}$St Vincent's Institute of Medical Research, 41 Victoria Parade Fitzroy Victoria 3065, Australia; and \\
-$^{\text{7}}$CRUK Cambridge Institute, Robinson Way, Cambridge CB2 0RE,
-Cambridgeshire, UK.}
+Roosevelt Drive, Oxford OX3 7BN, United Kingdom;\\
+$^{\text{3}}$Weatherall Institute for Molecular Medicine, University of Oxford, John Radcliffe Hospital, Oxford OX3 9DS, United Kingdom;\\
+$^{\text{4}}$Department of Physiology, Anatomy and Genetics, University of Oxford, South Parks Road, Oxford OX1 3QX, United Kingdom;\\
+$^{\text{5}}$St Vincent's Institute of Medical Research, 41 Victoria Parade, Fitzroy Victoria 3065, Australia; and \\
+$^{\text{6}}$CRUK Cambridge Institute, University of Cambridge, Robinson Way, Cambridge CB2 0RE, United Kingdom.
+% $^{\text{\sf 4}}$Department of Statistics, University of Oxford, Oxford, Oxfordshire, UK;
+}
 
 \begin{abstract}
-Supplementary Figures---an overview of the SCESet class; an overview of the \emph{scater} ecosystem; examples of using the \emph{scater} GUI.\\ \\
+Supplementary material and figures---details of package dependencies; an overview of the SCESet class; an overview of the \emph{scater} ecosystem; examples of using the \emph{scater} GUI.\\ \\
 \textbf{Contact:} \href{davis@ebi.ac.uk}{davis@ebi.ac.uk}
@@
 \maketitle
+
+\section*{Details of package dependencies}
+
+The package builds on many other R packages: \emph{Biobase} and \emph{BiocGenerics} for core Bioconductor functionality  \citep{Huber2015-en}; \emph{plyr} \citep{Wickham2015-kj}, \emph{reshape2} \citep{Wickham2012-ec}, \emph{dplyr} \citep{Wickham2015-la}, \emph{data.table} \citep{Dowle2015-zc} and \emph{magrittr} \citep{Bache2014-sa} for reading and
+tidying data; \emph{ggplot2} \citep{Wickham2016-dc} for plotting; \emph{biomaRt} \citep{Durinck2005-yz} for feature annotation; \emph{edgeR} \citep{Robinson2010-ky} for computation of normalisation size factors and counts-per-million values; \emph{limma} \citep{Ritchie2015-so} for efficient fitting of linear models to features; \emph{rhdf5} \citep{Fischer2016-me}, \emph{rjson} \citep{Couture-Beil2014-kk} and \emph{tximport} \citep{Soneson2015-fw} for reading in transcript-level expression values;
+\emph{viridis} \citep{Garnier2016-hk} for perceptually-uniform colour
+maps for plotting; \emph{parallel} for parallel computation; \emph{matrixStats} \citep{Bengtsson2016-tn} for computation of summary statistics from matrices; \emph{cowplot} \citep{Wilke2016-hj} for attractive plotting themes; \emph{destiny} \citep{Angerer2015-sw} for producing diffusion maps; \emph{Rtsne} \citep{Krijthe2015-is} for producing t-SNE plots; \emph{mvoutlier}
+\citep{Filzmoser2015-kx} for multivariate outlier detection from PCA of QC metrics; \emph{roxygen2} \citep{Wickham2015-pu}, \emph{BiocStyle} \citep{Huber2015-en}, \emph{knitr} \citep{Xie2013-bn} and \emph{rmarkdown} \citep{Allaire2016-dl} for generating documentation; and \emph{testthat} \citep{Wickham2011-cj} for unit testing.
+As well as functioning in the usual R environments, \emph{scater} also has a GUI built using
+\emph{shiny} \citep{Chang2016-of} and \emph{shinydashboard} \citep{Chang2015-bn} for intuitive and interactive data visualisation.
+Calling the \verb|scater_gui| function from within an R session opens up the GUI in a web browser.
 
@@
 Building on Bioconductor's ExpressionSet class, it is a fully-featured, sophisticated and flexible data class tailored to scRNA-seq data.}\label{fig:02}
+\end{figure}
+
+\begin{figure}[!tpb]
+\centerline{\includegraphics[width=0.7\textwidth]{figures/find-pcs_c1_machine.pdf}}
+\caption{A QC plot produced by the plotQC function in \emph{scater} showing violin, scatter- and boxplots of principal component values against the C1 machine used for each cell for the six principal components most strongly correlated with C1 machine used.}\label{fig:plotqc-c1-machine}
 \end{figure}
@@
 
-\section*{Details of package dependencies}
-
-The package builds on many other R packages: \emph{Biobase} and \emph{BiocGenerics} for core Bioconductor functionality  \citep{Huber2015-en}; \emph{plyr} \citep{Wickham2015-kj}, \emph{reshape2} \citep{Wickham2012-ec}, \emph{dplyr} \citep{Wickham2015-la}, \emph{data.table} \citep{Dowle2015-zc} and \emph{magrittr} \citep{Bache2014-sa} for reading and
-tidying data; \emph{ggplot2} \citep{Wickham2016-dc} for plotting; \emph{biomaRt} \citep{Durinck2005-yz} for feature annotation; \emph{edgeR} \citep{Robinson2010-ky} for computation of normalisation size factors and counts-per-million values; \emph{limma} \citep{Ritchie2015-so} for efficient fitting of linear models to features; \emph{rhdf5} \citep{Fischer2016-me}, \emph{rjson} \citep{Couture-Beil2014-kk} and \emph{tximport} \citep{Soneson2015-fw} for reading in transcript-level expression values;
-\emph{viridis} \citep{Garnier2016-hk} for perceptually-uniform colour
-maps for plotting; \emph{parallel} for parallel computation; \emph{matrixStats} \citep{Bengtsson2016-tn} for computation of summary statistics from matrices; \emph{cowplot} \citep{Wilke2016-hj} for attractive plotting themes; \emph{destiny} \citep{Angerer2015-sw} for producing diffusion maps; \emph{Rtsne} \citep{Krijthe2015-is} for producing t-SNE plots; \emph{mvoutlier}
-\citep{Filzmoser2015-kx} for multivariate outlier detection from PCA of QC metrics; \emph{roxygen2} \citep{Wickham2015-pu}, \emph{BiocStyle} \citep{Huber2015-en}, \emph{knitr} \citep{Xie2013-bn} and \emph{rmarkdown} \citep{Allaire2016-dl} for generating documentation; and \emph{testthat} \citep{Wickham2011-cj} for unit testing.
-As well as functioning in the usual R environments, \emph{scater} also has a GUI built using
-\emph{shiny} \citep{Chang2016-of} and \emph{shinydashboard} \citep{Chang2015-bn} for intuitive and interactive data visualisation.
-Calling the \verb|scater_gui| function from within an R session opens up the GUI in a web browser.
-
 \bibliographystyle{natbib}

### 2016-11-06 response_to_reviewers.md
Revisions to manuscript for resubmission

# Aaron's comments

Sweet! Here’s my comments, though keep in mind that this is after about 30 hours w/o sleep, so several grains of salt required:

> Specific comments:
> Major:
> - How do the wrapper functions for running kallisto and Salmon in scater deal with UMI data?  e.g. I don’t believe the current implementation of Salmon has built-in functionality for handle scRNA-seq data with UMIs?

Um. They don’t, I guess?

> - Could the authors add an assessment of how long it takes to perform basic tasks depending on number of genes/cells? With new droplet technologies to sort cells, will R be the best tool to analyze potentially 100s of 1000s of cells?

Point out C++ extension to allow memory- and time-efficient calculations of QC metrics, manipulations, etc. Scales to Zeisel data set (3000 cells, 10000 genes post-filtering) on a desktop machine (8 GB RAM). Larger data sets inevitably require HPC resources; perhaps can get more savings with sparse matrix representations, but this is not easily compatible with downstream methods (e.g., t-SNE, PCA) requiring the full matrix. So just suck it up and fork out some cash.

> - Are there functions to extract the highly variable (or expressed) genes? There are options to visualize most highly expressed features and dim reduction plots that use highly variable genes, but how to extract?  Also, how would you decide what set of genes to look at (e.g. DE genes, highly expressed or highly variable genes) depending on if you were interested in rare cell types versus identifying common major cell types?  Is there an option to control this in the data visualization tools?

Extended in scran - deviation-from-median (DM) method in Kolodziejczyk et al., CV2-based method (technicalCV2) by Brennecke et al., and Lun et al.’s variance of log-expression based methods. HVG selection can be a complicated procedure, so plotPCA/plotTSNE just uses sample variance as a quick-and-dirty default. However, selection based on these more complex methods can be performed by specifying the detected HVGs in feature_set=, allowing consideration of technical contributions to the variance, fitting of the mean-variance trend, rigorous hypothesis testing for HVGs, etc. See example from workflow for how this gets accommodated.

Also planning on testing and implementing the Gini index (Giniclust, can’t remember the paper) as a HVG selection option in scran; focuses on genes that are upregulated in rare cell types, to highlight these cells upon visualisation.

> - For normalization, does it make sense to use bioC packages such as svaseq where if you are trying to discover biological conditions (e.g. novel cell types) from scRNA-seq data, you could be removing the interesting biological variation in the latent factors? What are the author’s opinions of using spike-ins or control genes? Could the authors provide a discussion on this topic or at least refer the reader to other papers to help guide them?

Yes, you could potentially be removing interesting biological variation in the latent variables, in cases where discovery of new and interesting things is the main aim. Perhaps a note about this would be prudent.

Also, see F1000 workflow for discussion of benefits of spike-in normalisation vs non-DE normalisation. Blah blah blah.

> - How did the authors decide on the selection of variable genes? Is there a reason the authors applied the specific normalization of FPKMs to identify the variable genes instead of something like the distance-to-median approach in Kolodziejczyk et al. (2015) Cell Stem Cell. I would like to see a larger discussion of this and I would like the authors to consider other approaches of identifying highly variable genes and how this affects their downstream clustering results?

We used variance of log-values, I believe, in selecting the HVGs within plotPCA, etc.? This has several advantages over CV2-based methods:

1) plotPCA/plotTSNE use log-expression values, so HVGs based on high variance of logged values -> high Euclidean distances, maximise log-fold change and discrimination between cells.
2) log-transformation protect against genes with outlier expression patterns, i.e., very strong expression in only one cell.
3) logged counts are more normal-ish than the raw counts themselves, making it easier to construct tests for significance of HVG’ness.

In general, the quantitative differences between HVG selection methods are modest. Strong HVGs distinguishing cell types get picked up no matter what you do, so you should be able to see separate clusters. It is likely that differences occur in visualisation of more subtle features, e.g., rare cell types (CV2 is better, as it selects outlier-like genes) or weak clusters (where log might be better if it selects more discriminatory genes).

Note, the fact that we’re using FPKMs is irrelevant here, as the length scaling factor cancels out when computing the sample variance within a gene. No mean-variance trend is fitted so the absolute values don’t matter.

> - What if researchers have two types of data (e.g. scRNA-Seq and DNA methylation in scMT-seq https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-0950-z). How does the SCEset container handle this?  Are there different assays that would allow multiple types of information to be held in the same object?

Not well, as it's outside the design remit of the single cell expression set. But Vince Carey’s MultiAssay object (or something like that, can’t remember exact name) can be used.

> - Stegle et al. 2015 noted there is a big 3’ bias in scRNA-Seq protocols (particularly for the SMART-Seq protocol used by Fluidigm). Therefore, normalizing for transcript length (e.g. FPKM) can lead to an underestimate of expression in long transcripts and over expression in short transcripts. I would like to see the authors create a data visualization function in scater that would allow a user to check this in their own data.

Sure, I guess.

> Minor:
> - Pg 3, line 26: could the authors include new salmon reference listed in bioRxiv (http://biorxiv.org/content/early/2016/08/30/021592)

Yeah, fine, whatever. At least we know who this reviewer is likely to be, then. Who else would care?

> Reviewer: 2
>
> Comments to the Author
> The work by McCarthy and colleagues describes an R package to pre-process and QC single-cell RNA-seq data. The authors did a very nice job with documentation at all levels: manuscript, supplements and especially with the package vignette/case study, which was also provided as supplementary information.
>
> The package itself does not introduce any novel method to QC or normalize single-cell RNA-seq. Rather, it aims at becoming a one-stop solution for loading and organizing single-cell data, which can then be fed to existing methods. While the absence of novel analytical strategies diminishes the scope of the work and may somewhat mitigate its impact, I note that the design choices behind this package, particularly the data structures, are very elegant and polished, and fit well with the R/Bioconductor universe. Visualizations are useful, appealing and pleasant. A regular R user will immediately recognize the quality of this package, which therefore has the potential to become "the" choice for single-cell RNA-seq QC, pretty much like edgeR/DESeq are for differential gene expression.
>
> My points are minor and optional:
>
> - the authors may consider allocating more text to outlier identification. They provide a reference, but it may be worth summarizing briefly how outliers are identified. The metrics used for outlier detection are listed in the package vignette, but I feel this is important information that should go in the main text.

Okay…. are these the QC outliers detected with the PCA-based method in mvoutlier? Guess we could talk about it a bit more?

> - some packages already provide routines for batch correction. This is the case, for instance, of scde. Indeed, scde requires very little upstream pre-processing, mostly just filtering out outliers. So there is no one-size-fits-all workflow. In this regard, I wonder whether the authors might further elaborate on Figure 1 and Supplementary Figure 3 by providing schematics of the most common workflows for the most popular tools, highlighting which portions would be scater-specific and which would be third-party tool-specific?

I’m pretty sure scde requires a bit more work… at the very least, QC on genes and cells isn’t so simple as to be fobbed off in a single sentence. Probably mention the inputs to each of these tools; methods require a count matrix (scde, edgeR) or expression matrix (monocle) or log-expression matrix (plotTSNE, PCA, etc.). Last lot will require batch correction as batches cannot be modelled in the methods themselves.

scde appears to provide one function "clean.counts" for "QC" before running its functions, but this function will only filter cells by minimum library size and genes by minimum total reads and minimum number of cells a gene should be seen in. Thus, preparing a dataset with scater upstream of using scde offers many more possibilities for QC of both cells and genes. scde indeed offers the option to include a known batch effect in its DE models - scater again offers many options for exploring batch effects that should prove useful upstream of modelling with scde.

*Can very easily show some possible workflows with an extra Supp Fig for, for example DE with scde, DE with edgeR, clustering with SC3, ...*

> - the authors did a good job highlighting the possible influence of cell cycle on measurements. I wonder whether the effect of cell cycle should be also mentioned in the latent variable section of the case study. It is not infrequent that a treatment significantly alters the cell cycle, partially masking the phenotype of interest. This can be explicitly investigated by using cell cycle genes to drive PCA, for example, as the authors pointed out. However, there may be cases where the user did not suspect an involvement of cell cycle, but latent variable analysis may reveal it, prompting a more directed investigation.

Fine. Can combined with cyclone (implemented in scran) to determine if levels of the latent variable are associated with cell cycle classification.

> - I wonder whether scater might provide some simple summary stats regarding the distribution of genes. Roughly, genes may show 4 types of distributions: 1) noise 2) zero inflated 3) bimodal 4) log normal. It would be nice if scater provided related information, perhaps upon GMM fitting, such as the number of genes in each class for every sample, and then some basics such as the median expression and standard deviation of log normally distributed genes etc.

Sounds like a pain, not sure what it would tell you. Doesn’t tell you anything about QC - is having lots of bimodal genes good or bad? Who knows? We already get zero-inflation diagnostics from drop-out rate plots. Similarly, it doesn’t tell you much about the interesting genes - log-normally distributed genes can be high or low variance, in which case it is interesting and not interesting respectively. Finally, it’s a pain to draw a clear boundary between noise and zero-inflation (as large dispersions will automatically have a point mass at zero), bimodal and zero inflation (probably the same, if the mean of the non-zero component is large enough), and bimodal and log-normal (test for bimodality - can’t remember the name - seems to involve a few iffy assumptions).

> - minor details: how did the authors set the perplexity parameter of tSNE? How many of the top variable genes are used in PCA and dimensionality reduction in general? Are violin plots scaled to take into account that different groups might have different number of cells?

Default perplexity (20?), 500 genes, dunno, respectively.

-------------------

### 2016-11-06 biorxiv_template.tex
Revisions to manuscript for resubmission

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in,marginparwidth=2in]{geometry}

% use Unicode characters - try changing the option if you run into troubles with special characters (e.g. umlauts)
\usepackage[utf8]{inputenc}

% clean citations
\usepackage{cite}

% hyperref makes references clicky. use \url{www.example.com} or \href{www.example.com}{description} to add a clicky url
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% improves typesetting in LaTeX
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% text layout - change as needed
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% use adjustwidth environment to exceed text width (see examples in text)
\usepackage{changepage}

% adjust caption style
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}

% remove brackets from references
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% headrule, footrule and page numbers
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}

% use \textcolor{color}{text} for colored text (e.g. highlight to-do areas)
\usepackage{color}

% define custom colors (this one is for figure captions)
\definecolor{Gray}{gray}{.25}

% this is required to include graphics
\usepackage{graphicx}

% use if you want to put caption to the side of the figure - see example in text
\usepackage{sidecap}

% use for have text wrap around figures
\usepackage{wrapfig}
\usepackage[pscoord]{eso-pic}
\usepackage[fulladjust]{marginnote}
\reversemarginpar

% document begins here
\begin{document}
\vspace*{0.35in}

% title goes here:
\begin{flushleft}
{\Large
\textbf\newline{\LaTeX Template for Manuscripts with Embedded Figures.}
}
\newline
% authors go here:
\\
Author 1\textsuperscript{1},
Author 2\textsuperscript{2},
Author 3\textsuperscript{1},
Author 4\textsuperscript{1},
Author 5\textsuperscript{2},
Author 6\textsuperscript{2},
Author 7\textsuperscript{1,*}
\\
\bigskip
\bf{1} Affiliation A
\\
\bf{2} Affiliation B
\\
\bigskip
* correseponding@author.mail

\end{flushleft}

\section*{Abstract}
Fitting multiple figures into very tight manuscripts while keeping it pleasant to read is challenging. Therefore figures are often simply attached to the very end of a manuscript file. While easier for the authors, this practice is inconvenient for readers. This \LaTeX template shows how to generate a compiled PDF with figures embedded into the text. It provides several examples of how to embed figures or tables directly into the text thus giving you a range of options from which you should choose the one best suited for your manuscript. Check out Schlegel et al., (2016) as example of use \cite{Schlegel2016}.

% now start line numbers
\linenumbers

% the * after section prevents numbering
\section*{Introduction}
In the introduction you will see an example of text wrapping around the figure with a figure caption on the margin (Fig. \ref{fig1}). This is done by combining the \verb!wrapfigure! with % avoid blank space here
\marginpar{
\vspace{.7cm} % adjust vertical position relative to text with \vspace{} - note that you can enter negative numbers to move margin caption up
\color{Gray} % this gives caption a grey color to set it apart from text body
\textbf{Figure \ref{fig1}. Example of a margin caption.} % note that \ref{fig1} refers to the corresponding wrapfigure
Setting up your figure + caption like this looks fancy and does not disrupt the flow of the text. But it requires more manual adjustments (position, spacing, labeling) compared to using standard \LaTeX figure environments.
}
\begin{wrapfigure}[19]{l}{75mm}
% the number in [] of wrapfigure is optional and gives the number of text lines that should be wrapped around the text. Adjust according to your figures height
\includegraphics[width=75mm]{fig1.pdf}
\captionsetup{labelformat=empty} % makes sure dummy caption is blank
\caption{} % add dummy caption - otherwise \label won't work and figure numbering will not count up
\label{fig1} % use \ref{fig1} to reference to this figure
\end{wrapfigure} % avoid blank space here
the \verb!marginnote! environment. Please note that in this case the figure (\verb!wrapfigure!) and the figure caption (\verb!marginpar!) have to be separated as you can tell from the code. The \verb!wrapfigure! environment can be a bit tricky when it comes to text formatting. Thus some general hints: (a) try to avoid line breaks in the code as this may result in weird formatting around the figure, (b) the figure should not span multiple headlines (sections) and (c) if you encounter problems with the line break right after the \verb!wrapfigure! try using \verb!\mbox{}! to prevent premature line breaks (\mbox{see example in code}). As stated in the figure caption, setting up figures this way requires a bit more manual adjustments but it makes figures blend in nicely without interrupting flow of text.

\section*{Materials and Methods}

\subsection*{Tables.}

In this section you will find an example of a table using the \verb!table! plus the \verb!adjustwidth! environment and should give you a minimal example for tables in \LaTeX (Table \ref{tab1}).

\begin{table}[!ht]
\begin{adjustwidth}{-1.5in}{0in} % comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{{\bf Example Table.} This table demonstrates how to use the  adjustwidth environment if you need that tad bit of extra width for your figures or tables.}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multicolumn{4}{|l|}{\bf Heading 1} & \multicolumn{3}{|l|}{\bf Heading 2}\\ \hline
cell 1 - row 1 & cell 2 - row 1 & cell 3 - row 1 & cell 4 - row 1 & cell 5 - row 1 & cell 6 - row 1 & cell 7 - row 1 \\ \hline
cell 1 - row 2 & cell 2 - row 2 & cell 3 - row 2 & cell 4 - row 2 & cell 5 - row 2 & cell 6 - row 2 & cell 7 - row 2 \\ \hline
cell 1 - row 3 & cell 2 - row 3 & cell 3 - row 3 & cell 4 - row 3 & cell 5 - row 3 & cell 6 - row 3 & cell 7 - row 3 \\ \hline
\end{tabular}
\label{tab1}
\end{adjustwidth}
\end{table}

\paragraph{Paragraph.}
Instead of adding more and more subsections you can use the \verb!\paragraph{}! command to give structure to your manuscript.

\subsection*{Formulas.}
For mathematical formulas you should use the math environment. See this example:

\begin{center}
$f(A_{ik},A_{jk}) = min(A_{ik},A_{jk}) - C_{1} max(A_{ik},A_{jk}) e^{-C_{2}min(A_{ik},A_{jk})}$
\end{center}

% newpage forces a page break if you want to clearly separate materials from results
\newpage

\section*{Results}
\subsection*{Standard floating figures.}
Figure \ref{fig2} is wrapped into a standard floating environment. That means that \LaTeX will determine the exact placement of the figure. Even though you can state preferences (see code) it can be tricky to get the right placement - especially when working on very tight manuscripts. If you want exact placement, add \verb!\usepackage{float}! to this file's header and use [H] in the figure environment's placement options.

\begin{figure}[ht] %s state preferences regarding figure placement here

% use to correct figure counter if necessary
%\renewcommand{\thefigure}{2}

\includegraphics[width=\textwidth]{fig2.pdf}

\caption{\color{Gray} \textbf{Example of a standard floating figure}. \textbf{A-F}, This figure is wrapped into the standard floating environment.}

\label{fig2} % \label works only AFTER \caption within figure environment

\end{figure}

\subsection*{Page break in figures.}
The standard floating figures in \LaTeX do not cope well with page breaks which can make it difficult to fit in large figures. One way to deal with this is to separate figure and caption but \verb!\caption{}! might still give troubles at page breaks. Figure 3 demonstrates a way to manually set up figure and caption such that it continues onto the next page.
\vspace{.5cm} % set vertical space between text and figure
\begin{adjustwidth}{-2in}{0in}
\begin{flushright}
\includegraphics[width=163mm]{fig3.pdf}
\end{flushright}
\justify
\color{Gray}
\textbf {Figure 3. Example of a wide figure with multi-page caption.}
\textbf{A}, Proin lectus ex, venenatis vel ornare eget, hendrerit tempus justo. Pellentesque molestie purus sed pretium tincidunt. Curabitur facilisis, orci vitae mollis fringilla, elit erat fermentum justo, nec luctus nunc sapien vel dolor. Cras enim justo, ullamcorper ut commodo at, posuere et ex. Fusce cursus sapien id augue maximus convallis. Praesent egestas massa in enim volutpat varius. In aliquam turpis urna, at elementum turpis eleifend at. \textbf{B}, Proin risus erat, tincidunt quis massa non, sollicitudin congue metus. Aliquam quis magna vulputate, posuere est eu, tempor nisi. Cras gravida tempus felis, vitae lacinia lacus volutpat quis. Pellentesque et eros eu mi suscipit tempus. Proin in augue scelerisque. \textbf{C}, Donec a tempor tortor, et dignissim enim. Cras in ipsum sed velit bibendum imperdiet. Aenean aliquet mauris maximus, sodales ligula sit amet, placerat felis. In tristique nisi eu risus rutrum, ac lacinia lorem cursus. Nunc eget condimentum purus. Maecenas imperdiet nisl eu accumsan gravida. \textbf{D}, Nullam tincidunt, magna sed auctor ultrices, leo mi eleifend velit, quis varius ex diam non tellus. Nam tincidunt vehicula turpis, ut euismod turpis elementum vel.
\end{adjustwidth}

%\clearpage makes sure that all above content is printed at this point and does not invade into the upcoming content
%\clearpage

\section*{Discussion}
\subsection*{Subsection heading.}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Aliquam bibendum finibus diam, gravida sagittis lorem gravida vitae. Interdum et malesuada fames ac ante ipsum primis in faucibus. Nulla in diam tristique ante posuere tristique. Donec interdum purus sit amet nisl accumsan consectetur. Fusce aliquet libero mi, quis ornare dolor congue ullamcorper. Nulla nulla urna, molestie in urna sed, lacinia volutpat eros. Ut mi libero, elementum scelerisque ipsum vel, hendrerit fermentum turpis. Aliquam sit amet leo sodales, egestas augue id, fermentum nulla. Aenean vel cursus ante, et pellentesque eros. Nulla ac neque nec justo posuere commodo sit amet sit amet justo. Aliquam tincidunt tempor ex nec tincidunt. In ullamcorper vehicula lobortis.

%\clearpage

\section*{Supporting Information}
If you intend to keep supporting files separately you can do so and just provide figure captions here. Optionally make clicky links to the online file using \verb!\href{url}{description}!.

%These commands reset the figure counter and add "S" to the figure caption (e.g. "Figure S1"). This is in case you want to add actual figures and not just captions.
\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}

% You can use the \nameref{label} command to cite supporting items in the text.
\subsection*{S1 Figure}
\label{example_label}
{\bf Caption of Figure S1.} \textbf{A}, If you want to reference supporting figures in the text, use the \verb!\nameref{}!. command. This will reference the section's heading: \nameref{example_label}.

\subsection*{S2 Video}
\label{example_video}
{\bf Example Video.} Use \href{www.youtube.com}{clicky links} to the online sources of the files.

%\clearpage

\section*{Acknowledgments}
We thank just about everybody.

\nolinenumbers

%This is where your bibliography is generated. Make sure that your .bib file is actually called library.bib
\bibliography{library}

%This defines the bibliographies style. Search online for a list of available styles.
\bibliographystyle{abbrv}

\end{document}

### 2016-11-06 main.tex
Update to main text
Revisions to manuscript for resubmission

@@
 \documentclass{bioinfo}
+\newcommand{\revised}[1]{\textcolor{red}{#1}}
+
 \copyrightyear{2016} \pubyear{2016}
@@
 
-\title[scater package]{\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R}
-\author[McCarthy \textit{et~al}.]{Davis J.~McCarthy\,$^{\text{\sfb 1,2,5}*}$, Kieran R.~Campbell\,$^{\text{\sfb 2,4}}$, Aaron T.~L.~Lun\,$^{\text{\sfb 6}}$ and Quin F.~Wills\,$^{\text{\sfb 2,3}}$}
+\title[scater scRNA-seq package]{\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R}
+\author[McCarthy \textit{et~al}.~2016]{Davis J.~McCarthy\,$^{\text{\sfb 1,2,5}*}$, Kieran R.~Campbell\,$^{\text{\sfb 2,4}}$, Aaron T.~L.~Lun\,$^{\text{\sfb 6}}$ and Quin F.~Wills\,$^{\text{\sfb 2,3}}$}
 \address{$^{\text{\sf 1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, United Kingdom;\\
@@
 
-Numerous statistical methods and software tools have been published for scRNA-seq data \citep{Guo2015-wx,Kharchenko2014-rx,Finak2015-rd,Delmans2016-qc,Angerer2015-sw,Kiselev2016-fu,Julia2015-jt,Trapnell2014-gj}.
+Numerous statistical methods and software tools have been published for scRNA-seq data \citep{Guo2015-wx,Kharchenko2014-rx,Finak2015-rd,Angerer2015-sw,Julia2015-jt,Trapnell2014-gj}.
 However, all of these assume that quality control and normalisation have already been applied.
@@
 In particular, current visualisation methods designed for exploratory data analysis of bulk transcriptomic experiments are unsuited to datasets containing hundreds or thousands of cells.
-The large size of each dataset also favours methods such as \emph{kallisto} \citep{Bray2016-pj} and \emph{Salmon} \citep{Patro2015-jf} for rapidly quantifying gene expression.
+The large size of each dataset also favours methods such as \emph{kallisto} \citep{Bray2016-pj} and \emph{Salmon} \revised{\citep{Patro2016-rs}} for rapidly quantifying gene expression.
 Extensions to the current computational infrastructure are required to provide appropriate data structures and methods that can accommodate these rich scRNA-seq datasets for integrative analyses of expression and other assay data along with the accompanying metadata.
@@
 Further case studies using \emph{scater} on published data, for example from 3000 mouse cortex cells \citep{Zeisel2015-ab} and 1200 cells from early-development mouse embryos \citep{Scialdone2016-oa} are available at \href{http://dx.doi.org/10.5281/zenodo.59897}{dx.doi.org/10.5281/zenodo.59897}.
+All materials required to reproduce the results presented in this paper are available at \href{http://dx.doi.org/10.5281/zenodo.60139}{dx.doi.org/10.5281/zenodo.60139}.
 
@@
 \begin{figure}[!tpb]%figure1
-\centerline{\includegraphics[width=0.48\textwidth]{figures/scater_qc_workflow.pdf}}
+\centerline{\includegraphics[width=0.48\textwidth]{scater_qc_workflow.eps}}
 \caption{An overview of the \emph{scater} workflow, from raw sequenced reads to a tidy data set ready for higher-level downstream analysis.
@@
 The \emph{scater} package is an open-source R package available through     Bioconductor.
-Key aspects of the code are written in C++ to minimise computational time and memory use.
-The package builds on many other R packages, including \emph{Biobase} and \emph{BiocGenerics} for core Bioconductor functionality \citep{Huber2015-en}; \emph{destiny} \citep{Angerer2015-sw} and \emph{Rtsne} \citep{Krijthe2015-is} for dimensionality reduction; and \emph{edgeR} \citep{Robinson2010-ky} and \emph{limma} \citep{Ritchie2015-so} for model fitting and statistical analyses.
+Key aspects of the code are written in C++ to minimise computational time and memory use\revised{, and the package scales well to large datasets.
+For example, consider the \citet{Macosko2015-vt} dataset, which contains more than 44,000 cells.
+The core scater functions to create an SCESet object and calculate QC metrics took approximately two minutes to complete on an early 2015 MacBook Pro laptop with 2.9 GHz Intel Core i5 processor and 16 Gb of RAM.
+Subsetting the SCESet object takes only a few seconds, and producing a PCA plot with the plotPCA function takes less than a minute.}
+
+The package builds on many other R packages, including \emph{Biobase} and \emph{BiocGenerics} for core Bioconductor functionality \citep{Huber2015-en}; \emph{destiny} \citep{Angerer2015-sw} and \emph{Rtsne} for dimensionality reduction; and \emph{edgeR} \citep{Robinson2010-ky} and \emph{limma} \citep{Ritchie2015-so} for model fitting and statistical analyses.
 The plotting functionality in the package uses \emph{ggplot2} \citep{Wickham2016-dc}.
@@
 \begin{figure*}[!htpb]%figure3
-\centerline{\includegraphics[width=\textwidth]{figures/figure3_alt.pdf}}
+\centerline{\includegraphics[width=\textwidth]{figure3_alt.eps}}
 \caption{\textbf{Different types of QC plots that can be generated with \emph{scater}.} (\textbf{a}) Cumulative expression plot showing the proportion of the library accounted for by the top 1--500 most highly-expressed features.
@@
 
-The plotPCA function implements an approach to automatic outlier
+\revised{The plotPCA function implements an approach to automatic outlier
 detection using multivariate normal methods applied to the cell-level QC
-metrics \citep{Ilicic2016-dm}. Specifically, PCA is applied to the QC metrics
-for all cells and a plot is produced to automatically detect outliers in the
-higher-dimensional QC metric space (Figure~\ref{fig:03}b). These outliers
+metrics \citep{Ilicic2016-dm} and a PCA plot computed from QC metrics is produced showing detected outliers (Figure~\ref{fig:03}b).
+Briefly, semi-robust principal components are computed on robustly sphered QC metrics data to determine distances for each cell and weights for outlier detection, using the pcout function from the \emph{mvoutlier} R package, which implements the fast algorithm of \citet{Filzmoser2008-sy}.
+By default, the following QC metrics are used for automatic outlier detection: the percentage of counts from the top 100 features, the total number of features with detectable expression, the percentage of counts from control features, the number of detected feature controls, the log-scaled counts from endogenous features and log-transformed counts from feature controls are used.
+The user can specify QC metrics or other cell metadata variables to use for outlier detection with the "selected\_variables" argument to plotPCA.
+Detected} outliers
 correspond to low-quality cells with abnormal library characteristics (e.g.,
@@
 \begin{figure*}[!htpb]%figure4
-\centerline{\includegraphics[width=\textwidth]{figures/figure4.pdf}}
+\centerline{\includegraphics[width=\textwidth]{figure4.eps}}
 \caption{\textbf{Reduced dimension representations of cells and gene expression plots with \emph{scater.}} Plots are shown using all genes (\textbf{a-c}) and cell cycle genes only (\textbf{d-f}) using PCA (\textbf{a,d}), t-SNE (\textbf{b,e}) and diffusion maps (\textbf{c,f}), where each point represents a cell.
@@
 In addition, \emph{scater} is tightly integrated with the \emph{scran} package that implements a method utilising cell pooling and deconvolution to compute size factors better suited to scRNA-seq data \citep{Lun2016-sk}.
+\revised{\citet{Lun2016-uj} also offers further discussion of the respective benefits and drawbacks of spike-in normalisation and non-DE normalisation.}
 
@@
 large data sets to identify (potentially unknown) sources of unwanted variation
-\citep{Leek2010-nq,Hicks2015-qy,Tung2016-jy,Bacher2016-ay,Grun2015-xi}.
+\citep{Leek2010-nq,Hicks2015-qy,Grun2015-xi}.
 \emph{scater} is compatible with existing methods such as \emph{svaseq}
@@
 rather than C1 machine (Figure~\ref{fig:05}c).
+\revised{More targeted applications of these methods can be used to remove specific effects, for example, by identifying latent factors from cell cycle genes to remove the cell cycle effect.}
 
@@
 Where this is not possible (e.g., for visualisation), directly regressing out these uninteresting factors is required to obtain "corrected" expression values for further analysis.
+\revised{A general risk of removing latent factors is that interesting biological variation may be removed along with the presumed unwanted variation.
+Users should therefore apply such methods with appropriate caution, particularly when an analysis aims to discover biological conditions, such as new cell types.}
 
 \begin{figure*}[!tpb]%figure5
-\centerline{\includegraphics[width=\textwidth]{figures/figure5.pdf}}
-\caption{\textbf{Figure 5: Normalisation and batch correction with
+\centerline{\includegraphics[width=\textwidth]{figure5.eps}}
+\caption{\textbf{Normalisation and batch correction with
 \emph{scater}.} Principal component analysis plots showing cell
 structure in the first two PCA dimensions using various normalisation
-methods methods that can be easily applied in \emph{scater}, including
+methods that can be easily applied in \emph{scater}, including
 endogenous size-factor normalisation using methods from the \emph{scran}
@@
 cell cycle phase identification \citep{Scialdone2015-gj} and pseudotime
-computation \citep{Trapnell2014-gj,Angerer2015-sw,Julia2015-jt,Campbell2016-es,Campbell2015-nj,Haghverdi2016-is}. The \emph{scater} package bridges
+computation \citep{Trapnell2014-gj,Angerer2015-sw,Julia2015-jt,Campbell2016-es,Haghverdi2016-is}. The \emph{scater} package bridges
 the gap between raw reads and these downstream analysis tools by providing the
@@
 structure combining multiple data modalities and metadata necessary for
-convenient, robust and reproducible analyses of scRNA-seq data.
+convenient, robust and reproducible analyses of scRNA-seq data \revised{(see Supplementary Material for discussion of entry points to several third party tools from \emph{scater}).}
 
@@
 likely to compromise the validity of all downstream analyses
-\citep{Leek2010-nq,Hicks2015-qy,Tung2016-jy,Bacher2016-ay,Grun2015-xi}.
+\citep{Leek2010-nq,Hicks2015-qy,Grun2015-xi}.
 
@@
 Once identified, important covariates and latent variables can be flagged for inclusion in downstream statistical models or their effects regressed out of normalised expression values.
+\revised{The package is well documented and a recent step-by-step workflow article demonstrates detailed use of \emph{scater} in combination with other analysis packages in a range of scenarios \citep{Lun2016-uj}.}
 
@@
 
-This work was supported by the National Health and Medical Research Council of Australia [Early Career Fellowship to D.J.M.], by core funding from the European Molecular Biology Laboratory [D.J.M.], core funding from Cancer Research UK [A17197 to A.T.L.L.), the United Kingdom Medical Research Council [studentship to K.R.C.) and the Oxford Single Cell Biology Consortium [Q.F.W.].\vspace*{-12pt}
+This work was supported by the National Health and Medical Research Council of Australia [APP1112681 to D.J.M.], by core funding from the European Molecular Biology Laboratory [D.J.M.], core funding from Cancer Research UK [A17197 to A.T.L.L.), the United Kingdom Medical Research Council [studentship to K.R.C.) and the Oxford Single Cell Biology Consortium [Q.F.W.].\vspace*{-12pt}
 

### 2016-11-06 main_biorxiv.tex
Revisions to manuscript for resubmission

--- 2016-08-14 main.tex
@@
-\documentclass{bioinfo}
-\copyrightyear{2016} \pubyear{2016}
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in,marginparwidth=2in]{geometry}
 
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Manuscript Category}
+% use Unicode characters - try changing the option if you run into troubles with special characters (e.g. umlauts)
+\usepackage[utf8]{inputenc}
+\usepackage[english]{babel}
 
+% clean citations
+%\usepackage{cite}
+\usepackage[round]{natbib}
+
+% hyperref makes references clicky. use \url{www.example.com} or \href{www.example.com}{description} to add a clicky url
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% improves typesetting in LaTeX
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% text layout - change as needed
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in
+\textheight 8.75in
+
+% use adjustwidth environment to exceed text width (see examples in text)
+\usepackage{changepage}
+
+% adjust caption style
+%\usepackage[labelfont=bf,labelsep=period,singlelinecheck=off]{caption}
+
+% remove brackets from references
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% headrule, footrule and page numbers
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+
+% use \textcolor{color}{text} for colored text (e.g. highlight to-do areas)
+\usepackage{color}
+
+% define custom colors (this one is for figure captions)
+\definecolor{Gray}{gray}{.25}
+
+% this is required to include graphics
+\usepackage{graphicx}
+
+% use if you want to put caption to the side of the figure - see example in text
+\usepackage{sidecap}
+
+% use for have text wrap around figures
+\usepackage{wrapfig}
+\usepackage[pscoord]{eso-pic}
+\usepackage[fulladjust]{marginnote}
+\reversemarginpar
+
+% document begins here
 \begin{document}
-\firstpage{1}
+\vspace*{0.35in}
 
-\subtitle{Subject Section}
+% title goes here:
+\begin{flushleft}
+{\Large
+\textbf\newline{\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R}
+}
+\newline
+% authors go here:
+\\
+Davis J.~McCarthy\,\textsuperscript{1,2,5,*},
+Kieran R.~Campbell\,\textsuperscript{2,4},
+Aaron T.~L.~Lun\,\textsuperscript{6},
+Quin F.~Wills\,\textsuperscript{2,3}
+\\
+\bigskip
+\bf{1} European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, United Kingdom;
+\\
+\bf{2} Wellcome Trust Centre for Human Genetics, University of Oxford,
+Roosevelt Drive, Oxford OX3 7BN, United Kingdom;
+\\
+\bf{3} Weatherall Institute for Molecular Medicine, University of Oxford, John Radcliffe Hospital, Oxford OX3 9DS, United Kingdom;
+\\
+\bf{4} Department of Physiology, Anatomy and Genetics, University of Oxford, South Parks Road, Oxford OX1 3QX, United Kingdom;
+\\
+\bf{5} St Vincent's Institute of Medical Research, 41 Victoria Parade, Fitzroy Victoria 3065, Australia; and
+\\
+\bf{6} CRUK Cambridge Institute, University of Cambridge, Robinson Way, Cambridge CB2 0RE, United Kingdom.
+\\
+\bigskip
+* \textbf{Contact:} \href{davis@ebi.ac.uk}{davis@ebi.ac.uk}
 
-\title[scater package]{\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R}
-\author[McCarthy \textit{et~al}.]{Davis J.~McCarthy\,$^{\text{\sfb 1,2,5}*}$, Kieran R.~Campbell\,$^{\text{\sfb 2,4}}$, Aaron T.~L.~Lun\,$^{\text{\sfb 6}}$ and Quin F.~Wills\,$^{\text{\sfb 2,3}}$}
-\address{$^{\text{\sf 1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, United Kingdom;\\
-$^{\text{\sf 2}}$Wellcome Trust Centre for Human Genetics, University of Oxford,
-Roosevelt Drive, Oxford OX3 7BN, United Kingdom;\\
-$^{\text{\sf 3}}$Weatherall Institute for Molecular Medicine, University of Oxford, John Radcliffe Hospital, Oxford OX3 9DS, United Kingdom;\\
-$^{\text{\sf 4}}$Department of Physiology, Anatomy and Genetics, University of Oxford, South Parks Road, Oxford OX1 3QX, United Kingdom;\\
-$^{\text{\sf 5}}$St Vincent's Institute of Medical Research, 41 Victoria Parade, Fitzroy Victoria 3065, Australia; and \\
-$^{\text{\sf 6}}$CRUK Cambridge Institute, University of Cambridge, Robinson Way, Cambridge CB2 0RE, United Kingdom.
-% $^{\text{\sf 4}}$Department of Statistics, University of Oxford, Oxford, Oxfordshire, UK;
-}
+\end{flushleft}
 
-\corresp{$^\ast$To whom correspondence should be addressed.}
-
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
-
-\editor{Associate Editor: XXXXXXX}
-
-\abstract{\textbf{Motivation:} Single-cell RNA sequencing (scRNA-seq) is increasingly used to study gene expression at the level of individual cells.
+\section*{Abstract}
+\textbf{Motivation:} Single-cell RNA sequencing (scRNA-seq) is increasingly used to study gene expression at the level of individual cells.
 However, preparing raw sequence data for further analysis is not a straightforward process.
-Biases, artifacts, and other sources of unwanted variation are present in the data, requiring substantial time and effort to be spent on pre-processing, quality control (QC) and normalisation.\\
+Biases, artifacts, and other sources of unwanted variation are present in the data, requiring substantial time and effort to be spent on pre-processing, quality control (QC) and normalisation.\\ \noindent
 \textbf{Results:} We have developed the R/Bioconductor package \emph{scater} to facilitate rigorous pre-processing, quality control, normalisation and visualisation of scRNA-seq data.
 The package provides a convenient, flexible workflow to process raw sequencing reads into a high-quality expression dataset ready for downstream analysis.
-\emph{scater} provides a rich suite of plotting tools for single-cell data and a flexible data structure that is compatible with existing tools and can be used as infrastructure for future software development.\\
+\emph{scater} provides a rich suite of plotting tools for single-cell data and a flexible data structure that is compatible with existing tools and can be used as infrastructure for future software development.\\ \noindent
 \textbf{Availability:} The open-source code, along with installation instructions, vignettes and case studies, is available through Bioconductor at
-\href{http://bioconductor.org/packages/scater}{http://bioconductor.org/packages/scater}.\\
-\textbf{Contact:} \href{davis@ebi.ac.uk}{davis@ebi.ac.uk}\\
-\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}
+\href{http://bioconductor.org/packages/scater}{http://bioconductor.org/packages/scater}.\\ \noindent
+\textbf{Supplementary information:} Supplementary material is available online at \emph{bioRxiv} accompanying this manuscript, and all materials required to reproduce the results presented in this paper are available at \href{http://dx.doi.org/10.5281/zenodo.60139}{dx.doi.org/10.5281/zenodo.60139}.
 
-\maketitle
+% now start line numbers
+\linenumbers
 
-\section{Introduction}\label{introduction}
+\section*{Introduction}\label{introduction}
 
@@
 
-\begin{methods}
-\section{Methods, Data and Implementation}\label{methods-data-and-implementation}
+\section*{Methods, Data and Implementation}\label{methods-data-and-implementation}
 
-\subsection{Case study with scRNA-seq data}\label{case-study-with-scrna-seq-data}
+\subsection*{Case study with scRNA-seq data}\label{case-study-with-scrna-seq-data}
 
@@
 
-\begin{figure}[!tpb]%figure1
-\centerline{\includegraphics[width=0.48\textwidth]{figures/scater_qc_workflow.pdf}}
-\caption{An overview of the \emph{scater} workflow, from raw sequenced reads to a tidy data set ready for higher-level downstream analysis.
+\begin{figure}[tpb]%figure1
+\centerline{\includegraphics[width=0.52\textwidth]{scater_qc_workflow}}
+\caption{\small An overview of the \emph{scater} workflow, from raw sequenced reads to a tidy data set ready for higher-level downstream analysis.
 For step 5, explanatory variables include experimental covariates like batch, cell source and other recorded information, as well as QC metrics computed from the data.
@@
 
-\subsection{Implementation}\label{implementation}
+\subsection*{Implementation}\label{implementation}
 
@@
 Key aspects of the code are written in C++ to minimise computational time and memory use.
-The package builds on many other R packages, including \emph{Biobase} and \emph{BiocGenerics} for core Bioconductor functionality \citep{Huber2015-en}; \emph{destiny} \citep{Angerer2015-sw} and \emph{Rtsne} \citep{Krijthe2015-is} for dimensionality reduction; and \emph{edgeR} \citep{Robinson2010-ky} and \emph{limma} \citep{Ritchie2015-so} for model fitting and statistical analyses.
+The package builds on many other R packages, including \emph{Biobase} and \emph{BiocGenerics} for core Bioconductor functionality \citep{Huber2015-en}; \emph{destiny} \citep{Angerer2015-sw} and \emph{Rtsne} for dimensionality reduction; and \emph{edgeR} \citep{Robinson2010-ky} and \emph{limma} \citep{Ritchie2015-so} for model fitting and statistical analyses.
 The plotting functionality in the package uses \emph{ggplot2} \citep{Wickham2016-dc}.
@@
 
-\end{methods}
+\section*{Results}\label{results}
 
-\section{Results}\label{results}
-
-\subsection{The \emph{scater} package}\label{the-scater-package}
+\subsection*{The \emph{scater} package}\label{the-scater-package}
 
@@
 
-\subsection{SCESet: a data structure for single-cell expression
+\subsection*{SCESet: a data structure for single-cell expression
 data}\label{sceset-a-data-structure-for-single-cell-expression-data}
@@
 
-\subsection{Data pre-processing}\label{data-pre-processing}
+\subsection*{Data pre-processing}\label{data-pre-processing}
 
@@
 
-\enlargethispage{6pt}
-
-\begin{figure*}[!htpb]%figure3
-\centerline{\includegraphics[width=\textwidth]{figures/figure3_alt.pdf}}
-\caption{\textbf{Different types of QC plots that can be generated with \emph{scater}.} (\textbf{a}) Cumulative expression plot showing the proportion of the library accounted for by the top 1--500 most highly-expressed features.
-(\textbf{b}) PCA plot produced using a subset of the QC metrics computed with \emph{scater's} calculateQCMetrics function.
-(\textbf{c}) Plot of frequency of expression (percentage of cells in which the feature is deemed expressed) against mean expression level across cells.
-The vertical dotted line shows the median of the gene mean expression levels, and the horizontal dotted line indicates 50\% frequency of expression.
-(\textbf{d}) Plot of the 20 most highly-expressed features (computed according to the highest total read counts) across all cells in the data set.
-For each feature, the circle represents the percentage of counts that the gene accounts for computed from counts pooled across all cells.
-The genes are ordered by this value.
-The bars for each gene show the percentage of counts accounted for by the gene for each individual cell, providing a visualisation of the distribution across cells.
-(\textbf{e}) Density plot showing the percentage of variance explained by a set of explanatory variables across all genes.
-Each individual plot is produced by a single call with either the function plot (\textbf{a}), plotPCA (\textbf{b}) or plotQC (\textbf{c--e}).}\label{fig:03}
-\end{figure*}
-
-\subsection{Data quality control}\label{data-quality-control}
+\subsection*{Data quality control}\label{data-quality-control}
 
@@
 low- and zero-expression observations in scRNA-seq data.
+
+\begin{figure*}[tpb]%figure3
+\begin{adjustwidth}{-2in}{0in}
+\begin{flushright}
+\includegraphics[width=1.3\textwidth]{figure3_alt}
+\end{flushright}
+\justify
+\color{Gray}
+\caption{\small \textbf{Different types of QC plots that can be generated with \emph{scater}.} (\textbf{a}) Cumulative expression plot showing the proportion of the library accounted for by the top 1--500 most highly-expressed features.
+(\textbf{b}) PCA plot produced using a subset of the QC metrics computed with \emph{scater's} calculateQCMetrics function.
+(\textbf{c}) Plot of frequency of expression (percentage of cells in which the feature is deemed expressed) against mean expression level across cells.
+The vertical dotted line shows the median of the gene mean expression levels, and the horizontal dotted line indicates 50\% frequency of expression.
+(\textbf{d}) Plot of the 20 most highly-expressed features (computed according to the highest total read counts) across all cells in the data set.
+For each feature, the circle represents the percentage of counts that the gene accounts for computed from counts pooled across all cells.
+The genes are ordered by this value.
+The bars for each gene show the percentage of counts accounted for by the gene for each individual cell, providing a visualisation of the distribution across cells.
+(\textbf{e}) Density plot showing the percentage of variance explained by a set of explanatory variables across all genes.
+Each individual plot is produced by a single call with either the function plot (\textbf{a}), plotPCA (\textbf{b}) or plotQC (\textbf{c--e}).}\label{fig:03}
+\end{adjustwidth}
+\end{figure*}
 
@@
 
-\begin{figure*}[!htpb]%figure4
-\centerline{\includegraphics[width=\textwidth]{figures/figure4.pdf}}
-\caption{\textbf{Reduced dimension representations of cells and gene expression plots with \emph{scater.}} Plots are shown using all genes (\textbf{a-c}) and cell cycle genes only (\textbf{d-f}) using PCA (\textbf{a,d}), t-SNE (\textbf{b,e}) and diffusion maps (\textbf{c,f}), where each point represents a cell.
-In the top row (\textbf{a-c}), points are coloured by patient of origin, sized by total features (number of genes with detectable expression) and the shape indicates the C1 machine used to process the cells.
-In the second row (\textbf{d-f}), points are coloured by the expression of \emph{CCND2} (a gene associated with the G1/S phase transition of the cell cycle) in each cell.
-With the plotExpression function, gene expression can be plotted against any cell metadata variables or the expression of another gene---here, expression for the CD86, IGH44 and IGHV4-34 genes in each cell is plotted against the patient of origin (\textbf{g}).
-The function automatically detects whether the x-axis variable is categorical or continuous and plots the data accordingly, with x-axis values ``jittered'' to avoid excessive overplotting of points with the same x coordinate.}\label{fig:04}
-\end{figure*}
-
 The \emph{scater} graphical user interface (GUI) provides convenient
@@
 
-\subsection{Data visualisation}\label{data-visualisation}
+\subsection*{Data visualisation}\label{data-visualisation}
 
@@
 genes that are associated with a biological process of interest, and construct
-plots using only these features. For example, \citealp{Scialdone2015-gj}
+plots using only these features. For example, \citet{Scialdone2015-gj}
 found that using prior knowledge to define feature sets is vital for
@@
 
-\subsection{Data normalisation and batch correction} \label{data-normalisation-and-batch-correction}
+\subsection*{Data normalisation and batch correction} \label{data-normalisation-and-batch-correction}
 
@@
 In addition, \emph{scater} is tightly integrated with the \emph{scran} package that implements a method utilising cell pooling and deconvolution to compute size factors better suited to scRNA-seq data \citep{Lun2016-sk}.
+
+\begin{figure*}[htpb]%figure4
+\begin{adjustwidth}{-2in}{0in}
+\begin{flushright}
+\includegraphics[width=1.3\textwidth]{figure4}
+\end{flushright}
+\justify
+\color{Gray}
+\caption{\small \textbf{Reduced dimension representations of cells and gene expression plots with \emph{scater.}} Plots are shown using all genes (\textbf{a-c}) and cell cycle genes only (\textbf{d-f}) using PCA (\textbf{a,d}), t-SNE (\textbf{b,e}) and diffusion maps (\textbf{c,f}), where each point represents a cell.
+In the top row (\textbf{a-c}), points are coloured by patient of origin, sized by total features (number of genes with detectable expression) and the shape indicates the C1 machine used to process the cells.
+In the second row (\textbf{d-f}), points are coloured by the expression of \emph{CCND2} (a gene associated with the G1/S phase transition of the cell cycle) in each cell.
+With the plotExpression function, gene expression can be plotted against any cell metadata variables or the expression of another gene---here, expression for the CD86, IGH44 and IGHV4-34 genes in each cell is plotted against the patient of origin (\textbf{g}).
+The function automatically detects whether the x-axis variable is categorical or continuous and plots the data accordingly, with x-axis values ``jittered'' to avoid excessive overplotting of points with the same x coordinate.}\label{fig:04}
+\end{adjustwidth}
+\end{figure*}
 
@@
 
-\begin{figure*}[!tpb]%figure5
-\centerline{\includegraphics[width=\textwidth]{figures/figure5.pdf}}
-\caption{\textbf{Figure 5: Normalisation and batch correction with
+\begin{figure*}[tpb]%figure5
+\begin{adjustwidth}{-2in}{0in}
+\begin{flushright}
+\includegraphics[width=1.3\textwidth]{figure5}
+\end{flushright}
+\justify
+\color{Gray}
+\caption{\small \textbf{Normalisation and batch correction with
 \emph{scater}.} Principal component analysis plots showing cell
 structure in the first two PCA dimensions using various normalisation
-methods methods that can be easily applied in \emph{scater}, including
+methods that can be easily applied in \emph{scater}, including
 endogenous size-factor normalisation using methods from the \emph{scran}
@@
 expression in the cell.}\label{fig:05}
+\end{adjustwidth}
 \end{figure*}
 
-\subsection{Software and data integration}\label{software-and-data-integration}
+\subsection*{Software and data integration}\label{software-and-data-integration}
 
@@
 
-\section{Discussion}\label{discussion}
+\section*{Discussion}\label{discussion}
 
@@
 
-\section{Conclusion}\label{conclusion}
+\section*{Conclusion}\label{conclusion}
 
@@
 
-\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{plain}
-%\bibliographystyle{bioinformatics}
+\nolinenumbers
+
+\bibliographystyle{abbrvnat}
+
+%This is where your bibliography is generated. Make sure that your .bib file is actually called library.bib
 \bibliography{main}

### 2016-11-06 scater_case_study/timing_macosko.Rmd
Revisions to manuscript for resubmission

---
title: "Timings for using scater with Macosko dataset"
author: Davis McCarthy, Kieran Campbell, Aaron Lun & Quin Wills
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        theme: united
        highlight: tango
        number_sections: false
        code_folding: show
---

```{r setup, include=FALSE}
library("scater")
```

# Introduction

This notebook shows timing results for using scater with the Macosko et al Drop-seq dataset consisting of approximately 45,000 cells. The dataset in its current format was supplied by Vladimir Kiselev.

# Read data

```{r read-data}
counts_mat <- readRDS("../../../../Dropbox/Projects/single-cell-data/macosko2015/macosko_norm.rds")
dim(counts_mat)
```

# Create `SCESet` object

We can make a minimal `SCESet` object from the count matrix.

```{r create-sceset}
cluster <- colnames(counts_mat)
colnames(counts_mat) <- paste0("Cell", 1:ncol(counts_mat))
system.time(
    sce <- newSCESet(countData = counts_mat)
)
```

This object now contains the count data, and the data transformed to log2-counts-per-million.

```{r show-sceset}
rm(counts_mat)
sce$cluster <- cluster
sce
```

# Calculate QC metrics

We do not appear to have ERCC spike-ins in this dataset, so use mitochondrial genes as feature controls.

```{r calc-qc-metrics}
mtgenes <- grepl("MT:", rownames(sce))
system.time(
    sce <- calculateQCMetrics(sce, feature_controls = list(MT = mtgenes))
)
```

We can look at plots of QC metrics:

```{r plot-pheno}
plotPhenoData(sce, aes(x = log10_total_counts, y = total_features, 
                       colour = pct_counts_feature_controls))
```

Such phenotype data plots take little time to produce, even for ~45,000 cells.

# Subsetting

Subsetting a random sample of 10,000 cells from the `SCESet` object only take a matter of seconds, so filtering out unwanted cells will not be a major hurdle.

````{r subsetting}
system.time(
    sce[, sample(ncol(sce), size = 10000)]
)
```

# Automatic outlier detection without output

Automatic outlier detection using PCA on QC metrics does not take too long.

```{r auto-outlier, results='hide'}
system.time(
    sce <- plotPCA(sce, pca_data_input = "pdata", detect_outliers = TRUE,
                   return_SCESet = TRUE)
)
```

# PCA using expression values

PCA using expression values still takes less than a couple of minutes.

```{r pca}
system.time(
    p1 <- plotPCA(sce, colour_by = "outlier")
)
p1
```

We can look at the original clusters defined by the authors in a PCA plot.

```{r pca-cluster}
system.time(
    p2 <- plotPCA(sce, colour_by = "cluster")
)
p2
```

# Automatic outlier detection with output

Automatic outlier detection using PCA on QC metrics does not take too long.

```{r auto-outlier-with-output}
system.time(
    sce <- plotPCA(sce, pca_data_input = "pdata", detect_outliers = TRUE,
                   return_SCESet = TRUE)
)
```

### 2016-11-06 supplement_figs.tex
Revisions to manuscript for resubmission

@@
 Calling the \verb|scater_gui| function from within an R session opens up the GUI in a web browser.
+
+\section*{Entry points for third-party tools from \emph{scater}}
+
+The \emph{scater} package serves to prepare data for a wide variety of downstream analyses with third-party tools.
+Given the diverse nature of analyses that can be done with scRNA-seq data, the entry points for various third-party tools in terms of data preparation and transformation/format can vary.
+Below we discuss entry points from \emph{scater} into example third-party tools representing major categories of downstream analysis.
+
+\paragraph{Differential expression analysis with \emph{scde} or \emph{edgeR}} Differential expression analysis tools for RNA-seq data, including \emph{scde} and \emph{edgeR} take raw counts as input data and can handle known batch effects in their statistical models.
+Therefore, the upstream data processing with \emph{scater} is straight-forward: QC on cells and genes should be carried out with \emph{scater} as usual, and the count data supplied to the DE tool.
+Accessing the count matrix is as simple as applying the counts() function to an SCESet object.
+For an \emph{edgeR} analysis, size-factors computed with \emph{scran} should also be supplied, and the convertTo function from \emph{scran} makes it very easy to convert an SCESet object from a \emph{scater} workflow to a DGEList object needed for an \emph{edgeR} analysis.
+
+\paragraph{Clustering with \emph{SC3}} Clustering results can be negatively impacted by the inclusion of poor quality cells, so QC of cells and genes as usual with \emph{scater} is necessary before supplying data to a clustering algorithm.
+In general, clustering algorithms to not account internally for batch effects and similar, so it will often be desirable to normalise expression data with size factors and regress out known batch effects or other technical effects.
+In the case of clustering with \emph{SC3}, a QC'd and normalised SCESet object can be supplied directly to the sc3 function for clustering.
+
+\paragraph{Pseudotime estimation with \emph{monocle}} As with clustering, current pseudotime estimation methods do not internally account for batch effects.
+Thus, before applying pseudotime tools, data should be QC'd with \emph{scater} to remove problematic cells and genes.
+Size-factor normalisation of data is advisable, and the normaliseExprs function in \emph{scater} can be used to regress out known batch or technical effects.
+Once a filtered and normalised SCESet object has been obtained, the convertTo function in \emph{scran} can be used to convert the SCESet object to a CellDataSet object used in \emph{monocle}.
+
+The three examples above demonstrate that the QC steps in \emph{scater} are necessary before any downstream analyses.
+The entry point from \emph{scater} varies for different third-party tools, but is nevertheless straight forward in most cases.
 

### 2016-11-06 supplement_figs_biorxiv.tex
Revisions to manuscript for resubmission

--- 2016-11-06 supplement_figs.tex
@@
-\documentclass{revtex4}
-\usepackage{hyperref}
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in,marginparwidth=2in]{geometry}
+
+% use Unicode characters - try changing the option if you run into troubles with special characters (e.g. umlauts)
+\usepackage[utf8]{inputenc}
+\usepackage[english]{babel}
+
+% clean citations
+%\usepackage{cite}
+\usepackage[round]{natbib}
+
+% hyperref makes references clicky. use \url{www.example.com} or \href{www.example.com}{description} to add a clicky url
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% improves typesetting in LaTeX
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% text layout - change as needed
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in
+\textheight 8.75in
+
+% use adjustwidth environment to exceed text width (see examples in text)
+\usepackage{changepage}
+
+% adjust caption style
+%\usepackage[labelfont=bf,labelsep=period,singlelinecheck=off]{caption}
+
+% remove brackets from references
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% headrule, footrule and page numbers
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+
+% use \textcolor{color}{text} for colored text (e.g. highlight to-do areas)
+\usepackage{color}
+
+% define custom colors (this one is for figure captions)
+\definecolor{Gray}{gray}{.25}
+
+% this is required to include graphics
 \usepackage{graphicx}
+
+% use if you want to put caption to the side of the figure - see example in text
+\usepackage{sidecap}
+
+% use for have text wrap around figures
+\usepackage{wrapfig}
+\usepackage[pscoord]{eso-pic}
+\usepackage[fulladjust]{marginnote}
+\reversemarginpar
+
+% document begins here
 \begin{document}
+\vspace*{0.35in}
 
-\title[scater supplementary figures]{Supplementary Figures---``\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R''}
-\author{Davis J.~McCarthy\,$^{\text{1,2,5}*}$, Kieran R.~Campbell\,$^{\text{2,4}}$, Aaron T.~L.~Lun\,$^{\text{6}}$ and Quin F.~Wills\,$^{\text{2,3}}$}
-\address{$^{\text{1}}$European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, United Kingdom;\\
-$^{\text{2}}$Wellcome Trust Centre for Human Genetics, University of Oxford,
-Roosevelt Drive, Oxford OX3 7BN, United Kingdom;\\
-$^{\text{3}}$Weatherall Institute for Molecular Medicine, University of Oxford, John Radcliffe Hospital, Oxford OX3 9DS, United Kingdom;\\
-$^{\text{4}}$Department of Physiology, Anatomy and Genetics, University of Oxford, South Parks Road, Oxford OX1 3QX, United Kingdom;\\
-$^{\text{5}}$St Vincent's Institute of Medical Research, 41 Victoria Parade, Fitzroy Victoria 3065, Australia; and \\
-$^{\text{6}}$CRUK Cambridge Institute, University of Cambridge, Robinson Way, Cambridge CB2 0RE, United Kingdom.
-% $^{\text{\sf 4}}$Department of Statistics, University of Oxford, Oxford, Oxfordshire, UK;
+% title goes here:
+\begin{flushleft}
+{\Large
+\textbf\newline{Supplementary Material and Figures---\emph{scater:} pre-processing, quality control, normalisation and visualisation of single-cell RNA-seq data in R}
 }
+\newline
+% authors go here:
+\\
+Davis J.~McCarthy\,\textsuperscript{1,2,5,*},
+Kieran R.~Campbell\,\textsuperscript{2,4},
+Aaron T.~L.~Lun\,\textsuperscript{6},
+Quin F.~Wills\,\textsuperscript{2,3}
+\\
+\bigskip
+\bf{1} European Molecular Biology Laboratory - European Bioinformatics Institute (EMBL-EBI), Hinxton CB10 1SD, United Kingdom;
+\\
+\bf{2} Wellcome Trust Centre for Human Genetics, University of Oxford,
+Roosevelt Drive, Oxford OX3 7BN, United Kingdom;
+\\
+\bf{3} Weatherall Institute for Molecular Medicine, University of Oxford, John Radcliffe Hospital, Oxford OX3 9DS, United Kingdom;
+\\
+\bf{4} Department of Physiology, Anatomy and Genetics, University of Oxford, South Parks Road, Oxford OX1 3QX, United Kingdom;
+\\
+\bf{5} St Vincent's Institute of Medical Research, 41 Victoria Parade, Fitzroy Victoria 3065, Australia; and
+\\
+\bf{6} CRUK Cambridge Institute, University of Cambridge, Robinson Way, Cambridge CB2 0RE, United Kingdom.
+\\
+\bigskip
+* \textbf{Contact:} \href{davis@ebi.ac.uk}{davis@ebi.ac.uk}
 
-\begin{abstract}
-Supplementary material and figures---details of package dependencies; an overview of the SCESet class; an overview of the \emph{scater} ecosystem; examples of using the \emph{scater} GUI.\\ \\
+\end{flushleft}
+
+\section*{Abstract}
+Supplementary material and figures---details of package dependencies; entry points for common third-party tools from \emph{scater}; an overview of the SCESet class; an overview of the \emph{scater} ecosystem; examples of using the \emph{scater} GUI.\\ \noindent
 \textbf{Contact:} \href{davis@ebi.ac.uk}{davis@ebi.ac.uk}
-\end{abstract}
 
-\maketitle
+% now start line numbers
+\linenumbers
 
@@
 Once a filtered and normalised SCESet object has been obtained, the convertTo function in \emph{scran} can be used to convert the SCESet object to a CellDataSet object used in \emph{monocle}.
+\newline
 
@@
 
-\begin{figure}[!tpb]%figure2
+\begin{figure}[!htpb]%figure2
 \centerline{\includegraphics[width=0.95\textwidth]{figures/sceset_outline.pdf}}
@@
 
-\bibliographystyle{natbib}
+\clearpage
+
+\nolinenumbers
+
+\bibliographystyle{abbrvnat}
 %\bibliographystyle{achemnat}

### 2016-11-07 main.tex
Final revisions for resubmission to Bioinformatics

@@
 
-One of the most popular frameworks for interactive analysis is the R programming language, extended for biological data analysis through the Bioconductor project \citep{Huber2015-en}.
+One of the most \revised{widely used} frameworks for interactive analysis is the R programming language, extended for biological data analysis through the Bioconductor project \citep{Huber2015-en}.
 While Bioconductor packages have been widely used for bulk RNA-seq data, the existing data structures (like the ExpressionSet class) are not sufficient for scRNA-seq data.
@@
 In particular, current visualisation methods designed for exploratory data analysis of bulk transcriptomic experiments are unsuited to datasets containing hundreds or thousands of cells.
-The large size of each dataset also favours methods such as \emph{kallisto} \citep{Bray2016-pj} and \emph{Salmon} \revised{\citep{Patro2016-rs}} for rapidly quantifying gene expression.
+\revised{The large size of each dataset also favours methods such as \emph{kallisto} \citep{Bray2016-pj} and \emph{Salmon} \revised{\citep{Patro2016-rs}} for rapidly quantifying gene expression, but support for the output of these methods is currently limited}.
 Extensions to the current computational infrastructure are required to provide appropriate data structures and methods that can accommodate these rich scRNA-seq datasets for integrative analyses of expression and other assay data along with the accompanying metadata.
@@
 RNA-seq reads were were mapped to a custom genome reference, consisting of Homo sapiens GRCh37 (primary assembly from \href{ftp://ftp.ensembl.org/pub/release-75/fasta/homo_sapiens/dna/}{ftp.ensembl.org/pub/release-75/fasta/homo\_sapiens/dna/}, last accessed 14.08.2015), Epstein-Barr Virus type 1 (B95-8 strain, Accession NC\_007605.1) and ERCC RNA spike-ins (ThermoFisher).
-Reads in fastq format were aligned using TopHat2 v2.0.12 \citep{Kim2013-qb} using Bowtie2 v2.2.3.0 \citep{Langmead2012-yc} as the core-mapping engine (\verb|--mate-inner-dist 190| \verb|--mate-std-dev 40| \verb|--report-secondary-alignments|) and other default parameters.
+Reads in fastq format were aligned \revised{with} TopHat2 v2.0.12 \citep{Kim2013-qb}, using Bowtie2 v2.2.3.0 \citep{Langmead2012-yc} as the core mapping engine (\verb|--mate-inner-dist 190| \verb|--mate-std-dev 40| \verb|--report-secondary-alignments|) and other default parameters.
 Potential PCR duplicates were marked with Picard MarkDuplicates v1.92(1464).
@@
 \centerline{\includegraphics[width=0.48\textwidth]{scater_qc_workflow.eps}}
-\caption{An overview of the \emph{scater} workflow, from raw sequenced reads to a tidy data set ready for higher-level downstream analysis.
+\caption{An overview of the \emph{scater} workflow, from raw sequenced reads to a \revised{high quality} data set ready for higher-level downstream analysis.
 For step 5, explanatory variables include experimental covariates like batch, cell source and other recorded information, as well as QC metrics computed from the data.
@@
 A common subsequent step for these methods is to collapse transcript-level expression to gene-level expression.
-Exploiting the \emph{biomaRt} R/Bioconductor package, \emph{scater} provides a convenient function for using Ensembl annotations to obtain
-gene-level expression values and gene or transcript annotations \citep{Yates2016-ia}.
+Exploiting the \emph{biomaRt} R/Bioconductor package, \emph{scater} provides a convenient function for using Ensembl annotations \revised{\citep{Yates2016-ia}} to obtain gene-level expression values and gene or transcript annotations.
 
@@
 (\textbf{d}) Plot of the 20 most highly-expressed features (computed according to the highest total read counts) across all cells in the data set.
-For each feature, the circle represents the percentage of counts that the gene accounts for computed from counts pooled across all cells.
-The genes are ordered by this value.
-The bars for each gene show the percentage of counts accounted for by the gene for each individual cell, providing a visualisation of the distribution across cells.
+\revised{For each feature, the circle represents the percentage of counts across all cells that correspond to that feature.
+The features are ordered by this value.
+The bars for each feature show the percentage of counts corresponding to the feature in each individual cell}, providing a visualisation of the distribution across cells.
 (\textbf{e}) Density plot showing the percentage of variance explained by a set of explanatory variables across all genes.
@@
 
-\revised{The plotPCA function implements an approach to automatic outlier
-detection using multivariate normal methods applied to the cell-level QC
-metrics \citep{Ilicic2016-dm} and a PCA plot computed from QC metrics is produced showing detected outliers (Figure~\ref{fig:03}b).
-Briefly, semi-robust principal components are computed on robustly sphered QC metrics data to determine distances for each cell and weights for outlier detection, using the pcout function from the \emph{mvoutlier} R package, which implements the fast algorithm of \citet{Filzmoser2008-sy}.
-By default, the following QC metrics are used for automatic outlier detection: the percentage of counts from the top 100 features, the total number of features with detectable expression, the percentage of counts from control features, the number of detected feature controls, the log-scaled counts from endogenous features and log-transformed counts from feature controls are used.
+\revised{The plotPCA function implements an approach to automatic outlier detection using multivariate normal methods applied to the cell-level QC
+metrics \citep{Ilicic2016-dm}.
+It produces a PCA plot computed from QC metrics, where cells corresponding to detected outliers are marked (Figure~\ref{fig:03}b).
+Briefly, semi-robust principal components are computed from robustly sphered QC metrics data, using the pcout function from the mvoutlier R package \citep{Filzmoser2008-sy}.
+These components are used to calculate distances for each cell, which are then used to compute weights for outlier detection.
+By default, the following QC metrics are used in this procedure: the percentage of counts from the top 100 features, the total number of features with detectable expression, the percentage of counts from control features, the number of detected feature controls, the log-scaled counts from endogenous features and log-transformed counts from feature controls.
 The user can specify QC metrics or other cell metadata variables to use for outlier detection with the "selected\_variables" argument to plotPCA.
@@
 such as a plot to visualise the frequency of expression of features against their average expression level (Figure~\ref{fig:03}c).
-Such plots are useful, because scRNA-seq is characterised by a high frequency of ``dropout'' events,
-that is, no observed expression (such as no read counts) in a particular
-cell for a gene that is actually expressed in that cell. Typically only
-a small set of genes are observed with detectable expression in every
-cell. With plotQC, control features can be highlighted easily in the plot, and
-typical scRNA-seq datasets will show a broadly sigmoidal relationship between
-average expression level and frequency of expression across cells. This is
-consistent with expected behaviour where genes with greater average expression
-are more readily captured during library preparation and are detected at a
-greater frequency \citep{Brennecke2013-zv,Kim2015-xd,Vallejos2015-ww}.
+\revised{Such plots are useful because scRNA-seq data are characterised by a high frequency of "dropout" events, i.e., no observed expression (such as no read counts) in a particular cell for a gene that is actually expressed in that cell.
+Indeed, most genes will not have detectable expression in every cell.
+With plotQC, control features that should be present in each cell can be highlighted easily in the plot, allowing technical dropouts to be distinguished from biological heterogeneity of expression.
+Typical scRNA-seq datasets will show a broadly sigmoidal relationship between average expression level and frequency of expression across cells.} This is consistent with expected behaviour where genes with greater average expression are more readily captured during library preparation and are detected at a greater frequency \citep{Brennecke2013-zv,Kim2015-xd,Vallejos2015-ww}.
 
-With plotQC we can also produce a plot to visualise the most
+With plotQC, we can also produce a plot to visualise the most
 highly-expressed features in the dataset (Figure~\ref{fig:03}d). This
@@
 the phenoData slot of the SCESet object, we fit a linear model for each
-feature with just that variable as the explanatory variable. We then plot the distribution of the marginal $R^2$ values across all features for the variables with the most
-explanatory power for the dataset (Figure~\ref{fig:03}e). The variables are ranked by median $R^2$ across features in the plot, allowing users to identify variables that may need to be considered
+feature with \revised{only} that variable as the explanatory variable.
+We then plot the distribution of the marginal $R^2$ values across all features for the variables with the most explanatory power for the dataset (Figure~\ref{fig:03}e).
+The variables are ranked by median $R^2$ across features in the plot, allowing users to identify variables that may need to be considered
 during normalisation or statistical modelling. The plotQC function can
@@
 In the second row (\textbf{d-f}), points are coloured by the expression of \emph{CCND2} (a gene associated with the G1/S phase transition of the cell cycle) in each cell.
-With the plotExpression function, gene expression can be plotted against any cell metadata variables or the expression of another gene---here, expression for the CD86, IGH44 and IGHV4-34 genes in each cell is plotted against the patient of origin (\textbf{g}).
+\revised{Furthermore, with} the plotExpression function, gene expression can be plotted against any cell metadata variables or the expression of another gene---here, expression for the CD86, IGH44 and IGHV4-34 genes in each cell is plotted against the patient of origin (\textbf{g}).
 The function automatically detects whether the x-axis variable is categorical or continuous and plots the data accordingly, with x-axis values ``jittered'' to avoid excessive overplotting of points with the same x coordinate.}\label{fig:04}
@@
 Frequently used methods for scaling normalization include the trimmed mean of M-values \citep{Robinson2010-dy}, relative log-expression \citep{Anders2010-kv} and upper-quartile methods \citep{Bullard2010-ui}, all of which are available for use in \emph{scater}.
-In addition, \emph{scater} is tightly integrated with the \emph{scran} package that implements a method utilising cell pooling and deconvolution to compute size factors better suited to scRNA-seq data \citep{Lun2016-sk}.
+In addition, \emph{scater} is tightly integrated with the \emph{scran} package, \revised{which} implements a method utilising cell pooling and deconvolution to compute size factors better suited to scRNA-seq data \citep{Lun2016-sk}.
 \revised{\citet{Lun2016-uj} also offers further discussion of the respective benefits and drawbacks of spike-in normalisation and non-DE normalisation.}
@@
 \citep{Leek2007-rg,Leek2014-nu} and \emph{RUVSeq} \citep{Risso2014-np} to
-identify and remove these unwanted sources of variation, and the
-removeBatchEffect method in the \emph{limma} package \citep{Ritchie2015-so}
-to account for known batch effects. Here, just removing the first latent
-variable identified by the RUVs method from \emph{RUVSeq} is sufficient to
-remove the machine effect, as the PCA plot now separates cells by patient
-rather than C1 machine (Figure~\ref{fig:05}c).
+identify and remove these unwanted sources of variation.
+Here, just removing the first latent variable identified by the RUVs method from \emph{RUVSeq} is sufficient to remove the machine effect, as the PCA plot now separates cells by patient rather than C1 machine (Figure~\ref{fig:05}c).
 \revised{More targeted applications of these methods can be used to remove specific effects, for example, by identifying latent factors from cell cycle genes to remove the cell cycle effect.}
@@
 Where this is not possible (e.g., for visualisation), directly regressing out these uninteresting factors is required to obtain "corrected" expression values for further analysis.
-\revised{A general risk of removing latent factors is that interesting biological variation may be removed along with the presumed unwanted variation.
+\revised{Furthermore, a general risk of removing latent factors is that interesting biological variation may be removed along with the presumed unwanted variation.
 Users should therefore apply such methods with appropriate caution, particularly when an analysis aims to discover biological conditions, such as new cell types.}
@@
 Once identified, important covariates and latent variables can be flagged for inclusion in downstream statistical models or their effects regressed out of normalised expression values.
-\revised{The package is well documented and a recent step-by-step workflow article demonstrates detailed use of \emph{scater} in combination with other analysis packages in a range of scenarios \citep{Lun2016-uj}.}
+\revised{The package is thoroughly documented and a recent step-by-step workflow article demonstrates detailed use of \emph{scater} in combination with other analysis packages in a range of scenarios \citep{Lun2016-uj}.}
 
@@
 producing a high-quality single-cell expression dataset for downstream analysis.
-The intuitive GUI implemented in \emph{scater} provides an easier entry point into rigorous analysis of scRNA-seq data for users without a
+The intuitive GUI implemented in \emph{scater} provides an \revised{easy} entry point into rigorous analysis of scRNA-seq data for users without a
 computational background, enabling them to process raw reads into high-quality expression data within a single computing environment.
-Experienced users can take advantage of \emph{scater}'s data structures, wide array of tools, suitability for scripted analyses and seamless integration with many other R/Bioconductor analysis tools.
+Experienced users can take advantage of \emph{scater}'s data structures, wide array of \revised{methods}, suitability for scripted analyses and seamless integration with many other R/Bioconductor analysis tools.
 The data structures and methods in
@@
 
-This work was supported by the National Health and Medical Research Council of Australia [APP1112681 to D.J.M.], by core funding from the European Molecular Biology Laboratory [D.J.M.], core funding from Cancer Research UK [A17197 to A.T.L.L.), the United Kingdom Medical Research Council [studentship to K.R.C.) and the Oxford Single Cell Biology Consortium [Q.F.W.].\vspace*{-12pt}
+This work was supported by the National Health and Medical Research Council of Australia [APP1112681 to D.J.M.], by core funding from the European Molecular Biology Laboratory [D.J.M.], core funding from Cancer Research UK [A17197 to A.T.L.L.], the United Kingdom Medical Research Council [studentship to K.R.C.] and the Oxford Single Cell Biology Consortium [Q.F.W.].\vspace*{-12pt}
 

### 2016-11-07 supplement_figs.tex
Final revisions for resubmission to Bioinformatics

@@
 \paragraph{Differential expression analysis with \emph{scde} or \emph{edgeR}} Differential expression analysis tools for RNA-seq data, including \emph{scde} and \emph{edgeR} take raw counts as input data and can handle known batch effects in their statistical models.
-Therefore, the upstream data processing with \emph{scater} is straight-forward: QC on cells and genes should be carried out with \emph{scater} as usual, and the count data supplied to the DE tool.
+Therefore, the upstream data processing with \emph{scater} is straightforward: QC on cells and genes should be carried out with \emph{scater} as usual, and the count data supplied to the DE tool.
 Accessing the count matrix is as simple as applying the counts() function to an SCESet object.
@@
 \paragraph{Clustering with \emph{SC3}} Clustering results can be negatively impacted by the inclusion of poor quality cells, so QC of cells and genes as usual with \emph{scater} is necessary before supplying data to a clustering algorithm.
-In general, clustering algorithms to not account internally for batch effects and similar, so it will often be desirable to normalise expression data with size factors and regress out known batch effects or other technical effects.
+In general, clustering algorithms do not account internally for batch effects and similar, so it will often be desirable to normalise expression data with size factors and regress out known batch effects or other technical effects.
 In the case of clustering with \emph{SC3}, a QC'd and normalised SCESet object can be supplied directly to the sc3 function for clustering.
@@
 The three examples above demonstrate that the QC steps in \emph{scater} are necessary before any downstream analyses.
-The entry point from \emph{scater} varies for different third-party tools, but is nevertheless straight forward in most cases.
+The entry point from \emph{scater} varies for different third-party tools, but is nevertheless straightforward in most cases.
 
